\documentclass[11pt,a4paper,openany]{book}

% --- Preamble ---
\input{preamble}

\title{Large Language Models as Engineered Systems \\
       \large A Systems, Statistical, and Human-Factors Primer}
\author{Lodewijk Bonebakker \\[0.4em]
        \small\textit{with drafting assistance from Claude Opus 4.7 (Anthropic)}}
\date{\today}

\begin{document}

\frontmatter

\maketitle
\tableofcontents

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% PREFACE
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\chapter*{Preface: A Letter to the Reader}
\addcontentsline{toc}{chapter}{Preface: A Letter to the Reader}

\section*{The thing in front of you}
\addcontentsline{toc}{section}{The thing in front of you}

You have, almost certainly, already used a Large Language Model. You
asked it to draft an email, or to debug a stubborn piece of code, or
to explain a concept that a textbook had failed to make stick. The
answer it gave you was probably useful. It was also, at some level,
a surprise --- not because the technology is new (it is not, really;
its lineage is more than a century old), but because the experience
of being addressed in fluent prose by a system with no body, no
mind, and no opinions felt different from anything you had asked a
computer to do before.

It is worth pausing on that surprise. The interesting question, for
the kind of engineer who wants to build systems around these models
or repair them when they misbehave, is not whether the answer was
correct. The interesting question is what the system was actually
\emph{doing}. We are about to take it apart, slowly, and put it back
together. By the time you finish this reader you will not, in any
deep sense, find these systems mysterious. You may, however, find
them more remarkable than you do today, and you will certainly find
them less trustworthy in some specific and useful ways.

\section*{How we got here}
\addcontentsline{toc}{section}{How we got here}

It is worth remembering that the field has a history. The Russian
mathematician Andrey Markov, in 1913, sat down with a copy of
Pushkin's \emph{Eugene Onegin} and counted the alternation of vowels
and consonants. He did this, in part, to make a philosophical point:
that conditional probability does not require independence. He gave
us a model of language as a sequence of dependencies a full century
before we had the data or the hardware to make that model useful.

Claude Shannon followed in 1948 with his theory of information,
which gave us a way to measure how surprising a sequence is, and
therefore how well a model has captured a language's regularities.
For decades after that the field crawled. The n-gram era of speech
recognition counted word triples on hard drives that filled rooms.
The connectionist revival of the 1980s put neural networks back on
the table. Bengio, Ducharme, Vincent, and Jauvin built a neural
language model in 2003 that almost nobody outside the field
noticed. Mikolov and his colleagues showed in 2013 that words could
be embedded in a vector space where ``king minus man plus woman''
landed near ``queen''. Sutskever, Vinyals, and Le followed with
sequence-to-sequence translation in 2014. Bahdanau, Cho, and Bengio
introduced attention in 2015. Vaswani and his coauthors at Google
removed the recurrence in 2017 and gave us the Transformer.

And then the field stopped crawling. GPT-1 in 2018, GPT-2 in 2019,
GPT-3 in 2020, ChatGPT at the end of 2022, GPT-4 the following
spring. Each of these was a substantial engineering effort by
hundreds of people whose names will not appear in this book. But
the lineage matters. None of these systems came out of nowhere.
None of them required anything that we could not, in retrospect,
have predicted from the trajectory. The mathematics of the
Transformer is the mathematics of attention, which is the
mathematics of weighted sums, which is, in the end, the mathematics
of probability. The whole edifice rests on a few sturdy ideas, and
this reader is in the business of giving you those ideas one at a
time.

\section*{How to read this book}
\addcontentsline{toc}{section}{How to read this book}

This is a reader for engineers. It assumes you have the
mathematical background of an undergraduate in a quantitative
discipline --- some calculus, some linear algebra, some probability,
the rough shape of an optimization argument. It does not assume you
have any prior exposure to deep learning or to LLMs in particular.

The book is organised into thirteen chapters, one per topic. Every
chapter follows the same structure. It opens with a short narrative
section called \emph{Setting the Scene}, which describes the
problem space the chapter is about, the lineage of ideas that
brought us to the current best practice, what that current best
practice is, and the characteristic ways the practice fails. The
narrative is followed by a notation block that fixes the symbols
and standing assumptions. Then come the derivations, the worked
numeric examples, the figures, and a final \emph{Proved vs Assumed}
recap that separates what the chapter has rigorously established
from what it has empirically observed. The chapter closes with
bridges to other chapters and a compact \emph{Derivation Ledger}
that indexes the central identities.

The narrative comes first deliberately. The math in the body of
each chapter is the bookkeeping for what the narrative describes.
A reader who knows what a thing is for tends to derive it more
easily; a reader who derives it first sometimes loses track of why
it matters. We write the why first. Then the how.

If you are reading the book front-to-back you will find that the
chapters build on one another. Cross-entropy from Chapter~1
reappears in Chapter~5 as the pretraining objective. The
mixed-precision memory budget from Chapter~2 reappears in
Chapter~7 as the LoRA budget. The trust boundary defined in
Chapter~9 reappears in Chapter~11 as the agent-loop invariant. The
dependency graph below makes the cross-references explicit so that
a reader who already knows part of the material can skip in safely.

\section*{What goes wrong, and why this is the most important
section of the book}
\addcontentsline{toc}{section}{What goes wrong}

A language model trained on cross-entropy is, by construction, the
most probable thing the model can say given the prompt. The most
probable thing is not the same as the true thing. The model has
no native concept of truth at all. It has a beautifully calibrated
sense of plausibility and a great deal of memorised plausibility,
which is sometimes the same as truth and sometimes is not. This is
the seed of \emph{hallucination} --- the polite name we give to a
fluent, confident, plausible, wrong answer.

Cross-entropy training cares about the average. It does not, by
itself, care about edge cases. The careful question that produces
a careful answer often differs from the careless question that
produces a wrong one only by a few tokens. We will see this as
\emph{prompt brittleness}, the fact that two prompts a human would
read as equivalent can produce two qualitatively different outputs.

A model that is fluent is not, by default, a model that is
helpful. A pretrained base model will happily complete an essay
that begins ``the best way to commit fraud is''; making it refuse
to do so is the work of \emph{alignment}, and alignment is its own
discipline of trade-offs. We will see how the discipline arrived,
what it bought, and what it cost --- the so-called \emph{alignment
tax} on capability, the \emph{sycophancy} that emerges when the
model is rewarded too much for agreeing with the annotator, the
\emph{reward hacking} that lets the model find a high-reward path
that no annotator would ever endorse.

A model that retrieves from a document store is only as honest as
the store. We will see how retrieval can hallucinate, how the
model can be \emph{lost in the middle} of a long context, and how
a malicious tool output can be crafted to look like a system
instruction --- the class of attack we call \emph{prompt
injection}, which is to LLM systems roughly what SQL injection
once was to web applications.

These failure modes are not embarrassments to be hidden. They are
the working vocabulary of the engineer who deploys these systems
into the world. Every chapter ends with a discussion of the
characteristic failures of the technique it presents. Read those
sections carefully. They are where the unhelpful intuition that
``the model just knows'' meets the more useful intuition of
``the model is doing exactly what we asked, and what we asked is
not quite what we meant.''

\section*{A note on tone}
\addcontentsline{toc}{section}{A note on tone}

The narrative sections in this book are written in plain language
because the technical sections are not. The technical sections are
written precisely because precision is what they are for. The
narrative sections aim at intuition because intuition is what
allows you to recover when the precision fails --- which it will,
because every model is wrong about something and every system has
a failure mode the documentation did not anticipate. We have tried
to acknowledge the people who built the ideas on which this field
stands, because giving credit is a courtesy, and because knowing
who first solved a problem helps you guess who has thought hardest
about its limits.

If a passage feels like it is trying to dazzle you, please tell us;
we will rewrite it. The intent is not awe. The intent is that you
put the book down able to reason about a class of system that did
not exist when you began reading.

\section*{On the writing of this book}
\addcontentsline{toc}{section}{On the writing of this book}

This reader was written in collaboration with Claude, a language
model produced by Anthropic. The subject matter of the book makes
that fact worth disclosing rather than concealing, and you should
know it before you read further.

The author chose the scope of the book, the chapters it would
contain, the historical threads it would trace, the voice the
narrative sections would adopt, and the derivations the technical
sections would carry. Each chapter was drafted, then read, then
revised, then committed. The model produced the prose and worked
through the algebra inside those choices. The author made the
choices, read the drafts, redirected the ones that drifted, and
approved each chapter before it landed.

Errors that survived this process are the author's, since
responsibility is a thing an author has and a language model does
not. Readers who find passages that are wrong, evasive, or fluent
in a way that papers over a real difficulty are encouraged to say
so. Any future edition will be better for the saying.

\section*{Reading Order and Chapter Dependencies}
\addcontentsline{toc}{section}{Reading Order and Chapter Dependencies}

The chapters are arranged so that each one's notation and results
build on its predecessors. Figure~\ref{fig:chapter_dependencies}
shows the directed dependency graph; arrows point from a source
chapter to any chapter that reuses its definitions, identities, or
assumptions. Readers familiar with parts of the stack can enter
later in the graph, but should still read any chapter whose
outputs feed the target chapter along the arrows shown.

\input{figures/chapter_dependencies}

\mainmatter

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% CHAPTERS
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\input{weeks/ch01_statistical_foundations}

\input{weeks/ch02_optimization_generalization}

\input{weeks/ch03_neural_networks}

\input{weeks/ch04_tokenization_embeddings}

\input{weeks/ch05_transformer_architecture}

\input{weeks/ch06_pretraining_incontext}

\input{weeks/ch07_efficiency_scaling_moe}

\input{weeks/ch08_adaptation_lora_quantization}

\input{weeks/ch09_alignment_preference_optimization}

\input{weeks/ch10_llm_systems_rag_tools_evaluation}

\input{weeks/ch11_governance_assurance_future}

\input{weeks/ch12_access_apis_agents_mcp}

\input{weeks/ch13_human_centric_error_model}

\backmatter

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% REFERENCES
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\bibliographystyle{apalike}
\bibliography{references}

\end{document}

