%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% CHAPTER 3 -- NEURAL NETWORKS
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\chapter{Neural Networks}
\label{sec:neural-networks}

\section*{Setting the Scene}
\addcontentsline{toc}{section}{Setting the Scene}

\paragraph{The problem.}
You have, by this point, two of the three things you need to train
a Large Language Model. From Chapter~1 you have the loss --- the
cross-entropy that asks, of any probability distribution over next
tokens, how surprised it was by the token that actually came next.
From Chapter~2 you have the optimiser --- stochastic gradient
descent, with its learning-rate schedules and its bias-corrected
moments and its Chinchilla-shaped compute budgets. What you do not
yet have is the \emph{function class} that the optimiser is
updating. That is the subject of this chapter. Before we can
compose the Transformer in Chapter~4, we need to introduce, one at
a time, the pieces it composes: a layer, a stack of layers, a way
to compute gradients through that stack, a choice of activation, a
method for initialising the weights, a normalisation, and a trick
for keeping deep stacks trainable. This chapter is the place where
each of those pieces enters the book.

\paragraph{How we got here.}
The idea that a network of simple units could learn a function from
examples is older than the computer that could run it. Frank
Rosenblatt, at Cornell in 1958, built the perceptron --- a single
linear threshold unit that could be trained to classify inputs by
adjusting its weights in proportion to its errors. The perceptron
was, briefly, a sensation. It was also limited: in 1969 Marvin
Minsky and Seymour Papert proved that no single linear threshold
unit could represent the exclusive-or of two inputs, and in the
backlash that followed the field receded for a decade.

The decisive response was depth. Paul Werbos, in his 1974 Harvard
PhD thesis, wrote down a recognisable derivation of what we now
call backpropagation --- the algorithm by which the chain rule of
calculus computes gradients through a multi-layer composition.
Backpropagation sat largely unread until 1986, when David
Rumelhart, Geoffrey Hinton, and Ronald Williams published a paper
in \emph{Nature} that brought it to the attention of an audience
ready to use it. Yann LeCun, at AT\&T, showed in 1989 that the
same algorithm could train a network to read handwritten zip
codes; the field had its first large-scale demonstration that
depth on real problems was tractable. Sepp Hochreiter and J\"urgen
Schmidhuber, in 1997, diagnosed the vanishing-gradient problem in
deep recurrent networks and proposed the gated cell that solved
it.

The next move was to make plain feed-forward depth work without
elaborate pre-training. Geoffrey Hinton, Simon Osindero, and
Yee-Whye Teh proposed a layer-wise pre-training recipe in 2006
that briefly closed the gap. Xavier Glorot and Yoshua Bengio in
2010 introduced the variance-preserving initialisation that made
the recipe unnecessary; the same year, Vinod Nair and Geoffrey
Hinton showed that the rectified linear unit --- a function so
simple it had been overlooked --- removed one of the main
bottlenecks in the gradient signal. The 2012 AlexNet paper, by
Alex Krizhevsky, Ilya Sutskever, and Geoffrey Hinton, took these
pieces to ImageNet at scale and won by a margin that ended the
debate about whether deep networks worked.

The remaining pieces converged in the years that followed. Sergey
Ioffe and Christian Szegedy added batch normalisation in 2015.
Kaiming He and his coauthors gave us, in two papers that same year,
both the ReLU-tuned variant of the Glorot init and the residual
connection that made networks of more than a hundred layers
trainable. Jimmy Ba, Jamie Ryan Kiros, and Geoffrey Hinton in 2016
introduced layer normalisation, the variant that survived into
the Transformer. Dan Hendrycks and Kevin Gimpel, also in 2016,
replaced the rectifier with a smoothed version called GeLU. Biao
Zhang and Rico Sennrich, in 2019, simplified layer normalisation
into RMSNorm. And Noam Shazeer, in 2020, identified the gated
activation variants (SwiGLU among them) that now sit inside almost
every Transformer's feed-forward block. Six decades of named
contributions, and the modern neural network is what you get when
you compose them.

\paragraph{Where we stand.}
A modern neural network used inside a Large Language Model is a
stack of differentiable transformations. Each transformation is a
linear layer (multiply by a weight matrix, add a bias vector)
followed by an elementwise nonlinearity. The whole network is
trained by gradient descent, with the gradients computed by
backpropagation. The depths that turned out to be useful --- tens
to hundreds of layers --- are made trainable by three engineering
moves: weight initialisation that preserves the variance of
activations across layers, a normalisation applied at each layer
to re-establish that variance after training has perturbed it, and
a residual connection that lets the optimiser learn small
corrections to an identity mapping rather than the full
transformation outright. Each of these moves was a response to a
specific failure of the design that came before; none was
inevitable. The Transformer of Chapter~4 is what you get when you
take this neural-network template and add the attention mechanism
on top.

\paragraph{What goes wrong.}
The failure modes of neural networks are the engineering
vocabulary of the field, and they recur often enough that you
should learn to recognise them. The first is \emph{vanishing
gradients}: in a deep network with the wrong initialisation or
the wrong activation, the gradient signal shrinks exponentially
with depth, and the early layers cease to learn. The second is
\emph{exploding gradients}: the same mechanism in reverse, with
activations and gradients growing past the range of floating-point
arithmetic until the network produces NaNs and the run dies. The
third is the \emph{dead-neuron problem}: a unit whose
pre-activation is permanently negative under a ReLU receives zero
gradient forever, becoming an ineffective contributor. The fourth
is the \emph{expressive-versus-trainable} gap: a network
architecture that \emph{can} represent the function we want is
not necessarily a network that the optimiser \emph{will find}
such a representation in. Each of these failures has a concrete
remedy --- careful initialisation, the right activation,
normalisation, residual paths --- and each remedy is one of the
moves this chapter unpacks.

Before that unpacking begins, a note on prerequisites. If you
have taken a course in deep learning and can write down
backpropagation on a whiteboard, you can skim or skip this
chapter; the material that follows is the foundational floor the
rest of the book stands on, and it will not surprise you. If you
have not, this is the chapter where the rest of the book becomes
legible. Either way, the goal is the same: to arrive at Chapter~4
with the function class in hand, ready to meet the architecture
that has, for now, eaten the field.

% --- END Setting the Scene ---

\section{Learning Objectives}

By the end of this chapter, you should be able to:
\begin{itemize}
  \item Define a multilayer perceptron as a composition of affine
        layers and elementwise nonlinearities, and state the
        universal-approximation result and its limits.
  \item Derive backpropagation by applying the calculus chain rule
        to a multi-layer composition; state the forward / backward
        FLOP equivalence that Chapters~2 and~4 use without proof.
  \item Pick an activation function (sigmoid, tanh, ReLU, GeLU,
        SwiGLU) given a task, a depth, and the resulting
        gradient-flow constraints.
  \item Initialise the weights of a feed-forward network using
        Xavier or He, justify the variance-preservation argument
        underneath both, and connect that argument to the
        $1/\sqrt{d_k}$ scaling Chapter~4 derives for attention.
  \item Distinguish BatchNorm, LayerNorm, and RMSNorm, and explain
        why Transformers use the latter two and not the first.
  \item Explain residual connections as identity-path stability,
        and state why depth becomes trainable when they are
        present and stalls when they are not.
  \item Identify the MLP block as the foundational primitive that
        Chapter~4 composes, alongside attention, into the
        Transformer's feed-forward sublayer.
\end{itemize}

\section{LLM Components Covered}

\textbf{Foundational primitives}
\begin{itemize}
  \item Affine (linear) layers and the multilayer perceptron (MLP).
  \item Backpropagation as the calculus-chain-rule implementation
        of the gradient through a layered composition.
  \item Activation functions: sigmoid, tanh, ReLU, GeLU, GLU and
        SwiGLU.
  \item Weight initialisation: Xavier/Glorot for tanh-style
        activations, He/Kaiming for ReLU-family activations.
  \item Normalisation layers: BatchNorm (overview only), LayerNorm,
        and RMSNorm.
  \item Residual / skip connections and the identity-path argument.
\end{itemize}

\textbf{Forward references}
\begin{itemize}
  \item Chapter~4 composes the primitives above into the
        Transformer's feed-forward block, derives the pre-norm
        placement, and gives the full residual-stream
        interpretation.
  \item Chapter~7 inserts low-rank deltas (LoRA) into the linear
        layers introduced here and uses He initialisation in the
        process.
  \item Chapter~8 reuses the logistic sigmoid as the link function
        in the Bradley--Terry preference model; the activation
        and the link function happen to be the same scalar
        transform.
\end{itemize}

\section{Notation and Standing Assumptions}
\label{sec:nn-notation}

\begin{notationbox}
A neural network in this chapter is a function
$f_\theta : \R^{d_{\mathrm{in}}} \to \R^{d_{\mathrm{out}}}$
with parameter vector $\theta$. The network has $L$ layers, indexed
$\ell \in \{1, \dots, L\}$. We write $h^{(\ell)} \in \R^{d_\ell}$
for the activations at layer $\ell$, with $h^{(0)} = x$ the input
and $h^{(L)} = f_\theta(x)$ the output. Each layer's parameters are
the weight matrix $W^{(\ell)} \in \R^{d_\ell \times d_{\ell-1}}$
and the bias vector $b^{(\ell)} \in \R^{d_\ell}$. The
\emph{pre-activation} at layer $\ell$ is
$z^{(\ell)} = W^{(\ell)} h^{(\ell-1)} + b^{(\ell)}$; the
\emph{post-activation} is
$h^{(\ell)} = \varphi(z^{(\ell)})$ where $\varphi : \R \to \R$ is
the elementwise nonlinearity (applied componentwise). The scalar
loss being minimised is $\mathcal{L}(\theta)$; from Chapter~1, this
is typically the cross-entropy under the empirical distribution.

\textbf{On the symbol $\varphi$ versus $\sigma$.}
We deliberately use $\varphi$ for the generic activation function,
not $\sigma$. The reason is that the symbol $\sigma$ is overloaded
in the rest of the book. Chapter~4 uses $\sigma$ both for the
activation inside the FFN equation and, in the LayerNorm
expression, for the per-token standard deviation. Chapter~8 uses
$\sigma(\cdot)$ for the logistic sigmoid as the link function in
the Bradley--Terry log-likelihood. To avoid confusion, this
chapter reserves $\sigma$ for the specific logistic sigmoid
$\sigma(z) = (1 + e^{-z})^{-1}$ and uses $\varphi$ when we mean
``some elementwise activation function, ReLU or GeLU or whatever
the design choice happens to be.''

\textbf{On the phrase ``chain rule''.}
Chapter~1 used the phrase ``chain rule'' to refer to the
\emph{probability} chain rule that factorises a joint distribution
over a sequence into a product of conditionals,
$P(x_1, \dots, x_T) = \prod_t P(x_t \mid x_{<t})$. This chapter
uses ``chain rule'' to refer to the \emph{calculus} chain rule
that propagates gradients through a composition of differentiable
functions, $(f \circ g)'(x) = f'(g(x)) \cdot g'(x)$. The two are
different theorems with the same name. Where ambiguity is
possible, we will say ``probability chain rule'' or ``calculus
chain rule'' explicitly.

\textbf{Standing assumptions.}
\begin{itemize}
  \item[A3.1] Each layer is differentiable almost everywhere.
        ReLU's non-differentiability at $z = 0$ is handled by
        convention: we take $\varphi'(0) := 0$, so that the
        backward pass is well-defined at every input.
  \item[A3.2] Inputs and parameters are real-valued and finite.
        Implementations evaluate the network in IEEE-754
        floating-point (typically fp32 master weights with bf16
        compute, per Chapter~2's mixed-precision recipe); we
        ignore the resulting rounding error in derivations and
        flag it explicitly when it matters operationally.
  \item[A3.3] The loss $\mathcal{L}$ is the empirical
        cross-entropy from Chapter~1, averaged over a minibatch
        of (context, token) pairs. Worked examples in this
        chapter occasionally use the squared-error loss for
        regression-style illustrations; this is a pedagogical
        choice and does not contradict A3.3, which describes the
        loss for the LLM training the rest of the book is about.
\end{itemize}
\end{notationbox}

% --- END Scaffolding ---

\section{The Neural Network as a Function Class}
\label{sec:nn-function-class}

\subsection{The Perceptron}

The earliest trainable neural unit is the \emph{perceptron},
introduced by \citet{rosenblatt1958perceptron}. A perceptron with
input dimension $d$ takes a real vector $x \in \R^d$, applies a
weighted sum and a threshold, and emits a binary output:
\begin{equation}
  y \;=\; \mathbb{1}\!\left[ w^\top x + b > 0 \right],
  \qquad w \in \R^d,\; b \in \R.
  \label{eq:nn-perceptron}
\end{equation}
The decision boundary is the hyperplane $w^\top x + b = 0$, which
in two dimensions is a line, in three dimensions a plane, and in
general a $(d-1)$-dimensional affine subspace separating $\R^d$
into two halves. Rosenblatt's contribution was a learning rule:
given a labelled example $(x^{(i)}, y^{(i)})$ with $y^{(i)} \in
\{0, 1\}$, update $w \leftarrow w + \eta (y^{(i)} - \hat{y}^{(i)})
x^{(i)}$ where $\hat{y}^{(i)}$ is the perceptron's prediction. If
the data are linearly separable, this rule is guaranteed to find a
separating hyperplane in finitely many updates. The perceptron
established that adjusting weights in proportion to errors was a
viable mechanism for learning.

\subsection{The Minsky--Papert Observation}

The perceptron's strength is also its limit. \citet{minsky1969perceptrons}
proved that a single perceptron cannot represent the
exclusive-or function: the four points
$\{(0,0), (0,1), (1,0), (1,1)\}$ with XOR labels $\{0, 1, 1, 0\}$
cannot be separated by any single line in the plane. The argument
is geometric: any line $w_1 x_1 + w_2 x_2 + b = 0$ partitions
$\R^2$ into two half-planes; one of those half-planes must
contain at least three of the four points; but the XOR labels
require that the two ``1''-labelled points and the two
``0''-labelled points sit on opposite sides of the boundary, in a
configuration where any straight line separating them would also
require the same straight line to be on its other side. There is
no such line. The result is sharp: it identifies a function so
simple that we can describe it in one word, and yet the
single-layer perceptron cannot compute it. Worked Example~B
later in this chapter walks the geometry explicitly.

The 1969 result was correct, narrow, and consequential. It was
correct because XOR really cannot be linearly separated. It was
narrow because the same argument does not apply to networks with
more than one trainable layer. It was consequential because the
field, in the climate of the time, treated the result as a verdict
on neural networks generally rather than as the specific
limitation of a specific architecture, and funding contracted
accordingly. The decade that followed is sometimes called the AI
winter; the multilayer response that ended it is the subject of
the next subsection.

\subsection{The Multilayer Response}

Adding a layer of nonlinear units \emph{between} the input and the
output produces a function class that can represent XOR (and much
else besides). Concretely: a multilayer perceptron (MLP) with $L$
layers, hidden widths $d_1, \dots, d_{L-1}$, and an elementwise
activation $\varphi$ is the composition
\begin{align}
  h^{(0)} &\;=\; x, \nonumber \\
  z^{(\ell)} &\;=\; W^{(\ell)} h^{(\ell-1)} + b^{(\ell)},
    \quad \ell = 1, \dots, L, \nonumber \\
  h^{(\ell)} &\;=\; \varphi\!\bigl(z^{(\ell)}\bigr),
    \quad \ell = 1, \dots, L-1, \nonumber \\
  f_\theta(x) &\;=\; h^{(L)} \;=\; z^{(L)},
  \label{eq:nn-mlp}
\end{align}
with parameters $\theta = (W^{(\ell)}, b^{(\ell)})_{\ell=1}^{L}$.
The final layer typically omits the activation when the output is
a real-valued vector (regression) or applies a softmax when the
output is a categorical distribution (classification, including
language modelling per Chapter~1). The intermediate layers
\emph{must} apply a nonlinearity: a composition of affine
transformations alone is itself affine, so depth without
nonlinearity buys nothing. We will see in §3 that the choice of
$\varphi$ matters; for now it suffices to know that any
non-polynomial $\varphi$ that is nonlinear gives the function
class real expressive reach.

\subsection{Universal Approximation, in One Paragraph}

The classical existence theorems make the expressive reach
precise. \citet{cybenko1989approximation} proved that an MLP with
one hidden layer and a sigmoidal activation can approximate any
continuous function on a compact subset of $\R^d$ to arbitrary
accuracy, given sufficient hidden width.
\citet{hornik1991approximation} extended the result: any
non-polynomial elementwise activation suffices. These are
existence theorems about \emph{representability}; they tell us
what an MLP \emph{can} compute. They tell us nothing about the
hidden width required (which can be exponentially large in the
input dimension), nothing about the number of training examples
needed, and nothing about whether stochastic gradient descent
will actually \emph{find} a good approximation in a reasonable
amount of time. For the engineering question of whether SGD
trains a usable model, universal approximation is necessary but
profoundly insufficient. The remainder of this chapter is about
the operational machinery that turns the existence theorem into a
trainable network.

\subsection{What This Chapter Will and Will Not Derive}

The function class of \eqref{eq:nn-mlp} is the object of study.
The chapter's job is to give the reader the operational machinery
that makes this class trainable in practice: an algorithm to
compute gradients (\S\ref{sec:nn-backprop}, backpropagation), a
choice of activation that does not stall those gradients
(\S\ref{sec:nn-activations}), an initialisation that does not
make the network start in a degenerate regime
(\S\ref{sec:nn-init}), a normalisation that keeps the activation
distribution from drifting during training
(\S\ref{sec:nn-norm}), and a residual connection that lets depth
be added without compounding the optimisation problem
(\S\ref{sec:nn-residual}). Existence theorems are
acknowledged. Trainability is the working subject.

% --- END §1 ---

\section{Backpropagation}
\label{sec:nn-backprop}

\subsection{The Problem}

Chapter~2 told us how to update parameters once we have a
gradient: stochastic gradient descent and its momentum-and-Adam
descendants take a noisy estimate of $\nabla_\theta \mathcal{L}$
and step against it. Chapter~2 did not tell us how to compute
that gradient. The model in question has, on the small side, a
hundred million parameters; on the large side, a trillion. A
direct numerical estimate --- perturb each parameter, re-evaluate
the loss, take the difference --- would cost $P$ forward passes
per gradient, each forward pass itself costing $O(P)$ work, for a
total of $O(P^2)$ FLOPs per training step. At $P = 10^{11}$ this
is $10^{22}$ FLOPs to compute one gradient. The training run
would never finish.

\emph{Backpropagation} is the algorithm that brings this cost down
to $O(P)$ --- the same order as a single forward pass --- by
storing the intermediate activations from the forward pass and
walking the chain rule of calculus backwards through the
computation graph. \citet{werbos1974beyond} wrote down a
recognisable version of the method in his Harvard PhD thesis;
\citet{rumelhart1986learning} brought it to the field's attention
with a Nature paper that triggered the modern revival of neural
networks. The mechanics have not changed since.

\subsection{The Calculus Chain Rule, Briefly}

The calculus chain rule states that if a real-valued scalar
$\mathcal{L}$ is a function of an intermediate vector $z$ which
in turn depends on a parameter $\theta$, the derivative of
$\mathcal{L}$ with respect to $\theta$ is the inner product of the
upstream gradient and the local Jacobian:
\begin{equation}
  \frac{\partial \mathcal{L}}{\partial \theta}
  \;=\;
  \frac{\partial \mathcal{L}}{\partial z}
  \cdot
  \frac{\partial z}{\partial \theta}.
  \label{eq:nn-chain-rule}
\end{equation}
If $z$ is itself produced by a longer chain, the rule iterates:
each link in the chain multiplies the upstream gradient by the
local Jacobian of that link's output with respect to its input.
This is the calculus chain rule, distinct from (though sharing a
name with) the probability chain rule of Chapter~1, as flagged in
the notation block above.

\subsection{A Two-Layer Derivation, by Hand}

Concrete first; general second. Take a two-layer MLP with scalar
input $x \in \R^{d_0}$, hidden width $d_1$, scalar output, and
loss $\mathcal{L}$ that depends only on the output:
\begin{align*}
  z^{(1)} &\;=\; W^{(1)} x + b^{(1)}, \\
  h^{(1)} &\;=\; \varphi\!\bigl(z^{(1)}\bigr), \\
  z^{(2)} &\;=\; W^{(2)} h^{(1)} + b^{(2)}, \\
  \mathcal{L} &\;=\; \mathcal{L}(z^{(2)}; y).
\end{align*}
We want $\partial \mathcal{L} / \partial W^{(\ell)}$ and
$\partial \mathcal{L} / \partial b^{(\ell)}$ for $\ell \in
\{1, 2\}$. Walk the chain rule outward from $\mathcal{L}$ through
the output layer first.

\textbf{Output layer.} Let
$\delta^{(2)} := \partial \mathcal{L} / \partial z^{(2)}$ ---
the gradient of the loss with respect to the output
\emph{pre-activation}. For a regression loss
$\mathcal{L} = \tfrac{1}{2}\,(z^{(2)} - y)^2$ this is just
$\delta^{(2)} = z^{(2)} - y$; for a softmax-cross-entropy head it
is $\delta^{(2)} = \mathrm{softmax}(z^{(2)}) - \mathrm{onehot}(y)$;
the specific form depends on the loss. Once we have $\delta^{(2)}$,
the parameter gradients of the output layer follow from the chain
rule applied to $z^{(2)} = W^{(2)} h^{(1)} + b^{(2)}$:
\[
  \frac{\partial \mathcal{L}}{\partial W^{(2)}}
  \;=\; \delta^{(2)} \, h^{(1)\top},
  \qquad
  \frac{\partial \mathcal{L}}{\partial b^{(2)}}
  \;=\; \delta^{(2)}.
\]

\textbf{Hidden layer.} Now propagate the gradient back to the
hidden pre-activation $z^{(1)}$. By the chain rule,
\[
  \delta^{(1)}
  \;:=\; \frac{\partial \mathcal{L}}{\partial z^{(1)}}
  \;=\; \frac{\partial \mathcal{L}}{\partial h^{(1)}}
        \odot \varphi'\!\bigl(z^{(1)}\bigr)
  \;=\; \bigl(W^{(2)\top} \delta^{(2)}\bigr)
        \odot \varphi'\!\bigl(z^{(1)}\bigr),
\]
where the first equality applies the chain rule through the
elementwise activation (so the local Jacobian of $\varphi$ is the
diagonal matrix $\mathrm{diag}(\varphi'(z^{(1)}))$, and matrix
multiplication by a diagonal becomes elementwise multiplication
$\odot$), and the second equality applies the chain rule through
the linear map $z^{(2)} = W^{(2)} h^{(1)} + b^{(2)}$ to express
$\partial \mathcal{L} / \partial h^{(1)}$ as $W^{(2)\top}
\delta^{(2)}$. With $\delta^{(1)}$ in hand, the hidden-layer
parameter gradients have the same shape as the output-layer ones:
\[
  \frac{\partial \mathcal{L}}{\partial W^{(1)}}
  \;=\; \delta^{(1)} \, x^\top,
  \qquad
  \frac{\partial \mathcal{L}}{\partial b^{(1)}}
  \;=\; \delta^{(1)}.
\]

The structural lesson is worth flagging. The gradient of
$\mathcal{L}$ with respect to the parameters of layer $\ell$
depends on two things: the upstream gradient $\delta^{(\ell)}$
(arriving from the layer above) and the post-activation of the
layer below ($h^{(\ell-1)}$, with $h^{(0)} = x$). The same
$\delta^{(\ell)}$ is what the next-deeper layer needs to compute
its own $\delta^{(\ell-1)}$. The forward pass produces and stores
all the $h^{(\ell)}$ and $z^{(\ell)}$; the backward pass walks
from $\delta^{(L)}$ down to $\delta^{(1)}$, computing parameter
gradients as it goes.

\subsection{The General Multi-Layer Pattern}

The two-layer derivation generalises directly. For an $L$-layer
MLP with the composition of \eqref{eq:nn-mlp}, fix the loss
$\mathcal{L}$ and define $\delta^{(\ell)} := \partial \mathcal{L}
/ \partial z^{(\ell)}$ as the gradient of the loss with respect
to the layer-$\ell$ pre-activation. The recursion is
\begin{equation}
  \delta^{(L)}
  \;=\; \frac{\partial \mathcal{L}}{\partial z^{(L)}},
  \qquad
  \delta^{(\ell-1)}
  \;=\; \bigl(W^{(\ell)\top} \delta^{(\ell)}\bigr)
        \odot \varphi'\!\bigl(z^{(\ell-1)}\bigr),
  \quad \ell = L, L-1, \dots, 2,
  \label{eq:nn-delta-recursion}
\end{equation}
and the parameter gradients at every layer are
\begin{equation}
  \frac{\partial \mathcal{L}}{\partial W^{(\ell)}}
  \;=\; \delta^{(\ell)} \, h^{(\ell-1)\top},
  \qquad
  \frac{\partial \mathcal{L}}{\partial b^{(\ell)}}
  \;=\; \delta^{(\ell)}.
  \label{eq:nn-param-grads}
\end{equation}
The final-layer term $\delta^{(L)} = \partial \mathcal{L} /
\partial z^{(L)}$ is whatever the loss says: the regression and
softmax cross-entropy forms above are the two cases this book
will use repeatedly.

Read as an algorithm, this is two passes. The forward pass
computes and stores $z^{(\ell)}$ and $h^{(\ell)}$ for every
$\ell$. The backward pass starts at $\delta^{(L)}$, computes the
output-layer parameter gradients via \eqref{eq:nn-param-grads},
recurses down to $\delta^{(L-1)}$ via \eqref{eq:nn-delta-recursion},
computes the layer-$L{-}1$ parameter gradients, and so on until
all layers have been visited. Each parameter is touched exactly
once on the backward pass, mirroring the forward pass.

\subsection{Forward / Backward FLOP Equivalence}

Counting FLOPs makes the cost structure explicit. Each layer's
forward pass is dominated by the matrix multiply $W^{(\ell)}
h^{(\ell-1)}$, costing $2 d_\ell d_{\ell-1}$ FLOPs (one MAC per
weight-input pair, two FLOPs per MAC). Summed across layers, the
forward pass costs roughly $2 P$ FLOPs per training token, where
$P$ is the parameter count.

The backward pass does two matrix multiplies per layer. The
first, $W^{(\ell)\top} \delta^{(\ell)}$, propagates the gradient
to the layer below; this costs another $2 d_\ell d_{\ell-1}$
FLOPs per layer, or $2 P$ FLOPs in total. The second,
$\delta^{(\ell)} h^{(\ell-1)\top}$, computes the parameter
gradient at this layer; this costs another $2 P$ FLOPs in total.
The backward pass therefore costs roughly $4 P$ FLOPs per
training token, twice the forward pass.

A full training step --- one forward pass to compute activations
and loss, plus one backward pass to compute gradients --- costs
roughly $2 P + 4 P = 6 P$ FLOPs per training token. With $D$
tokens of training data and one pass over them, the total
training compute is $\approx 6 P D$ FLOPs. \emph{This is the
``$6 N D$'' rule that Chapter~2 used without justification when
deriving the Chinchilla compute-optimal point}: the factor of six
is exactly the three MACs per parameter per token (one forward,
two backward) times two FLOPs per MAC. Backpropagation is what
makes the rule cheap enough to be worth applying.

% --- END §2 ---

\section{Activations}
\label{sec:nn-activations}

\subsection{Why We Need Nonlinearity at All}

The activation function $\varphi$ in \eqref{eq:nn-mlp} is not a
decoration. Without it, the network collapses to a single linear
map regardless of depth. The argument is one line. Compose two
affine layers without an elementwise nonlinearity between them:
\[
  W^{(2)}\bigl(W^{(1)} x + b^{(1)}\bigr) + b^{(2)}
  \;=\; \bigl(W^{(2)} W^{(1)}\bigr) x
       + \bigl(W^{(2)} b^{(1)} + b^{(2)}\bigr).
\]
The right-hand side is itself an affine function $W' x + b'$ with
$W' = W^{(2)} W^{(1)}$ and $b' = W^{(2)} b^{(1)} + b^{(2)}$. The
two-layer network has the same expressive power as a one-layer
network. The same collapse extends to any depth: a stack of $L$
purely affine layers is equivalent to a single affine layer.
\emph{Depth without nonlinearity buys nothing.} The choice of
$\varphi$ --- a strictly elementwise, strictly nonlinear function ---
is what makes deep networks more expressive than shallow ones.

This section walks the historical progression of activation
choices from the sigmoid of the 1980s to the SwiGLU of today's
Transformers. The progression is not arbitrary: each newer choice
was a response to a specific failure of the one before it.

\subsection{Sigmoid and Tanh: the Vanishing-Gradient Origin Story}

The activations of the connectionist era of the 1980s were the
logistic sigmoid and the hyperbolic tangent:
\[
  \sigma(z) \;=\; \frac{1}{1 + e^{-z}},
  \qquad
  \tanh(z) \;=\; \frac{e^{z} - e^{-z}}{e^{z} + e^{-z}}.
\]
Their derivatives are well-behaved at $z = 0$ ---
$\sigma'(0) = 1/4$, $\tanh'(0) = 1$ --- but they saturate as
$|z|$ grows. For the sigmoid, $\sigma'(z) = \sigma(z)(1 - \sigma(z))$,
which approaches zero for both large positive and large negative
$z$. For tanh, $\tanh'(z) = 1 - \tanh^2(z)$, with the same
saturation behaviour. A unit whose pre-activation has drifted
into either tail has a derivative within rounding distance of
zero.

This becomes a problem at depth. The backprop $\delta$-recursion
of \eqref{eq:nn-delta-recursion} multiplies the upstream gradient
by $\varphi'(z^{(\ell-1)})$ at every layer. If $|\varphi'| \le c < 1$
at most pre-activations and the network has $L$ layers, the
gradient at layer $1$ is bounded by $c^{L}$ times whatever
arrived at layer $L$. With sigmoidal activations and even
modestly large pre-activations, $c^{L}$ becomes
indistinguishable from zero for $L$ in the dozens. The early
layers stop learning. This is the \emph{vanishing-gradient
problem}, and it kept neural networks effectively shallow for
the better part of two decades.

\subsection{ReLU: the Breakthrough}

The Rectified Linear Unit, advocated for deep networks by
\citet{nair2010rectified}, is brutally simple:
\begin{equation}
  \mathrm{ReLU}(z) \;=\; \max(0,\, z),
  \qquad
  \mathrm{ReLU}'(z) \;=\;
  \begin{cases}
    1 & z > 0 \\
    0 & z < 0,
  \end{cases}
  \label{eq:nn-relu}
\end{equation}
with $\varphi'(0)$ taken by convention to be $0$ (per A3.1). On
the positive half-plane the derivative is exactly one ---
unsaturated, no decay --- so the gradient signal passes through
without attenuation. The Rumelhart-era saturation problem
disappears in the half of input space where the unit is active.

ReLU's defect is the symmetric one. On the negative half-plane
the derivative is exactly zero. A unit whose pre-activation is
permanently negative --- because the upstream activations and
the chosen weights drive its $z$ below zero on every training
example --- contributes zero gradient on every step, so its
weights never update, so its pre-activation stays negative. This
is the \emph{dead-neuron problem}. A network may begin training
with all units alive and end training with a substantial fraction
permanently dead. The remedy at the practitioner level is
careful initialisation (§\ref{sec:nn-init} below) and learning
rates that are not so aggressive that they push large fractions
of units into the dead region in a single update.

Variants ameliorate the failure mode. \emph{Leaky ReLU} replaces
the zero in the negative half with a small slope $\alpha z$ for
$\alpha \approx 0.01$, so the gradient is small but nonzero. The
\emph{exponential linear unit} (ELU) makes the same move with an
exponential curve. These are useful, never universal. The next
move was to smooth the kink itself.

\subsection{GeLU: Smoothed ReLU}

The Gaussian Error Linear Unit, due to
\citet{hendrycks2016gelu}, replaces ReLU's hard threshold with a
smooth gating by the standard normal CDF:
\begin{equation}
  \mathrm{GeLU}(z) \;=\; z \cdot \Phi(z),
  \qquad
  \Phi(z) \;=\; \tfrac{1}{2}\!\left(1 + \mathrm{erf}\!\bigl(z/\sqrt{2}\bigr)\right).
  \label{eq:nn-gelu}
\end{equation}
For computational convenience, the
$\tanh$-based approximation
$\mathrm{GeLU}(z) \approx z \cdot \tfrac{1}{2}\bigl(1 + \tanh(\sqrt{2/\pi}(z + 0.044715\,z^3))\bigr)$
is widely used in practice. The shape is familiar: GeLU
asymptotes to ReLU for large positive $z$, asymptotes to zero
for large negative $z$, and bends smoothly through the origin
with a derivative that is continuous everywhere. There is no
dead-neuron region in the strict sense: a unit with a slightly
negative pre-activation has a small but nonzero gradient and can
recover. GeLU is the default activation in BERT, the GPT line
through GPT-4, and most decoder-only Transformers released
through 2023.

\subsection{GLU and SwiGLU: Gated Variants (Preview)}

Modern Transformers do not stop at GeLU. Following
\citet{shazeer2020glu}, the feed-forward block now typically uses
a \emph{gated} variant in which the activation is itself a
function of two parallel linear projections:
\begin{equation}
  \mathrm{GLU}(x) \;=\; \bigl(W_g x\bigr) \odot \varphi\!\bigl(W_u x\bigr),
  \label{eq:nn-glu}
\end{equation}
where $\varphi$ is the elementwise activation (Swish, GeLU, etc.)
and $\odot$ is the elementwise (Hadamard) product. The
\emph{SwiGLU} variant takes $\varphi$ to be Swish, defined as
$\mathrm{Swish}(z) = z \cdot \sigma(z)$. SwiGLU is the activation
used inside the FFN block of Llama, PaLM, and most modern
decoder-only models. Chapter~4's discussion of the Transformer
feed-forward block derives the parameter count, FLOP cost, and
$d_f$ rescaling that SwiGLU implies; here we simply name the
form so that Chapter~4 has the primitive ready when it composes
the FFN.

\subsection{A Practitioner's Table}

The choice of activation is not a derivation; it is a design
decision informed by the depth, the data, and the surrounding
architecture. The table below summarises the choice that is
defensible by default in each context the rest of the book will
consider.

\begin{center}
\begin{tabular}{|l|l|l|}
\hline
\textbf{Context} & \textbf{Default activation} & \textbf{Why} \\
\hline
Output, binary classification        & $\sigma$ (logistic sigmoid) & Maps $\R \to (0, 1)$. \\
Output, $K$-way classification        & softmax ($\varphi$ at last layer)  & Probability simplex over $K$ classes. \\
Output, regression                    & identity (no activation)    & Real-valued target. \\
Hidden, recurrent gates              & $\sigma$, $\tanh$ (specific roles) & Gates require bounded outputs. \\
Hidden, modern feed-forward (vision) & ReLU                        & Cheap, well-understood, good with He init. \\
Hidden, Transformer FFN (language)   & GeLU or SwiGLU               & Smooth derivative; SwiGLU adds gating. \\
\hline
\end{tabular}
\end{center}

The historical progression --- sigmoid through ReLU through GeLU
through SwiGLU --- traces the same engineering arc as the rest
of the chapter: each newer choice was a response to a specific
failure of the one before it, and the modern Transformer is the
result of that arc converging on a small handful of robust
defaults.

% --- END §3 ---

\section{Weight Initialization}
\label{sec:nn-init}

\subsection{Why Initialisation Matters}

A neural network with the wrong initial weights does not train.
Two failure modes show up at depth, and they are mirror images of
each other.

\textbf{Variance collapse.} If the weights start too small, the
activations shrink toward zero as they propagate forward through
the layers, and so do the gradients as they propagate backward
through them. By the time the loss reaches the output layer the
activations are uniform; by the time the gradient reaches the
input layer it is rounding noise. The early layers do not
receive a useful gradient and never learn.

\textbf{Variance explosion.} If the weights start too large, the
activations grow exponentially in depth, the pre-activations
saturate any bounded nonlinearity (sigmoid, tanh) or push past
the range of floating-point arithmetic, and gradients spike to
NaN. The training run fails on the first few steps.

The middle ground --- weights chosen so that the
activation variance is preserved layer-to-layer at initialisation
--- is what \citet{glorot2010understanding} and
\citet{he2015delving} formalised. Their derivations are short and
worth working through.

\subsection{The Variance-Preservation Argument}

We want, at initialisation, $\Var(h^{(\ell)}) \approx
\Var(h^{(\ell-1)})$ across all layers. Consider a single linear
layer $z = W h$ with $W \in \R^{d_{\mathrm{out}} \times
d_{\mathrm{in}}}$. \Assum\ The weights $W_{ij}$ are drawn i.i.d.\
with $\E[W_{ij}] = 0$ and $\Var(W_{ij}) = \tau^2$, the input
components $h_j$ are i.i.d.\ with $\E[h_j] = 0$ and
$\Var(h_j) = \nu^2$, and $W$ and $h$ are independent of each
other. Then for the $i$-th output component
\[
  z_i \;=\; \sum_{j=1}^{d_{\mathrm{in}}} W_{ij}\, h_j,
  \qquad
  \E[z_i] \;=\; 0,
  \qquad
  \Var(z_i)
  \;=\; \sum_{j=1}^{d_{\mathrm{in}}} \Var(W_{ij}\, h_j)
  \;=\; d_{\mathrm{in}}\, \tau^2\, \nu^2,
\]
where we used $\Var(W_{ij} h_j) = \E[W_{ij}^2]\E[h_j^2] = \tau^2
\nu^2$ under independence and zero mean. Variance preservation
$\Var(z_i) = \Var(h_j)$ then requires
\begin{equation}
  \tau^2 \;=\; \frac{1}{d_{\mathrm{in}}}.
  \label{eq:nn-var-preserve}
\end{equation}
This is the headline. The variance of the linear output is the
input variance times $d_{\mathrm{in}}$ times the weight variance,
so to keep variance constant across the layer we set the weight
variance to $1/d_{\mathrm{in}}$.

\subsection{Xavier / Glorot Initialisation}

\citet{glorot2010understanding} observed that
\eqref{eq:nn-var-preserve} preserves variance on the
\emph{forward} pass, but the backward pass also needs to preserve
the gradient variance, and the relevant fan-in for the gradient
is the layer's $d_{\mathrm{out}}$ rather than $d_{\mathrm{in}}$.
Balancing the two goals gives the Xavier / Glorot initialiser
\begin{equation}
  \Var(W_{ij}) \;=\; \frac{2}{d_{\mathrm{in}} + d_{\mathrm{out}}},
  \label{eq:nn-xavier}
\end{equation}
typically realised by drawing $W_{ij}$ from $\mathcal{N}(0,
\Var(W))$ or from a uniform distribution with the same variance.
The argument assumes the activation $\varphi$ is symmetric around
zero with derivative roughly $1$ near the origin --- which is
true for tanh and roughly true for the logistic sigmoid scaled
appropriately. Glorot init was the first variance-preserving
initialiser to make plain feed-forward depth (without
pre-training) work reliably; it ended the layer-wise pre-training
era documented in §\ref{sec:nn-function-class}.

\subsection{He / Kaiming Initialisation}

The Glorot derivation breaks for ReLU activations. ReLU zeros
half of the input distribution in expectation, which halves the
output variance:
\[
  \Var(\mathrm{ReLU}(z)) \;=\; \tfrac{1}{2}\, \Var(z),
\]
under the assumption that $z$ is symmetric around zero (a
reasonable approximation at initialisation). To compensate,
\citet{he2015delving} doubled the weight variance:
\begin{equation}
  \Var(W_{ij}) \;=\; \frac{2}{d_{\mathrm{in}}},
  \label{eq:nn-he-init}
\end{equation}
which is variance-preserving across the linear-then-ReLU
combination. He init is the modern default for ReLU-family
networks and is the initialiser cited (typically by name) in the
LoRA setup of Chapter~7. The factor of two is the only
difference from a naive $1/d_{\mathrm{in}}$; the larger lesson is
that the right init depends on the activation, not just the
linear layer.

\subsection{Bridge to Chapter 4: the $1/\sqrt{d_k}$ Attention Scale}

The variance-preservation argument used here will reappear in
Chapter~4 in the context of attention. The dot product $q^\top k$
of a query and a key vector, both with $d_k$ components and
independent zero-mean unit-variance entries, has variance $d_k$
and standard deviation $\sqrt{d_k}$. Dividing by $\sqrt{d_k}$
restores unit variance, exactly the same logic that gave us
$\tau^2 = 1/d_{\mathrm{in}}$ above (square-root rather than full
inverse because we are scaling the standard deviation rather
than the variance). Chapter~4 derives this for the specific case
of attention scores; this chapter has given the general form. If
you find the $1/\sqrt{d_k}$ unmotivated when you reach
Chapter~4, return here.

% --- END §4 ---

\section{Normalization}
\label{sec:nn-norm}

\subsection{Why Normalize at All}

A network with sensible initialisation (\S\ref{sec:nn-init})
starts with activations whose variance is roughly preserved
across layers. Training does not preserve that property. Each SGD
step changes every weight in the network, and the activation
distribution at every layer drifts in response. By the time the
optimiser has taken a few thousand steps, the distribution that
each layer was implicitly tuned for has shifted underneath it.
The original framing called this \emph{internal covariate shift}
\citep{ioffe2015batchnorm}; the modern view treats the
mechanistic claim with caution but keeps the operational
observation. The activations drift, and the drift is what
normalisation layers are designed to undo.

A normalisation layer re-establishes a target activation
distribution at every layer of the network, on every forward
pass, regardless of how training has perturbed the weights so
far. Three benefits follow. The variance of activations stays
near unit at every depth, recovering the property that
initialisation gave us at step zero. Layer-wise effective
learning rates decouple from each other, because the input scale
to every layer is now standardised. Downstream layers see a
stationary distribution from upstream, even when upstream weights
are still moving fast. The result, in practice, is that deeper
networks are trainable with larger learning rates and less
hyperparameter babysitting.

\subsection{BatchNorm, Briefly}

The first widely-adopted normalisation layer was \emph{batch
normalisation} \citep{ioffe2015batchnorm}. It normalises each
feature \emph{across the batch}: for a feature index $j$ and a
batch of $B$ examples with pre-activations $\{z^{(1)}_j, \dots,
z^{(B)}_j\}$, BatchNorm computes the batch mean
$\mu_j = \tfrac{1}{B}\sum_b z^{(b)}_j$ and batch variance
$\sigma_j^2 = \tfrac{1}{B}\sum_b (z^{(b)}_j - \mu_j)^2$, then
emits $\gamma_j (z^{(b)}_j - \mu_j) / \sqrt{\sigma_j^2 +
\epsilon} + \beta_j$ for each example. Trainable scale $\gamma_j$
and shift $\beta_j$ recover whatever distribution downstream
layers prefer. At inference time, the batch statistics are
replaced by running averages collected during training.

BatchNorm is the right tool for vision (large mini-batches,
fixed-shape inputs, no ragged sequences). It is the wrong tool
for the autoregressive Transformers of this book. Three reasons.
First, language models pack variable-length sequences into a
fixed batch shape; the batch-statistic dependency interacts
poorly with that packing. Second, per-token effective batch
sizes inside a Transformer are large in tokens but small in
\emph{independent} examples (all tokens within a sequence share
correlated structure), and BatchNorm's variance estimate is
biased under correlation. Third, the train/eval discrepancy
between mini-batch statistics and running averages is fragile at
the scale and decoding regimes used in production. The field
moved away from BatchNorm in language modelling and has not
returned.

\subsection{LayerNorm}

\citet{ba2016layernorm} replaced the batch dimension with the
feature dimension. \emph{Layer normalisation} normalises each
\emph{individual} activation vector across its own features.
For a vector $x \in \R^d$,
\begin{equation}
  \mathrm{LN}(x) \;=\; \gamma \odot \frac{x - \mu(x)}{\sigma(x) + \epsilon} + \beta,
  \qquad
  \mu(x) = \tfrac{1}{d}\sum_j x_j,
  \quad
  \sigma(x)^2 = \tfrac{1}{d}\sum_j \bigl(x_j - \mu(x)\bigr)^2,
  \label{eq:nn-layernorm}
\end{equation}
with trainable per-feature scale $\gamma \in \R^d$ and shift
$\beta \in \R^d$. The mean and variance are computed over the
features of the single vector $x$; nothing about the batch enters
the computation. The same operation runs at training time and at
inference time without modification, removing the train/eval
discrepancy that BatchNorm has to manage with running averages.
LayerNorm has no batch dependency, no sequence-length sensitivity,
and no packing-friendliness penalty. It became the default in
sequence models and has been the workhorse normaliser of the
Transformer era.

\subsection{RMSNorm}

\citet{zhang2019rmsnorm} observed that the mean-centering term in
\eqref{eq:nn-layernorm} contributes little empirically: removing
it produces a layer that trains as well as LayerNorm does at
roughly seven percent lower compute per call. \emph{Root mean
square normalisation} drops the mean entirely:
\begin{equation}
  \mathrm{RMSNorm}(x) \;=\; \gamma \odot \frac{x}{\mathrm{RMS}(x) + \epsilon},
  \qquad
  \mathrm{RMS}(x) \;=\; \sqrt{\tfrac{1}{d}\sum_j x_j^2}.
  \label{eq:nn-rmsnorm}
\end{equation}
RMSNorm has a single trainable per-feature scale $\gamma$ and no
shift $\beta$, halving the parameter count of the layer compared
to LayerNorm. The operation rescales the vector to unit
root-mean-square magnitude in expectation, leaving the angular
direction unchanged. Llama, Mistral, and most open-weight models
released since 2023 use RMSNorm in place of LayerNorm. The two
are interchangeable from the perspective of the rest of this
book: the chapters that follow refer to ``the normaliser'' when
either choice is acceptable, and to LayerNorm or RMSNorm
specifically when the distinction matters.

\subsection{Bridge to Chapter 4}

This section has established the normalisers themselves.
Chapter~4 takes them up in two further directions. First,
\emph{placement}: should the normalisation be applied
\emph{before} or \emph{after} the residual sum at each
Transformer block (the \emph{pre-norm} versus \emph{post-norm}
distinction)? Chapter~4's argument is that pre-norm keeps the
residual path an identity at initialisation, which makes deep
stacks trainable without elaborate warmup; this is the
configuration used by GPT-2 onwards and almost every modern
LLM. Second, \emph{interpretation}: in a pre-norm Transformer the
residual connection carries a $d$-dimensional ``stream'' that
attention and feed-forward sublayers \emph{read from} (via the
normaliser) and \emph{write back into} (via the residual sum).
The normaliser, in that view, is what gives each sublayer a
clean, standardised view of the residual stream's current
state. Chapter~4 develops both points; the apparatus this
section has provided is what they sit on.

% --- END §5 ---

\section{Residual Connections}
\label{sec:nn-residual}

\subsection{The Depth Problem}

A plain MLP, even with the right activation
(\S\ref{sec:nn-activations}) and the right initialisation
(\S\ref{sec:nn-init}) and the right normalisation
(\S\ref{sec:nn-norm}), still refuses to train above some depth.
The empirical record of the period before 2015 was that as
network depth increased past a few dozen layers, training loss
\emph{stopped going down}. \citet{he2016resnet} called this the
\emph{degradation problem} and were careful to distinguish it
from overfitting: the deeper network's training loss was higher
than the shallower one's, so the failure was not memorisation
without generalisation, it was a failure to optimise. A
forty-layer network was unable to find a parameter setting that
matched the twenty-layer network on data the deeper network
should have been strictly more expressive on. The optimiser was
losing the gradient signal somewhere in the intervening layers,
and no amount of init or normalisation tweaking entirely
recovered it.

\subsection{The Residual Response}

The fix \citet{he2016resnet} proposed has a minimal-shock-to-the-system
character. Replace each layer's transformation
$h^{(\ell)} = F^{(\ell)}(h^{(\ell-1)})$ with
\begin{equation}
  h^{(\ell)} \;=\; h^{(\ell-1)} \;+\; F^{(\ell)}\!\bigl(h^{(\ell-1)}\bigr),
  \label{eq:nn-residual}
\end{equation}
where $F^{(\ell)}$ is whatever sublayer the architecture would
otherwise have used (typically a small MLP, or an attention block
in the Transformer case). The layer no longer computes a fresh
output; it computes a \emph{residual} that is added to the layer
below. At initialisation, when the parameters of $F^{(\ell)}$
draw weights small enough that the layer's output is near zero,
\eqref{eq:nn-residual} becomes $h^{(\ell)} \approx h^{(\ell-1)}$:
the layer is the identity map, and the network behaves at depth
$L$ as if it had only the depth at which the residual blocks
have already learned something useful. The optimiser is free to
learn \emph{small perturbations} on top of the identity rather
than learning the full transformation outright.

\subsection{Why This Makes Depth Trainable}

Two arguments, one about the gradient and one about the
optimisation landscape, explain why the change was decisive.

\textbf{Gradient flow.} Differentiating \eqref{eq:nn-residual}
with respect to the layer below,
\[
  \frac{\partial h^{(\ell)}}{\partial h^{(\ell-1)}}
  \;=\; I \;+\; \frac{\partial F^{(\ell)}}{\partial h^{(\ell-1)}}.
\]
The gradient signal flowing backwards from layer $\ell$ to layer
$\ell-1$ has an identity term --- it does not have to pass
through $F^{(\ell)}$ to survive. In a plain stack the same
derivative is just $\partial F^{(\ell)} / \partial h^{(\ell-1)}$,
which can have norm well below one and which compounds
multiplicatively across layers; the residual form keeps the
identity term in play at every layer, so the gradient at depth
$1$ is bounded below regardless of what the intermediate $F$
blocks do. Vanishing gradients (\S\ref{sec:nn-activations}) lose
their bite when the gradient has a guaranteed straight-shot path
back through the identity.

\textbf{Optimisation floor.} A residual network initialised with
near-zero $F$ blocks computes the same function as a shallower
network at the start of training. The optimiser cannot do
\emph{worse} than the shallow network, because finding $F = 0$
is always available. The empirical contribution of
\citet{he2016resnet} was that this floor turns into a ceiling
that grows monotonically with depth: networks with hundreds of
residual layers trained to lower loss than networks with tens of
plain ones. The architectural move solved the optimisation
problem, not the expressive-power problem; the existence
theorems of §\ref{sec:nn-function-class} had always said the
deep networks \emph{could} compute these functions, but plain
stacks were unable to find the parameters in practice.

\subsection{The Continuous-Depth View, in One Paragraph}

There is a clean limit to take. Treat the depth index $\ell$ as a
continuous variable $t$ and let the per-layer transformation
shrink as the depth grows. The residual recursion
\eqref{eq:nn-residual} becomes the Euler-discretisation of an
ordinary differential equation $dh/dt = F_t(h)$, with the
network's forward pass running an integrator from $t = 0$ to
$t = 1$ and the network's parameters parameterising the vector
field $F_t$. Neural ODEs \citep{chen2018neuralode} make this
limit explicit; they replace the discrete stack with an adaptive
integrator and learn the vector field directly. The view is
intellectually clean and operationally rare: most LLMs, and all
the Transformers in the rest of this book, use the discrete
residual form of \eqref{eq:nn-residual} rather than the
continuous-depth limit. We mention the connection here because
it explains why residual networks behave well at depth ---
solving an ODE numerically is well-conditioned in a way that
composing arbitrary layers is not --- not because the next ten
chapters require it.

\subsection{Bridge to Chapter 4: the Residual Stream}

The Transformer of Chapter~4 is built around residual connections
of the form \eqref{eq:nn-residual}. A Transformer block is two
residual sublayers stacked: an attention block and a feed-forward
block, each wrapped in a normaliser (\S\ref{sec:nn-norm}) and a
residual sum. The accumulation of all these residual additions
across the stack is what Chapter~4 calls the \emph{residual
stream}: a $d$-dimensional vector that propagates from the
embedding layer to the unembedding, with attention and feed-forward
sublayers reading from it (via the normaliser) and writing back
into it (via the residual add). The interpretability framing of
\citet{elhage2021mathematical}, which Chapter~4 develops, treats
this stream as the model's working memory and individual heads
as components that read specific subspaces and write into others.
The residual connection introduced here is the substrate that
makes that view possible.

% --- END §6 ---

\section{The MLP Block as a Transformer Primitive}
\label{sec:nn-mlp-block}

The pieces this chapter has assembled --- linear layers, an
activation choice, a variance-preserving initialiser, a
normaliser, and a residual connection --- compose into the
\emph{feed-forward block} that sits at the heart of every
Transformer layer. We will not derive the composition here;
that is Chapter~4's job. We will name it, so the reader knows
what is coming.

\subsection{The FFN}

Inside a Transformer block, between two attention sublayers, each
token is processed independently by a small two-layer MLP. The
canonical form is
\begin{equation}
  \mathrm{FFN}(x) \;=\; W_2 \, \varphi\!\bigl(W_1 x + b_1\bigr) + b_2,
  \qquad
  W_1 \in \R^{d_f \times d},
  \quad
  W_2 \in \R^{d \times d_f},
  \label{eq:nn-ffn}
\end{equation}
where $d$ is the token's embedding dimension and $d_f$ is the
hidden width of the feed-forward block, typically $d_f = 4d$ for
classical activations or $d_f = \tfrac{8}{3} d$ for SwiGLU
variants (so total parameters stay roughly fixed when the third
projection of SwiGLU is added; \S\ref{sec:nn-activations}). The
``position-wise'' qualifier means \eqref{eq:nn-ffn} is applied to
every token's representation independently, with the same
$(W_1, W_2, b_1, b_2)$ shared across positions. The block is
just a two-layer MLP of \eqref{eq:nn-mlp}, instantiated at every
position with shared weights.

\subsection{Why the FFN Dominates the Parameter Count}

A rough accounting tells the story. The attention sublayer has
four projection matrices $W_Q, W_K, W_V, W_O$, each of shape
$d \times d$, contributing $4 d^2$ parameters per layer. The
feed-forward block has two matrices of shape $d_f \times d = 4d
\times d$, contributing $2 \cdot 4 d^2 = 8 d^2$ parameters per
layer. Twelve total at $4d^2$ per matrix unit, eight of which
sit in the FFN: roughly two-thirds of a Transformer block's
parameters. The exact ratio depends on the architectural choices
(GQA reduces attention parameters; SwiGLU adds a third FFN
projection), but the headline does not change. \emph{The FFN is
where the Transformer's parameter budget lives.} Chapter~4 walks
the FLOP and parameter accounting in full; this section flags the
result so that adaptation discussions later in the book (LoRA in
Chapter~7, MoE in Chapter~6) make sense in advance.

\subsection{The Handoff to Chapter 4}

By the end of this chapter you have:

\begin{itemize}
  \item Linear layers and the multilayer perceptron
        (\S\ref{sec:nn-function-class}, Eq.~\ref{eq:nn-mlp});
  \item Backpropagation, the $\delta$-recursion, and the
        forward/backward FLOP equivalence underwriting
        Chapter~2's $6ND$ rule
        (\S\ref{sec:nn-backprop},
        Eqs.~\ref{eq:nn-delta-recursion}--\ref{eq:nn-param-grads});
  \item The progression of activations from sigmoid to SwiGLU,
        and the design rules for picking among them
        (\S\ref{sec:nn-activations}, Eqs.~\ref{eq:nn-relu},
        \ref{eq:nn-gelu}, \ref{eq:nn-glu});
  \item Variance-preserving initialisation (Xavier and He) and
        the bridge to attention's $1/\sqrt{d_k}$ scaling
        (\S\ref{sec:nn-init}, Eqs.~\ref{eq:nn-xavier},
        \ref{eq:nn-he-init});
  \item LayerNorm and RMSNorm, and the rationale for using them
        instead of BatchNorm
        (\S\ref{sec:nn-norm}, Eqs.~\ref{eq:nn-layernorm},
        \ref{eq:nn-rmsnorm});
  \item Residual connections and the gradient-flow and
        optimisation-floor arguments that explain why deep
        residual stacks train where deep plain stacks did not
        (\S\ref{sec:nn-residual}, Eq.~\ref{eq:nn-residual}).
\end{itemize}

Chapter~4 will compose these pieces, add the attention
mechanism, derive the $1/\sqrt{d_k}$ scaling using the
variance-preservation argument from \S\ref{sec:nn-init},
combine attention and FFN into a single Transformer block via
two residual sublayers, choose between LayerNorm and RMSNorm and
between pre-norm and post-norm placement, and walk the FLOP and
KV-cache accounting that follows. The Transformer of Chapter~4
is the arrangement of pieces this chapter has just laid on the
bench.

% --- END §7 ---

\section{Worked Examples}
\label{sec:nn-worked-examples}

The chapter's two worked examples are paired. The first
demonstrates the machinery on the simplest possible problem ---
how a forward pass produces activations, how the backward pass
produces gradients, how SGD moves weights. The second
demonstrates why that machinery is necessary: the same forward /
backward step on the canonical Minsky--Papert problem, where the
two-layer network can represent what a one-layer network cannot.

\subsection{Worked Example A: 1-D Regression (Mechanics)}

\begin{workedexample}[A: One forward pass, one backward pass, one SGD step]

\textbf{Setup.} Fit $y = \sin(2x)$ on $x \in [-\pi, \pi]$ with a
one-hidden-layer MLP of width $d_1 = 2$, GeLU activation, scalar
output, and squared-error loss
$\mathcal{L} = \tfrac{1}{2}(\hat y - y)^2$ on a single point.
The width is chosen for tractability: every number in this
example can be checked by hand. In practice you would use width
$8$ or more for this problem, but the mechanics are unchanged.

\textbf{Initialisation.} Pick a fixed seed and obtain the
following weights (rounded to two decimals):
\[
  W^{(1)} = \begin{bmatrix} 0.30 \\ -0.20 \end{bmatrix}, \;
  b^{(1)} = \begin{bmatrix} 0 \\ 0 \end{bmatrix}, \;
  W^{(2)} = \begin{bmatrix} 0.50 & 0.40 \end{bmatrix}, \;
  b^{(2)} = 0.
\]
Pick the training point $x = 0.5$, target $y = \sin(1) \approx 0.841$.

\textbf{Forward pass.} Apply \eqref{eq:nn-mlp} once.
\[
  z^{(1)} = W^{(1)} x + b^{(1)}
  = \begin{bmatrix} 0.30 \cdot 0.5 \\ -0.20 \cdot 0.5 \end{bmatrix}
  = \begin{bmatrix} 0.15 \\ -0.10 \end{bmatrix}.
\]
GeLU on each component:
$\mathrm{GeLU}(0.15) = 0.15 \cdot \Phi(0.15) \approx 0.15 \cdot 0.560 \approx 0.084$,
$\mathrm{GeLU}(-0.10) = -0.10 \cdot \Phi(-0.10) \approx -0.10 \cdot 0.460 \approx -0.046$.
So
\[
  h^{(1)}
  = \begin{bmatrix} 0.084 \\ -0.046 \end{bmatrix}.
\]
Then
\[
  \hat y \;=\; z^{(2)}
  \;=\; W^{(2)} h^{(1)} + b^{(2)}
  \;=\; 0.50 \cdot 0.084 + 0.40 \cdot (-0.046) + 0
  \;\approx\; 0.024.
\]
Loss: $\mathcal{L} = \tfrac{1}{2}(0.024 - 0.841)^2 \approx 0.334$.

\textbf{Backward pass.} Following \eqref{eq:nn-delta-recursion}.
The output-layer delta is the loss derivative w.r.t.\ $z^{(2)}$,
which for squared error is $\hat y - y$:
$\delta^{(2)} = 0.024 - 0.841 = -0.817$.
Output-layer parameter gradients (\eqref{eq:nn-param-grads}):
\[
  \frac{\partial \mathcal{L}}{\partial W^{(2)}}
  \;=\; \delta^{(2)}\, h^{(1)\top}
  \;=\; -0.817 \cdot \begin{bmatrix} 0.084 & -0.046 \end{bmatrix}
  \;\approx\; \begin{bmatrix} -0.069 & 0.038 \end{bmatrix},
  \qquad
  \frac{\partial \mathcal{L}}{\partial b^{(2)}}
  \;=\; -0.817.
\]
Propagate $\delta$ to the hidden layer.
$W^{(2)\top} \delta^{(2)} = [0.50; 0.40] \cdot (-0.817) = [-0.409; -0.327]$.
The GeLU derivative at each pre-activation is approximately
$\Phi(z) + z \phi(z)$ with $\phi$ the standard normal pdf:
$\mathrm{GeLU}'(0.15) \approx 0.560 + 0.15 \cdot 0.394 \approx 0.619$,
$\mathrm{GeLU}'(-0.10) \approx 0.460 + (-0.10) \cdot 0.397 \approx 0.421$.
Then
\[
  \delta^{(1)}
  \;=\; \bigl(W^{(2)\top} \delta^{(2)}\bigr) \odot \varphi'(z^{(1)})
  \;=\; \begin{bmatrix} -0.409 \cdot 0.619 \\ -0.327 \cdot 0.421 \end{bmatrix}
  \;\approx\; \begin{bmatrix} -0.253 \\ -0.138 \end{bmatrix}.
\]
Hidden-layer parameter gradients:
\[
  \frac{\partial \mathcal{L}}{\partial W^{(1)}}
  \;=\; \delta^{(1)} x^\top
  \;=\; \begin{bmatrix} -0.253 \\ -0.138 \end{bmatrix} \cdot 0.5
  \;\approx\; \begin{bmatrix} -0.127 \\ -0.069 \end{bmatrix},
  \qquad
  \frac{\partial \mathcal{L}}{\partial b^{(1)}}
  \;=\; \delta^{(1)}
  \;\approx\; \begin{bmatrix} -0.253 \\ -0.138 \end{bmatrix}.
\]

\textbf{One SGD step at $\eta = 0.05$.} Apply
$\theta \leftarrow \theta - \eta\, \nabla_\theta \mathcal{L}$.
\[
  W^{(2)} \leftarrow [0.50, 0.40] - 0.05 \cdot [-0.069, 0.038]
  \;\approx\; [0.503,\, 0.398],
  \qquad
  b^{(2)} \leftarrow 0 - 0.05 \cdot (-0.817)
  \;\approx\; 0.041.
\]
\[
  W^{(1)} \leftarrow \begin{bmatrix} 0.30 \\ -0.20 \end{bmatrix}
  - 0.05 \cdot \begin{bmatrix} -0.127 \\ -0.069 \end{bmatrix}
  \;\approx\; \begin{bmatrix} 0.306 \\ -0.197 \end{bmatrix},
  \qquad
  b^{(1)} \leftarrow [0, 0] - 0.05 \cdot [-0.253, -0.138]
  \;\approx\; [0.013,\, 0.007].
\]
After this single step, the prediction at $x = 0.5$ moves
slightly toward $0.841$. Iterate to convergence and the network
will fit the curve.

\textbf{The reader takeaway.} Every number in this example came
from \eqref{eq:nn-mlp}, \eqref{eq:nn-delta-recursion}, and
\eqref{eq:nn-param-grads}. There is no hidden machinery, no
trick the explanation glosses over. Backprop is just the chain
rule; SGD is just a step against the gradient; the rest is
careful bookkeeping. A network with a million parameters
and a hundred layers does the same thing, only at scale.
\end{workedexample}

\subsection{Worked Example B: XOR via 2-Layer MLP (Necessity)}

\begin{workedexample}[B: Why we need the hidden layer]

\textbf{Setup.} Inputs are the four corners of the unit square
$\{(0,0), (0,1), (1,0), (1,1)\}$ with the XOR labels
$\{0, 1, 1, 0\}$. The pedagogical question is whether a network
without a hidden layer can fit this data set.

\textbf{The single-layer attempt fails geometrically.} A
no-hidden-layer network with sigmoid output is
$\hat y = \sigma(w_1 x_1 + w_2 x_2 + b)$. Predicting the correct
label at every input requires $w_1 x_1 + w_2 x_2 + b$ to be
strongly positive for $(0,1)$ and $(1,0)$ and strongly negative
for $(0,0)$ and $(1,1)$. But the locus
$w_1 x_1 + w_2 x_2 + b = 0$ is a single straight line in the
plane, and any straight line that places $(0,1)$ and $(1,0)$ on
the positive side must also place at least one of $(0,0)$ or
$(1,1)$ on that side --- the four points sit at the corners of a
unit square, with the two ``$1$''-labelled points on one diagonal
and the two ``$0$''-labelled points on the other, and no straight
line separates two diagonals of a square into opposite halves.
\citet{minsky1969perceptrons} stated and proved this in 1969;
Worked Example~B has now stated and verified it geometrically.

\textbf{The two-layer solution.} Add a hidden layer of width $2$
with ReLU activations. The following weights solve XOR exactly:
\[
  W^{(1)} = \begin{bmatrix} 1 & 1 \\ 1 & 1 \end{bmatrix}, \;
  b^{(1)} = \begin{bmatrix} 0 \\ -1 \end{bmatrix}, \;
  W^{(2)} = \begin{bmatrix} 1 & -2 \end{bmatrix}, \;
  b^{(2)} = 0.
\]
Verify on all four inputs.

\smallskip
\begin{center}
\small
\begin{tabular}{|c|c|c|c|c|c|c|}
\hline
$(x_1, x_2)$ & $z^{(1)}$ & $h^{(1)} = \mathrm{ReLU}(z^{(1)})$ & $z^{(2)}$ & $\hat y$ & $y$ & error \\
\hline
$(0, 0)$ & $(0,\, -1)$ & $(0,\, 0)$ & $0$ & $0$ & $0$ & $0$ \\
$(0, 1)$ & $(1,\, 0)$ & $(1,\, 0)$ & $1$ & $1$ & $1$ & $0$ \\
$(1, 0)$ & $(1,\, 0)$ & $(1,\, 0)$ & $1$ & $1$ & $1$ & $0$ \\
$(1, 1)$ & $(2,\, 1)$ & $(2,\, 1)$ & $0$ & $0$ & $0$ & $0$ \\
\hline
\end{tabular}
\end{center}
\smallskip

The network outputs the correct label at every point. The two
hidden units perform two different linear projections of the
input, and the ReLU clips one of them at the origin; the output
layer subtracts twice the second from the first, producing the
XOR pattern. Geometrically, the hidden layer carves the input
plane into a piecewise-linear region whose boundary bends. A
straight line cannot bend, but the composition of two affine
layers with a nonlinearity between them can.

\textbf{Backprop on the converged solution.} Because the network
already produces zero error at every input, the squared-error
loss is exactly zero, $\delta^{(2)} = \hat y - y = 0$ at every
point, and every parameter gradient is zero. This is what
optimality looks like to the optimiser: nothing to update.

\textbf{Backprop at a perturbation.} To show the
$\delta$-recursion produce non-trivial numbers, perturb
$W^{(2)} = [1, -2]$ to $W^{(2)} = [0.9, -1.8]$ and walk one
backward pass on the input $(0, 1)$, target $y = 1$.

The forward pass at this perturbed configuration:
$z^{(1)} = (1, 0)$, $h^{(1)} = (1, 0)$,
$\hat y = z^{(2)} = 0.9 \cdot 1 + (-1.8) \cdot 0 = 0.9$.
The error is $\hat y - y = -0.1$, so $\delta^{(2)} = -0.1$.

Output-layer parameter gradients:
$\partial \mathcal{L} / \partial W^{(2)}
= -0.1 \cdot [1, 0] = [-0.1, 0]$,
$\partial \mathcal{L} / \partial b^{(2)} = -0.1$. The gradient on
$W^{(2)}_2$ is zero because the second hidden unit was inactive
at this input ($h^{(1)}_2 = 0$); there is no signal pulling
$W^{(2)}_2$ in any direction from this example.

Propagate $\delta$ to the hidden layer:
$W^{(2)\top} \delta^{(2)} = [0.9; -1.8] \cdot (-0.1) = [-0.09; 0.18]$.
The ReLU derivative at $z^{(1)} = (1, 0)$ is $(1, 0)$ (positive
input gives derivative $1$; zero input gives derivative $0$ by
the A3.1 convention). So
$\delta^{(1)} = [-0.09; 0.18] \odot [1; 0] = [-0.09; 0]$.

Hidden-layer parameter gradients:
$\partial \mathcal{L} / \partial W^{(1)}
= [-0.09; 0] \cdot [0, 1] = \begin{bmatrix} 0 & -0.09 \\ 0 & 0 \end{bmatrix}$,
$\partial \mathcal{L} / \partial b^{(1)} = [-0.09; 0]$.
An SGD step would update $W^{(1)}$, $b^{(1)}$, $W^{(2)}_1$, and
$b^{(2)}$ but leave $W^{(1)}_{2,*}$ and $W^{(2)}_2$ unchanged.

\textbf{The reader takeaway.} The same machinery from Worked
Example~A produces the same kind of numbers on a classification
problem with a hidden layer. The dead-neuron issue is concrete:
a hidden unit that is inactive at a given input contributes zero
to the gradient from that input. Over a full training set, units
that are inactive on every input become permanently dead. Init
choice (\S\ref{sec:nn-init}) and learning-rate choice are the
two practical knobs that prevent this from happening.
\end{workedexample}

% --- END Worked Examples ---

\begin{provedassumed}[Neural-network primitives: what is established vs assumed]
\textbf{Proved here.}
The composition form of an MLP and the explicit gradient
recursion via backpropagation
(\S\ref{sec:nn-backprop},
Eqs.~\ref{eq:nn-delta-recursion}--\ref{eq:nn-param-grads}) are
direct applications of the calculus chain rule to the affine /
elementwise composition of \eqref{eq:nn-mlp}. The
forward / backward FLOP equivalence follows from counting matrix
multiplies and is exact up to a small constant. The
variance-preservation derivation that motivates Xavier and He
init (\S\ref{sec:nn-init}, Eq.~\ref{eq:nn-var-preserve}) holds
under independence and zero-mean assumptions stated explicitly
in the derivation. The identity-path interpretation of residual
connections at initialisation, and the gradient-non-vanishing
argument through them
(\S\ref{sec:nn-residual}, Eq.~\ref{eq:nn-residual}), follow
from differentiating the residual recursion.

\textbf{Empirical / observed.}
The choice among ReLU, GeLU, and SwiGLU as the activation in
modern Transformers is empirical, not derived; the chapter
documents the historical progression but the current preference
is a benchmark consensus. The decision to use LayerNorm or
RMSNorm in language modelling rather than BatchNorm is
empirical, motivated by sequence-packing and small-batch
considerations rather than a theorem. The pre-norm
configuration is a recipe-conditional empirical observation
(\S\ref{sec:nn-norm}, with the full argument deferred to
Chapter~4). Universal approximation
(\S\ref{sec:nn-function-class}) is a true theorem but says
nothing about trainability and is not load-bearing for any
engineering decision in the rest of the book.

\textbf{Assumed.}
A3.1 (differentiability almost everywhere, with the convention
$\varphi'(0) := 0$ at ReLU's kink); A3.2 (finite floating-point
arithmetic; rounding error flagged where it matters); A3.3 (the
loss is the cross-entropy from Chapter~1 in the LLM context;
worked examples occasionally use squared error for regression
illustrations).
\end{provedassumed}

\section{Bridges to Later Chapters}

The chapter's seven primitives feed almost every chapter that
follows. The table below indexes the cross-references.

\begin{center}
\small
\begin{tabular}{|p{4.2cm}|p{8.2cm}|}
\hline
\textbf{Primitive established here} & \textbf{Used in} \\
\hline
Linear layer / MLP composition (\S\ref{sec:nn-function-class}) &
  Chapter~4 (FFN, $W_Q/W_K/W_V/W_O$ projections), Chapter~7 (LoRA
  low-rank deltas), Chapter~8 (reward-model and policy networks). \\
Backpropagation (\S\ref{sec:nn-backprop}) &
  Chapter~2's $6ND$ FLOP rule (justified here), Chapter~4's
  per-block FLOP budgets, every training run in the rest of the
  book. \\
Activations: GeLU, SwiGLU, $\sigma$
(\S\ref{sec:nn-activations}) &
  Chapter~4 (FFN activation), Chapter~8 (Bradley--Terry sigmoid
  link function), Chapter~5 (softmax decoding). \\
Variance-preserving init (\S\ref{sec:nn-init}) &
  Chapter~4's $1/\sqrt{d_k}$ attention scale (same argument),
  Chapter~7 (LoRA He-init initialisation of the $A$ factor). \\
LayerNorm / RMSNorm (\S\ref{sec:nn-norm}) &
  Chapter~4 (full LayerNorm/RMSNorm derivation and pre-norm
  placement), Chapter~6 (MoE block normalisation), Chapter~8
  (alignment-stage normalisation choices). \\
Residual connection (\S\ref{sec:nn-residual}) &
  Chapter~4's residual stream, Chapter~6's MoE block (residual
  around the routed FFN), Chapter~7 (depth-scaling intuitions
  for adapter placement). \\
The MLP block (\S\ref{sec:nn-mlp-block}) &
  Chapter~4 (FFN composition), Chapter~6 (MoE replaces the dense
  FFN with a sparse mixture), Chapter~7 (LoRA adapts the FFN's
  matrices). \\
\hline
\end{tabular}
\end{center}

\section{Key References}

\begin{itemize}
  \item \citet{rosenblatt1958perceptron} --- the perceptron and
        its learning rule.
  \item \citet{minsky1969perceptrons} --- the XOR critique that
        motivated multi-layer networks.
  \item \citet{werbos1974beyond}, \citet{rumelhart1986learning}
        --- backpropagation.
  \item \citet{cybenko1989approximation},
        \citet{hornik1991approximation} --- universal
        approximation.
  \item \citet{nair2010rectified} --- ReLU.
  \item \citet{hendrycks2016gelu} --- GeLU.
  \item \citet{shazeer2020glu} --- GLU and SwiGLU.
  \item \citet{glorot2010understanding} --- Xavier / Glorot init.
  \item \citet{he2015delving} --- He / Kaiming init.
  \item \citet{ioffe2015batchnorm} --- BatchNorm (overview only).
  \item \citet{ba2016layernorm} --- LayerNorm.
  \item \citet{zhang2019rmsnorm} --- RMSNorm.
  \item \citet{he2016resnet} --- residual connections.
  \item \citet{chen2018neuralode} --- neural ODE limit (one-line
        nod only).
\end{itemize}

\section{Recommended University Lectures}

\begin{itemize}
  \item Stanford CS231n --- the canonical undergraduate-friendly
        course on neural-network fundamentals; lectures on
        backpropagation, activations, and initialisation are
        directly aligned with this chapter.
  \item NYU Deep Learning (Yann LeCun) --- comprehensive
        treatment of the function-class, optimisation, and
        depth-trainability arguments; covers ResNet and the
        residual response in detail.
  \item MIT 6.S191 --- shorter course; useful for revisiting the
        backprop derivation.
\end{itemize}

\section{Practitioner Lab}

\begin{enumerate}
  \item \textbf{XOR from scratch.} Implement the 2-layer ReLU
        XOR solution from Worked Example~B in 20 lines of
        NumPy. Verify the forward pass on all four inputs.
        Initialise the network from random weights, train on
        the four points with squared-error loss and SGD, and
        watch the loss curve drop to near zero. Print the
        learned weights and compare to the canonical solution
        in Worked Example~B; SGD will typically find a
        permutation or rescaling of the canonical weights, not
        the canonical weights themselves.
  \item \textbf{Vanishing-gradient demonstration.} Train a
        20-layer MLP with $\tanh$ activations on a tiny
        regression task. Plot the gradient norm at each layer
        across the first 100 steps. Observe that the norm at
        the input-layer side decays exponentially with depth.
        Replace $\tanh$ with ReLU; observe the gradient norms
        recover. Add residual connections; observe that they
        recover further.
  \item \textbf{Init ablation.} Train a 10-layer MLP with ReLU
        on MNIST under three initialisations: zero variance
        (variance collapse), variance $1.0$ (variance
        explosion), and He init. Plot the loss curves. Verify
        that only the He-init network trains.
  \item \textbf{Forward / backward FLOP audit.} For a
        $\{d, d_f, L\} = \{4096, 16384, 32\}$ Transformer block
        configuration, compute the FLOPs of one forward pass
        and one backward pass per token. Verify that the
        backward pass is approximately twice the forward pass,
        consistent with the $6ND$ rule.
  \item \textbf{Normaliser comparison.} Train a small Transformer
        with LayerNorm and with RMSNorm; verify the empirical
        observation that they produce equivalent loss curves
        within run-to-run noise, with RMSNorm faster per step.
\end{enumerate}

\section*{Derivation Ledger (Ch. 3)}

\begin{derivledger}
Compact index of the chapter's central identities.
\begin{itemize}
  \item \textbf{MLP composition.}\
        $f_\theta(x) = h^{(L)}$ with
        $h^{(\ell)} = \varphi(W^{(\ell)} h^{(\ell-1)} + b^{(\ell)})$
        --- Eq.~\eqref{eq:nn-mlp}.
  \item \textbf{Backprop $\delta$-recursion.}\
        $\delta^{(\ell-1)} = (W^{(\ell)\top} \delta^{(\ell)}) \odot \varphi'(z^{(\ell-1)})$
        --- Eq.~\eqref{eq:nn-delta-recursion}.
  \item \textbf{Parameter gradients.}\
        $\partial \mathcal{L} / \partial W^{(\ell)} = \delta^{(\ell)} h^{(\ell-1)\top}$,
        $\partial \mathcal{L} / \partial b^{(\ell)} = \delta^{(\ell)}$
        --- Eq.~\eqref{eq:nn-param-grads}.
  \item \textbf{Forward / backward FLOP equivalence.}\
        Backward pass costs $\approx 2\times$ forward pass to
        within constants; total $\approx 6P$ FLOPs per training
        token (justifies the $6ND$ rule of Chapter~2).
  \item \textbf{ReLU.}\
        $\varphi(z) = \max(0, z)$, $\varphi'(z) = \mathbb{1}[z>0]$
        --- Eq.~\eqref{eq:nn-relu}.
  \item \textbf{GeLU.}\
        $\varphi(z) = z \cdot \Phi(z)$ --- Eq.~\eqref{eq:nn-gelu}.
  \item \textbf{GLU / SwiGLU.}\
        $\varphi(x) = (W_g x) \odot \varphi(W_u x)$
        --- Eq.~\eqref{eq:nn-glu}.
  \item \textbf{Variance preservation.}\
        $\Var(W) = 1/d_{\rm in}$
        --- Eq.~\eqref{eq:nn-var-preserve}.
  \item \textbf{Xavier init.}\
        $\Var(W) = 2/(d_{\rm in} + d_{\rm out})$
        --- Eq.~\eqref{eq:nn-xavier}.
  \item \textbf{He init.}\
        $\Var(W) = 2/d_{\rm in}$
        --- Eq.~\eqref{eq:nn-he-init}.
  \item \textbf{LayerNorm.}\
        $\mathrm{LN}(x) = \gamma \odot (x - \mu(x))/(\sigma(x) + \epsilon) + \beta$
        --- Eq.~\eqref{eq:nn-layernorm}.
  \item \textbf{RMSNorm.}\
        $\mathrm{RMSNorm}(x) = \gamma \odot x / (\mathrm{RMS}(x) + \epsilon)$
        --- Eq.~\eqref{eq:nn-rmsnorm}.
  \item \textbf{Residual connection.}\
        $h^{(\ell)} = h^{(\ell-1)} + F^{(\ell)}(h^{(\ell-1)})$
        --- Eq.~\eqref{eq:nn-residual}.
  \item \textbf{Transformer FFN.}\
        $\mathrm{FFN}(x) = W_2 \varphi(W_1 x + b_1) + b_2$
        --- Eq.~\eqref{eq:nn-ffn}.
\end{itemize}
\end{derivledger}

\section{Chapter 3 Takeaway}

Modern neural networks are not exotic. They are compositions of
linear layers, nonlinear activations, normalisations, and
residual connections, trained by gradient descent through the
chain rule. Each primitive in this chapter exists for a reason
that is empirically grounded and historically traceable. The
perceptron taught us that adjusting weights in proportion to
errors was a viable mechanism for learning. The XOR critique
taught us that depth was necessary. Backpropagation made depth
tractable. ReLU and GeLU removed the saturation problem that had
kept networks shallow. Xavier and He init kept the activation
variance from collapsing or exploding at depth. LayerNorm and
RMSNorm kept it stable through training. Residual connections
made networks of a hundred layers train as well as networks of
ten. None of these moves was inevitable; each was a response to
a specific failure of the design that came before.

The Transformer of Chapter~4 is what these primitives compose
into when you also add the attention mechanism. By the end of
this chapter you should be able to read Chapter~4 as a
particular arrangement of pieces you already understand, rather
than as a list of unfamiliar names. That is the point of putting
this chapter ahead of the architecture: to make the architecture
legible.

% --- END Chapter 3 ---
