\section{Week 6: Efficiency, Scaling, and Mixture-of-Experts}

\subsection{Learning Objectives}

By the end of this week, you should be able to:
\begin{itemize}
  \item Explain why efficiency dominates modern LLM design
  \item Understand the tradeoff between model capacity and compute cost
  \item Describe Mixture-of-Experts (MoE) architectures and routing
  \item Identify new failure modes introduced by conditional computation
  \item Reason about MoE from a systems-engineering perspective
\end{itemize}

\subsection{LLM Components Covered}

\textbf{Scaling and efficiency}
\begin{itemize}
  \item Dense vs sparse computation
  \item Conditional parameter activation
  \item Compute–memory tradeoffs
\end{itemize}

\textbf{Architectural patterns}
\begin{itemize}
  \item Mixture-of-Experts layers
  \item Routing networks
  \item Load balancing mechanisms
\end{itemize}

\subsection{Conceptual Core: Why Efficiency Matters}

Modern LLMs are constrained not by parameter count alone, but by:
\begin{itemize}
  \item training compute,
  \item inference latency,
  \item memory bandwidth,
  \item energy consumption.
\end{itemize}

A dense Transformer activates \emph{all} parameters for \emph{every} token.
This scales cost linearly with model size.

Mixture-of-Experts (MoE) architectures decouple:
\begin{quote}
\textbf{model capacity} from \textbf{per-token compute}.
\end{quote}

\subsection{Dense vs Sparse Computation}

In a standard Transformer, the feed-forward block computes:

\[
\mathrm{FFN}(x) = W_2 \sigma(W_1 x)
\]

In an MoE layer, this becomes:

\[
\mathrm{MoE}(x) = \sum_{i \in \mathcal{E}(x)} g_i(x) \cdot \mathrm{Expert}_i(x)
\]

where:
\begin{itemize}
  \item $\mathcal{E}(x)$ is a small subset of experts,
  \item $g_i(x)$ are gating weights,
  \item only selected experts are evaluated.
\end{itemize}

\subsection{Routing and Gating}

Routing is typically performed by a lightweight network that maps inputs to expert scores.

Common strategies:
\begin{itemize}
  \item top-1 routing,
  \item top-2 routing,
  \item load-balanced routing.
\end{itemize}

Routing decisions introduce discreteness into otherwise smooth optimization.

\subsection{Benefits of MoE}

MoE architectures allow:
\begin{itemize}
  \item massive parameter counts,
  \item lower inference cost,
  \item specialization of sub-networks.
\end{itemize}

This makes them attractive for frontier-scale models.

\subsection{New Failure Modes}

MoE introduces new challenges:
\begin{itemize}
  \item expert collapse,
  \item load imbalance,
  \item routing instability,
  \item harder debugging and reproducibility.
\end{itemize}

These are \emph{systems problems}, not merely modeling problems.

\subsection{Systems Engineering Implications}

\textbf{Requirements}
\begin{itemize}
  \item Predictable latency
  \item Robust routing under distribution shift
\end{itemize}

\textbf{Architecture}
\begin{itemize}
  \item Expert parallelism
  \item Routing instrumentation
\end{itemize}

\textbf{Verification \& Validation}
\begin{itemize}
  \item Load balance metrics
  \item Per-expert activation monitoring
\end{itemize}

\textbf{Operations}
\begin{itemize}
  \item Capacity planning
  \item Expert health checks
\end{itemize}

\subsection{Human Factors and Failure Modes}

MoE systems are harder to reason about:
\begin{itemize}
  \item behavior may vary subtly across inputs,
  \item failures may be localized to experts,
  \item debugging requires new mental models.
\end{itemize}

This increases operational cognitive load.

\subsection{Key References}

\begin{itemize}
  \item \citet{shazeer2017outrageously} — original MoE
  \item \citet{fedus2021switch} — Switch Transformers
\end{itemize}

\subsection{Recommended University Lectures}

\begin{itemize}
  \item Stanford CS324 — scaling and efficiency  
  \url{https://stanford-cs324.github.io/winter2022/}

  \item Berkeley CS182 — systems aspects of deep learning  
  \url{https://cs182.berkeley.edu}
\end{itemize}

\subsection{Practitioner Lab}

See \texttt{labs/week06/week06\_lab\_efficiency\_moe.ipynb}.

\subsection{Week 6 Takeaway}

\begin{quote}
Scaling LLMs is no longer a modeling problem alone — it is a distributed systems problem.
\end{quote}


