\section{Week 2: Optimization, Generalization, and Scaling Intuition}
\label{sec:week2-optgen}

\subsection{Learning Objectives}

By the end of this week, you should be able to:
\begin{itemize}
  \item Derive SGD, momentum, and Adam/AdamW updates from first principles and
        explain the role of bias correction.
  \item Reason quantitatively about learning-rate schedules (linear warmup,
        cosine decay, inverse-square-root) and their engineering tradeoffs.
  \item Describe why gradient clipping is necessary for deep Transformer
        stacks and choose a clipping threshold defensibly.
  \item State the Kaplan and Chinchilla scaling laws and derive the
        compute-optimal model/data ratio.
  \item Explain double descent and benign overfitting, and articulate why
        overparameterization can \emph{improve} generalization in practice.
  \item Evaluate mixed-precision training, loss scaling, gradient
        accumulation, and checkpointing as systems-level mechanisms.
  \item Design a reproducibility protocol (seeds, deterministic kernels,
        data ordering, regression gates) for production training pipelines.
\end{itemize}

\subsection{LLM Components Covered}

\textbf{Training dynamics}
\begin{itemize}
  \item Stochastic gradient estimators and their noise structure.
  \item First-order adaptive optimizers (Adam, AdamW) and their
        moment-estimate machinery.
  \item Learning-rate schedules, warmup, and decay.
  \item Gradient clipping, gradient accumulation, and loss scaling.
\end{itemize}

\textbf{Generalization machinery}
\begin{itemize}
  \item Overparameterization and implicit regularization.
  \item Double descent and benign overfitting.
  \item Flat-vs-sharp minima in the large-batch regime.
\end{itemize}

\textbf{Systems relevance}
\begin{itemize}
  \item Training instability, divergence, and regressions.
  \item Reproducibility under distributed training.
  \item Scaling-law-driven capacity planning.
  \item Cost / quality tradeoff under fixed compute budgets.
\end{itemize}

\subsection{Conceptual Core: Optimization Is Not Inference}

Training an LLM is, mechanically, a large-scale numerical optimization
problem. The learner does not \emph{reason} about the target; it performs a
long sequence of gradient updates against an empirical risk. The objective
from Week~1 is the empirical cross-entropy
\begin{equation}
\label{eq:week2-empirical-loss}
\mathcal{L}(\theta)
  \;=\; \frac{1}{N} \sum_{i=1}^{N}
        -\log p_{\theta}\!\left( x^{(i)}_{t_i} \,\middle|\, x^{(i)}_{<t_i} \right),
\end{equation}
optimized via variants of stochastic gradient descent:
\begin{equation}
\label{eq:week2-sgd}
\theta_{t+1} \;=\; \theta_t - \eta_t\, g_t,
\qquad
g_t \;=\; \frac{1}{B}\sum_{i \in \mathcal{B}_t}
          \nabla_\theta \bigl[-\log p_{\theta}(x^{(i)} \mid x^{(i)}_{<})\bigr],
\end{equation}
where $\eta_t$ is the learning rate and $\mathcal{B}_t$ is a minibatch of
size $B$. The key features that distinguish LLM optimization from the
textbook setting are:
\begin{itemize}
  \item \textbf{Dimensionality}: $\theta \in \mathbb{R}^d$ with
        $d \sim 10^{9}$--$10^{12}$ for frontier models.
  \item \textbf{Non-convexity}: the loss surface has a combinatorial
        number of saddle points, plateaus, and basins.
  \item \textbf{Stochasticity}: $g_t$ is an unbiased but noisy estimator
        of $\nabla \mathcal{L}$, with variance proportional to $1/B$.
  \item \textbf{Distributed execution}: gradients are aggregated across
        thousands of accelerators, making single-step cost high and
        divergence expensive.
\end{itemize}
That modern deep networks reliably converge under these conditions is an
empirical observation about the loss landscape, not a theorem.

\subsection{Stochastic Gradient Descent and Momentum}

\subsubsection{SGD as a Noisy Dynamical System}

Write the minibatch gradient as the true gradient plus noise,
$g_t = \nabla \mathcal{L}(\theta_t) + \xi_t$, with
$\mathbb{E}[\xi_t]=0$ and $\mathrm{Cov}(\xi_t) \approx \Sigma(\theta_t)/B$.
For small $\eta$, the continuous-time limit of
Eq.~\eqref{eq:week2-sgd} is the Langevin SDE
\begin{equation}
d\theta_t \;=\; -\nabla \mathcal{L}(\theta_t)\, dt
                + \sqrt{\eta\,\Sigma(\theta_t)/B}\; dW_t,
\end{equation}
which reveals two systems-relevant facts:
(i)~the stationary distribution is \emph{not} concentrated on the loss
minimum; it is biased toward flatter regions, which correlates with better
generalization~\citep{keskar2017large};
(ii)~the effective ``noise temperature'' scales with $\eta/B$, so doubling
batch size roughly halves gradient noise — this is the basis for the
linear scaling rule in large-batch training~\citep{goyal2017accurate}.

\subsubsection{Heavy-Ball and Nesterov Momentum}

Momentum introduces an exponentially-weighted moving average of past
gradients:
\begin{align}
m_t &= \beta\, m_{t-1} + (1-\beta)\, g_t, \label{eq:week2-momentum-m} \\
\theta_{t+1} &= \theta_t - \eta_t\, m_t.
\end{align}
Geometrically, momentum acts as a low-pass filter on the gradient signal:
high-frequency noise is attenuated while persistent descent directions
accumulate. Typical values are $\beta \in [0.9, 0.99]$. Nesterov's
look-ahead variant evaluates the gradient at the extrapolated point
$\theta_t - \eta \beta m_{t-1}$, which admits a stronger convergence
rate on convex problems but offers modest gains on LLM objectives.

\subsection{Adam and AdamW}

\subsubsection{Why Adaptive Per-Parameter Rates}

In a Transformer, embedding vectors for frequent tokens see large,
frequent gradients while embeddings for rare tokens see small, infrequent
ones. A single global $\eta$ cannot serve both regimes well: it is
either too aggressive for frequent tokens (overshooting) or too
conservative for rare ones (starvation). Adaptive methods scale the
update for each parameter by an online estimate of its gradient
magnitude~\citep{kingma2015adam}.

\subsubsection{Adam Update Rule}

Let $g_t = \nabla_\theta \mathcal{L}(\theta_t)$ be the minibatch
gradient. Adam maintains two exponential moving averages:
\begin{align}
m_t &= \beta_1\, m_{t-1} + (1-\beta_1)\, g_t
       & \text{(first moment)}, \label{eq:week2-adam-m}\\
v_t &= \beta_2\, v_{t-1} + (1-\beta_2)\, g_t^2
       & \text{(second moment)},
\end{align}
where $g_t^2$ is elementwise. Because $m_0 = v_0 = 0$, these estimates
are biased toward zero in the first several steps. Adam corrects this
bias analytically. Unrolling Eq.~\eqref{eq:week2-adam-m},
\[
m_t = (1-\beta_1) \sum_{k=1}^{t} \beta_1^{\,t-k} g_k,
\qquad
\mathbb{E}[m_t] = (1-\beta_1^{\,t})\,\mathbb{E}[g_t] + O(\text{drift}),
\]
so the debiased estimator is $\hat m_t = m_t / (1-\beta_1^{\,t})$; the
analogous correction applies to $v_t$. The parameter update becomes
\begin{equation}
\label{eq:week2-adam-update}
\theta_{t+1} \;=\; \theta_t - \eta_t\,\frac{\hat m_t}{\sqrt{\hat v_t} + \epsilon},
\qquad
\hat m_t = \frac{m_t}{1-\beta_1^{\,t}},
\quad
\hat v_t = \frac{v_t}{1-\beta_2^{\,t}},
\end{equation}
with default hyperparameters $\beta_1 = 0.9$, $\beta_2 = 0.999$,
$\epsilon = 10^{-8}$. For large Transformer training, practitioners
commonly reduce $\beta_2$ to $0.95$ to make the second-moment estimator
more responsive to nonstationarity.

\subsubsection{AdamW: Decoupled Weight Decay}

Classical L2 regularization adds $\tfrac{\lambda}{2}\|\theta\|^2$ to the
loss, which introduces a $\lambda \theta$ term into $g_t$. In Adam this
term is then divided by $\sqrt{\hat v_t}$, so parameters with small
gradients see disproportionately strong decay — the effective
regularization is entangled with the gradient magnitude. AdamW
\emph{decouples} weight decay by applying it directly to the parameters,
not through the loss gradient~\citep{loshchilov2019decoupled}:
\begin{equation}
\label{eq:week2-adamw-update}
\theta_{t+1} \;=\; \theta_t
  - \eta_t\!\left(\frac{\hat m_t}{\sqrt{\hat v_t} + \epsilon} + \lambda\,\theta_t\right).
\end{equation}
This subtle change yields consistently better generalization on
Transformer workloads and is the default optimizer for essentially all
production LLM training.

\begin{center}
\begin{tabular}{|l|l|l|}
\hline
Optimizer & Strengths & Risks / Caveats \\
\hline
SGD        & Simple, flatness-biased & Slow without tuning \\
SGD+Mom.   & Accelerates narrow valleys & Still sensitive to $\eta$ \\
Adam       & Fast early progress & Sharp minima, weight-decay bug \\
AdamW      & Decoupled decay, stable & Hyperparameter-sensitive, memory-heavy \\
Lion/Shampoo & Less memory, competitive & Less mature tooling \\
\hline
\end{tabular}
\end{center}

\subsubsection{Memory Cost of Optimizer State}

AdamW stores $m_t$ and $v_t$ at full precision for every parameter, so
optimizer state is roughly $2\times$ the size of the parameters (or
$\sim 8$ bytes/param at fp32). Combined with parameters themselves and
activation memory, this is what ZeRO-style sharding
\citep{rajbhandari2020zero} attacks in distributed training (revisited
in Week~6).

\subsection{Learning Rate Schedules}

The learning rate $\eta_t$ is the single most important hyperparameter
for Transformer training. Three families dominate LLM practice.

\subsubsection{Linear Warmup}

Start small and ramp up linearly over $T_w$ steps:
\begin{equation}
\eta_t \;=\; \eta_{\max} \cdot \min\!\left(1, \frac{t}{T_w}\right).
\end{equation}
Typical settings use $T_w \in [500, 5000]$ steps for models of
$10^{9}$--$10^{11}$ parameters. Warmup is essential because early
gradients are dominated by unstructured initialization: a full-size step
can drive parameters into a bad basin from which Adam's second-moment
estimator cannot recover quickly.

\subsubsection{Cosine Decay}

After warmup, the rate decays smoothly toward a minimum $\eta_{\min}$:
\begin{equation}
\eta_t \;=\; \eta_{\min} + \tfrac{1}{2}(\eta_{\max} - \eta_{\min})
            \left[1 + \cos\!\left(\pi\,\frac{t - T_w}{T_{\max} - T_w}\right)\right],
\qquad t > T_w.
\end{equation}
Cosine decay avoids the abrupt ``cliffs'' of step schedules and empirically
matches or exceeds more elaborate alternatives on LLM
pretraining~\citep{loshchilov2017sgdr}. A common variant holds
$\eta_{\min} = 0.1\,\eta_{\max}$ so that training never truly stops making
progress.

\subsubsection{Inverse-Square-Root (Noam) Schedule}

The original Transformer~\citep{vaswani2017attention} used
\begin{equation}
\eta_t \;=\; d_{\mathrm{model}}^{-1/2} \cdot
            \min\!\left(t^{-1/2},\; t\, T_w^{-3/2}\right),
\end{equation}
which combines linear warmup with an inverse-square-root decay.
Although cosine has largely displaced it for pretraining, inverse-sqrt
remains the default in many fine-tuning recipes.

\subsubsection{Engineering Guidance}

Choose schedules by matching the workload:
pretraining runs that consume a fixed compute budget pair warmup+cosine;
continual pretraining and fine-tuning that resume from a checkpoint
usually want warmup+constant or warmup+linear; very long-horizon runs
sometimes adopt \emph{one-cycle} / super-convergence
schedules~\citep{smith2018superconvergence} where $\eta$ rises and falls
once over the entire run.

\subsection{Gradient Clipping}

Deep Transformer stacks can produce exploding-norm gradients when
activations or attention logits grow during early training. Gradient
clipping bounds the global step size:
\begin{equation}
\label{eq:week2-gradclip}
g_t \;\leftarrow\; g_t \cdot \min\!\left(1, \frac{c}{\|g_t\|_2}\right),
\qquad c \in [0.5, 2.0].
\end{equation}
With the $\min$, the update direction is preserved while the magnitude
is capped at $c$. This is strictly preferable to elementwise clipping,
which distorts direction. Empirically, $c = 1.0$ is the default for
large LLM pretraining. Clipping is a cheap, principled defense against
rare but catastrophic optimizer divergence~\citep{pascanu2013difficulty},
and its activation rate is a useful operational telemetry signal: a
sudden increase typically precedes a loss spike by a few hundred steps.

\subsection{Mixed Precision, Loss Scaling, and Accumulation}

\subsubsection{Mixed-Precision Arithmetic}

LLM training uses mixed precision to increase throughput and reduce
memory~\citep{micikevicius2018mixed}: activations and gradients are held
in 16-bit (fp16 or bf16), while a ``master copy'' of parameters and
optimizer state is held in fp32. The chain is
\[
\theta_{\mathrm{fp32}} \xrightarrow{\text{cast}} \theta_{\mathrm{fp16}}
\xrightarrow{\text{forward/backward}} g_{\mathrm{fp16}}
\xrightarrow{\text{cast}} g_{\mathrm{fp32}}
\xrightarrow{\text{optimizer}} \theta_{\mathrm{fp32}}.
\]
The optimizer step is performed in fp32 to avoid accumulation error on
$m_t$, $v_t$. Bf16 is preferred over fp16 on modern hardware because its
wider exponent eliminates the need for loss scaling.

\subsubsection{Loss Scaling}

When using fp16 (not bf16), small gradients underflow to zero. Loss
scaling multiplies the loss by $s \in \{2^{10}, 2^{24}\}$ before
backpropagation, then divides gradients by $s$ before the optimizer
step. \emph{Dynamic} loss scaling halves $s$ whenever an $\mathrm{NaN}$
or $\mathrm{Inf}$ is detected and doubles it periodically otherwise.
This is an instance of adaptive numerical conditioning at the systems
layer, invisible to the mathematical model.

\subsubsection{Gradient Accumulation}

When the desired effective batch size exceeds accelerator memory,
gradients from $K$ micro-batches are summed in place before a single
optimizer step:
\[
g_t \;=\; \frac{1}{K}\sum_{k=1}^{K} g_t^{(k)}.
\]
Accumulation is algebraically equivalent to a $K$-times-larger batch at
the cost of $K$ extra forward/backward passes per update. It is the
primary mechanism by which organizations match the effective batch
sizes in published recipes while running on smaller clusters.

\subsubsection{Activation Checkpointing}

Activation memory dominates GPU usage. Activation checkpointing
\citep{chen2016training} trades compute for memory by discarding
intermediate activations during the forward pass and recomputing them
during backprop, typically saving $\sqrt{L}$ activation memory for $L$
layers at a $\sim 30\%$ compute overhead. This is how multi-hundred-billion
parameter models fit on realistic hardware.

\subsection{Overparameterization, Double Descent, and Benign Overfitting}

Classical statistical learning
theory~\citep{vapnik1998statistical,bishop2006prml} suggests that models
with more parameters than data should overfit catastrophically. Modern
deep networks contradict this: with $d \gg N$, training loss goes to
zero and test loss \emph{continues to improve}.

\subsubsection{Double Descent}

Plotted as a function of model size, test error exhibits the classical
U-curve up to the \emph{interpolation threshold} where training error
reaches zero. Beyond that threshold, test error descends
again~\citep{belkin2019reconciling,nakkiran2020double}. Three regimes:
\begin{itemize}
  \item \textbf{Underparameterized}: bias dominates; more capacity helps.
  \item \textbf{Interpolation threshold}: variance is maximized;
        test error spikes.
  \item \textbf{Overparameterized}: implicit regularization dominates;
        more capacity continues to help.
\end{itemize}
LLMs live deep in the third regime, which is why ``just train a bigger
model'' has been such a reliable strategy.

\subsubsection{Why Bigger Is Not Worse}

Several mechanisms are believed to contribute to benign overfitting:
\begin{itemize}
  \item SGD has an implicit bias toward low-norm, flat
        minima~\citep{keskar2017large}, which correlate with
        generalization;
  \item high-dimensional parameter spaces admit many zero-training-loss
        solutions, of which the optimizer tends to pick well-behaved
        ones;
  \item training data is structured and redundant, so capacity is used
        for coverage rather than memorization.
\end{itemize}
None of these arguments is a full theory, but together they justify the
engineering decision to scale up rather than regularize down.

\subsection{Scaling Laws}

\subsubsection{Kaplan Scaling}

\citet{kaplan2020scaling} observed that cross-entropy loss follows
smooth power laws in model size $N$, dataset size $D$, and compute $C$:
\begin{equation}
\label{eq:week2-kaplan}
L(N) \;\approx\; \left(\frac{N_c}{N}\right)^{\alpha_N},
\qquad
L(D) \;\approx\; \left(\frac{D_c}{D}\right)^{\alpha_D},
\qquad
L(C) \;\approx\; \left(\frac{C_c}{C}\right)^{\alpha_C},
\end{equation}
with fitted exponents $\alpha_N \approx 0.076$,
$\alpha_D \approx 0.095$, and $\alpha_C \approx 0.05$. Equivalently,
\[
\log L \;=\; \mathrm{const} - \alpha_N \log N,
\]
so on log--log axes loss is a straight line in $N$ over six orders of
magnitude. This predictability is what makes LLM training an
engineering discipline rather than a craft.

\subsubsection{Chinchilla Scaling}

\citet{hoffmann2022chinchilla} refitted the joint law with better
hyperparameter tuning and reached a very different conclusion. Their
form is
\begin{equation}
\label{eq:week2-chinchilla}
L(N, D) \;=\; E + \frac{A}{N^{\alpha}} + \frac{B}{D^{\beta}},
\end{equation}
with approximately $\alpha \approx 0.34$, $\beta \approx 0.28$,
$E \approx 1.69$. Given a fixed compute budget $C \approx 6\,N D$ FLOPs,
minimizing $L(N,D)$ subject to $C$ yields the \emph{compute-optimal}
ratio
\begin{equation}
\label{eq:week2-chinchilla-opt}
N^\star \propto C^{a},
\qquad
D^\star \propto C^{b},
\qquad a \approx b \approx 0.5,
\end{equation}
i.e.\ model parameters and tokens should scale roughly equally with
compute. The Kaplan recommendation of ``scale model size faster than
dataset'' was an artifact of undertuned baselines. The practical
implication is dramatic: for any fixed compute budget there is a
\emph{cost-optimal} model size, and training a larger model on fewer
tokens is strictly worse than training a smaller model on more tokens.

\subsubsection{What Scaling Laws Do and Do Not Tell You}

Scaling laws predict \emph{training loss} as a function of resources.
They do \emph{not} predict:
\begin{itemize}
  \item emergent capabilities (discussed in Week~5 via
        \citet{wei2022emergent} and \citet{schaeffer2023mirage});
  \item performance on specific downstream tasks;
  \item the effect of data quality, deduplication, or mixture ratios.
\end{itemize}
They are compute-planning tools, not capability forecasts.

\subsection{Systems Engineering Implications}

\begin{center}
\begin{tabular}{|p{3.4cm}|p{3.4cm}|p{6.8cm}|}
\hline
V-model stage & Optimizer / training property & System requirement or activity \\
\hline
Requirements & Fixed compute budget & Use Chinchilla law to size $N, D$. \\
 & Reproducibility target & Seeded data order, deterministic kernels. \\
\hline
Architecture & AdamW state $\sim 2\times$ params & Budget optimizer memory (ZeRO). \\
 & fp16 vs bf16 & Choose based on hardware and stability. \\
 & Activation memory & Plan recompute vs sharding. \\
\hline
Implementation & Warmup + cosine & Expose schedule params to config. \\
 & Gradient accumulation & Log effective vs nominal batch. \\
\hline
V\&V & Loss spikes and plateaus & Divergence alarms; checkpoint rollback. \\
 & Clip activation rate & Telemetry feed into regression gates. \\
 & Small-scale proxies & Shadow runs at $10^{-3}\times$ compute
   \citep{wortsman2023stable}. \\
\hline
Operations & Regressions between runs & Rerun on fixed seed + data; bisect. \\
 & Cost / quality tracking & Dollars per perplexity point. \\
\hline
\end{tabular}
\end{center}

\subsection{Reproducibility in Distributed Training}

Deterministic training under data-parallel, tensor-parallel, and
pipeline-parallel execution is substantially harder than it sounds.
Common sources of nondeterminism:
\begin{itemize}
  \item asynchronous all-reduce summation order;
  \item nondeterministic cuDNN kernels (e.g., convolutions);
  \item dataloader worker scheduling;
  \item mixed-precision dynamic loss scaling decisions;
  \item floating-point non-associativity across restart boundaries.
\end{itemize}
A defensible reproducibility protocol fixes:
(i)~the random seed for parameter init, data shuffling, and dropout;
(ii)~the global data order (checkpoint the dataloader state);
(iii)~the kernel selection via deterministic-mode flags;
(iv)~the loss-scale trajectory (deterministic schedule rather than
dynamic halving).
Even then, \emph{bit-exact} reproducibility across hardware generations
is rarely achievable; the practical goal is \emph{statistical}
reproducibility, i.e., loss curves within noise bands across reruns.

\subsection{Human Factors and Failure Modes}

A recurring failure mode is the attribution of \emph{training behavior}
to \emph{model intelligence}. Loss plateaus, loss drops after restarts,
apparent ``insight moments'' — all of these are best explained as
optimization dynamics. Concrete examples:
\begin{itemize}
  \item Attributing a mid-training loss jump to a capability emerging,
        when the actual cause was an inadvertent change in data
        ordering after a checkpoint restore.
  \item Interpreting a flat training loss as ``the model has learned
        everything it can,'' when the actual cause is a learning-rate
        schedule that has decayed too far.
  \item Overreading the output of small-scale probes: noise at
        $10^9$ parameters does not monotonically forecast behavior at
        $10^{11}$ parameters.
\end{itemize}
The countermeasure is disciplined telemetry: track gradient norm,
clip-rate, parameter norm, activation statistics, and token-ordering
hash on every run. Most training ``mysteries'' resolve to an operator
mistake that better instrumentation would have caught.

\subsection{Bridges to Later Chapters}

\begin{itemize}
  \item \textbf{Week 3 (Tokenization and Embeddings)}: embedding tables
        are the largest single contributor to optimizer state memory;
        tokenizer choice directly affects the effective $D$ in
        Eq.~\eqref{eq:week2-chinchilla}.
  \item \textbf{Week 4 (Transformer Architecture)}: the layer
        normalization placement (pre-LN vs post-LN)
        \citep{xiong2020onlayer} interacts with warmup schedule length.
  \item \textbf{Week 5 (Pretraining \& In-Context)}: the scaling-law
        formulas set the budget; emergent capability claims must be
        checked against them for confounds~\citep{schaeffer2023mirage}.
  \item \textbf{Week 6 (Efficiency, Scaling, MoE)}: ZeRO
        \citep{rajbhandari2020zero} and pipeline parallelism
        \citep{narayanan2021efficient} are direct responses to the
        optimizer-memory constraint derived here.
  \item \textbf{Week 7 (Adaptation, LoRA, Quantization)}: adaptation
        methods replace a full-fidelity optimizer over $10^{11}$
        parameters with one over $\le 10^{7}$, tracing back to the
        cost equation laid out in \S\,Scaling Laws.
  \item \textbf{Week 10 (Governance \& Assurance)}: reproducibility
        protocols become evidence in an assurance case; training
        regression detection is an operational-safety control.
\end{itemize}

\subsection{Key References}

\begin{itemize}
  \item \citet{goodfellow2016dl} — canonical treatment of optimization
        and generalization in deep learning.
  \item \citet{kingma2015adam} — original Adam paper.
  \item \citet{loshchilov2019decoupled} — AdamW, decoupled weight
        decay.
  \item \citet{loshchilov2017sgdr} — cosine-decay schedules.
  \item \citet{goyal2017accurate} — large-batch warmup and the linear
        scaling rule.
  \item \citet{pascanu2013difficulty} — origin of gradient clipping.
  \item \citet{micikevicius2018mixed} — mixed-precision training.
  \item \citet{chen2016training} — activation checkpointing.
  \item \citet{kaplan2020scaling} — scaling laws for neural language
        models.
  \item \citet{hoffmann2022chinchilla} — compute-optimal scaling.
  \item \citet{nakkiran2020double}, \citet{belkin2019reconciling} —
        double descent and benign overfitting.
  \item \citet{keskar2017large} — large-batch training and flat
        minima.
  \item \citet{zhang2017understanding} — empirical study of
        overparameterization.
  \item \citet{wortsman2023stable} — small-scale proxies for
        large-training instabilities.
\end{itemize}

\subsection{Recommended University Lectures}

\begin{itemize}
  \item MIT 6.S191 --- optimization and deep learning practice,
        \citet{mit_6s191}.
  \item Berkeley CS182 --- deep learning foundations,
        \citet{berkeley_cs182}.
  \item Stanford CS229 --- gradient descent and generalization,
        \citet{stanford_cs229}.
  \item Stanford CS324 --- Large Language Models, module on training
        dynamics, \citet{stanford_cs324}.
  \item Caltech CS/CNS/EE~156 --- Learning from Data,
        \citet{caltech_cs156}.
\end{itemize}

\subsection{Practitioner Lab}

\begin{enumerate}
  \item \textbf{Optimizer comparison.} Train a $\sim\!10$\,M-parameter
        Transformer on WikiText-2 with SGD, SGD+momentum, Adam, and
        AdamW. Plot training and validation loss; report steps-to-target
        and final perplexity.
  \item \textbf{Bias correction.} Remove the
        $1-\beta_1^{\,t}, 1-\beta_2^{\,t}$ terms from
        Eq.~\eqref{eq:week2-adam-update}. Measure the impact on the first
        500 steps.
  \item \textbf{LR sweep.} Sweep $\eta_{\max} \in \{1, 3, 10, 30, 100\}
        \times 10^{-5}$ with a warmup+cosine schedule. Plot loss at step
        1{,}000 vs $\eta_{\max}$ to locate the instability knee.
  \item \textbf{Gradient clipping telemetry.} Log the clip activation
        rate and gradient norm $\|g_t\|_2$ every step. Correlate spikes
        in $\|g_t\|_2$ with subsequent loss jumps.
  \item \textbf{Mini-Chinchilla.} At fixed compute $C$, train
        $(N, D)$ pairs $(N, 20 N)$ and $(2N, 10 N)$ and $(N/2, 40 N)$.
        Verify that $(N, 20 N)$ yields the lowest loss.
  \item \textbf{Reproducibility audit.} Run the same training script
        twice with the same seed; diff the loss curves. Identify and
        eliminate all nondeterminism sources until the curves match to
        within numerical tolerance.
\end{enumerate}

\subsection{Week 2 Takeaway}

\begin{quote}
Training behavior in LLMs is governed less by ``intelligence'' than by
optimization dynamics operating in very high dimensions. Adam/AdamW,
warmup, cosine decay, gradient clipping, and mixed precision are not
interchangeable tuning knobs; each responds to a specific pathology of
large-scale stochastic optimization, and scaling laws turn the resulting
process into a compute-planning exercise that can be reasoned about
engineering-first.
\end{quote}
