%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% WEEK 1 -- STATISTICAL LEARNING FOUNDATIONS FOR LLMs
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{Week 1: Statistical Learning Foundations for Large Language Models}
\label{sec:week1}

\subsection{Learning Objectives}

By the end of this week, you should be able to:
\begin{itemize}
  \item Formally interpret a Large Language Model (LLM) as a conditional probabilistic model over sequences of discrete symbols.
  \item Re-derive the cross-entropy loss used in next-token prediction from first principles, both as a maximum-likelihood estimator and as a consequence of the information-theoretic identity $H(p,q) = H(p) + \KL(p \parallel q)$.
  \item Work through a softmax computation numerically and show why the logit representation is invariant under additive constant shifts (justifying the log-sum-exp trick).
  \item Distinguish the three quantities that are routinely conflated in downstream discussions: \emph{probability}, \emph{confidence}, and \emph{correctness}.
  \item Explain why cross-entropy minimization does not, on its own, imply semantic correctness, and why this is the root cause of several characteristic LLM failure modes.
  \item Reason quantitatively about model calibration using reliability diagrams, Expected Calibration Error (ECE), and temperature scaling.
  \item Explain the information-theoretic interpretation of training objectives, including the equivalence between cross-entropy and perplexity.
  \item State the standard proper scoring rules used in probabilistic forecasting and explain why log-loss is a natural choice.
  \item Identify the human-factor risks that arise when probabilistic outputs are interpreted as authoritative assertions, and describe first-line mitigations.
\end{itemize}

\subsection{LLM Components Covered}

This week focuses on the statistical substrate underlying all LLM systems. Every concept introduced here will reappear in later weeks under increasingly engineering-heavy disguises, so the aim is to establish vocabulary and intuition that later chapters can take as given.

\textbf{Modeling and Training}
\begin{itemize}
  \item Conditional probability modeling over discrete tokens.
  \item Maximum likelihood estimation (MLE) as a training principle.
  \item Cross-entropy as the empirical-risk objective.
  \item Softmax parameterization and logit representation.
  \item Calibration and uncertainty quantification.
\end{itemize}

\textbf{Systems Perspective}
\begin{itemize}
  \item Interpretation of model outputs as distributions rather than assertions.
  \item Implications of probabilistic behavior for downstream system design, monitoring, and human-in-the-loop review.
\end{itemize}

\subsection{Conceptual Core: What Kind of Object Is an LLM?}

A Large Language Model defines a probability distribution over sequences of discrete symbols drawn from a finite vocabulary $V$. We denote an individual sequence as $x = (x_1, x_2, \ldots, x_T)$ where each $x_t \in V$ and $|V|$ is typically on the order of $10^4$ to $10^5$ tokens for modern subword vocabularies. The object of study is the joint distribution
\begin{equation}
P_\theta(x_1, x_2, \ldots, x_T),
\label{eq:joint}
\end{equation}
parameterized by $\theta \in \R^d$, the collection of all trainable weights of the network.

Modelling the joint distribution directly is combinatorially infeasible: the sample space has cardinality $|V|^T$, which for even modest context lengths is astronomically large. Instead, using the chain rule of probability, we factor (\ref{eq:joint}) into an ordered product of conditionals:
\begin{equation}
P_\theta(x_1, \ldots, x_T) = \prod_{t=1}^{T} P_\theta(x_t \mid x_1, \ldots, x_{t-1}).
\label{eq:chain}
\end{equation}
This identity is exact and holds for any distribution over sequences. An LLM commits to a particular \emph{parameterization} of the conditional $P_\theta(x_t \mid x_{<t})$, where we use the shorthand $x_{<t} = (x_1, \ldots, x_{t-1})$. The chosen parameterization is the neural architecture (Week~4), and the weights $\theta$ are learned by fitting this conditional to empirical text (Week~5). Everything else---including the reasoning-like behavior, the hallucinations, and the striking context sensitivity---is a downstream consequence of this one modelling choice.

An immediate observation is that an LLM does \emph{not} explicitly store facts, rules, or symbols. It stores numerical parameters that, together with its architecture, implement a family of conditional distributions. The model's ``knowledge'' is whatever makes those conditionals assign high probability to corpus text. This framing is the single most important piece of intuition a systems engineer needs. Many behaviors that look like reasoning failures---confident fabrications, brittle chains of thought, inconsistency under paraphrase---follow directly from reading the model's outputs through a probabilistic lens rather than a symbolic one.

\subsubsection{Why Left-to-Right?}

The chain rule in equation (\ref{eq:chain}) is written left-to-right, but it could just as well be written right-to-left, or in any other ordering. Any particular factorization is mathematically equivalent on the joint. Neural language models nonetheless commit to a canonical ordering---left-to-right for GPT-style autoregressive models, or a masked variant for BERT-style bidirectional encoders---because the \emph{parameter sharing} induced by that ordering is what makes the model tractable. We return to this choice in Week~5, where we contrast autoregressive, masked, and prefix-LM objectives.

\subsection{Maximum Likelihood and Next-Token Prediction}

\subsubsection{Training Objective}

Suppose we are given a dataset of token sequences
\[
\mathcal{D} = \big\{ x^{(i)} = (x_1^{(i)}, \ldots, x_{T_i}^{(i)}) \big\}_{i=1}^{N}.
\]
Under an i.i.d. assumption across sequences, the likelihood of $\mathcal{D}$ under model $P_\theta$ is
\begin{equation}
P_\theta(\mathcal{D}) = \prod_{i=1}^{N} P_\theta(x^{(i)}) = \prod_{i=1}^{N} \prod_{t=1}^{T_i} P_\theta(x^{(i)}_t \mid x^{(i)}_{<t}).
\end{equation}
Maximum likelihood estimation chooses $\hat\theta = \arg\max_\theta \log P_\theta(\mathcal{D})$. Taking the logarithm makes the product tractable and replaces it with a sum:
\begin{equation}
\hat\theta = \arg\max_\theta \sum_{i=1}^{N} \sum_{t=1}^{T_i} \log P_\theta(x^{(i)}_t \mid x^{(i)}_{<t}).
\label{eq:mle}
\end{equation}
This is the training objective for essentially every modern language model. Every innovation in the last decade---Transformers, sparse mixtures, alignment, retrieval---is layered on top of equation (\ref{eq:mle}).

\subsubsection{Equivalence with Cross-Entropy}

Equation (\ref{eq:mle}) is most commonly rewritten in the language of empirical risk minimization. Define the empirical distribution $\hat p$ induced by $\mathcal{D}$: for each context $x_{<t}$ observed in the corpus, $\hat p(\cdot \mid x_{<t})$ is the empirical distribution of tokens following that context. Let $q_\theta(\cdot \mid x_{<t}) := P_\theta(\cdot \mid x_{<t})$. The negative log-likelihood per token is
\begin{align}
\mathcal{L}_{\text{NLL}}(\theta)
  &= -\frac{1}{N_{\text{tok}}} \sum_{i,t} \log q_\theta(x_t^{(i)} \mid x_{<t}^{(i)}) \\
  &= -\E_{(x_{<t}, x_t) \sim \hat p}\!\left[ \log q_\theta(x_t \mid x_{<t}) \right]
  = H(\hat p, q_\theta),
\label{eq:nll-ce}
\end{align}
where $N_{\text{tok}} = \sum_i T_i$. The object $H(\hat p, q_\theta)$ is the empirical cross-entropy between $\hat p$ and $q_\theta$. Thus maximizing log-likelihood under the data is \emph{identical} to minimizing cross-entropy against the empirical distribution. This is not a coincidence and is not specific to language modelling: it is a general fact about exponential-family distributions fitted by MLE.

To see the connection to KL divergence, note that for any fixed reference $p$,
\begin{equation}
H(p, q) = H(p) + \KL(p \parallel q).
\label{eq:ce-kl}
\end{equation}
Since $H(p)$ does not depend on $\theta$, minimizing cross-entropy against $p$ is equivalent to minimizing $\KL(p \parallel q_\theta)$. In training, $p = \hat p$ is the empirical distribution, so we are driving $q_\theta$ toward the corpus distribution in KL. This is a critical observation with a double edge: on the one hand it tells us that cross-entropy is the ``right'' loss in a principled sense; on the other, it reminds us that the target is the \emph{data} distribution, not the \emph{truth}. If the corpus contains systematic errors or biases, a low-cross-entropy model faithfully reproduces them.

\subsubsection{Softmax Parameterization and Numerical Stability}

The conditional $q_\theta(\cdot \mid x_{<t})$ is realized by the model as a vector of real-valued \emph{logits} $z \in \R^{|V|}$ computed by the final layer, followed by a softmax:
\begin{equation}
q_\theta(x_t = v \mid x_{<t})
  = \frac{\exp(z_v)}{\sum_{v' \in V} \exp(z_{v'})}.
\label{eq:softmax}
\end{equation}
Two properties of the softmax deserve attention. First, it is \emph{shift-invariant}: for any constant $c \in \R$, the distribution is unchanged if every logit is shifted by $c$. This invariance is the justification for the \emph{log-sum-exp} (LSE) trick used in every production implementation:
\begin{equation}
\log q_\theta(x_t = v \mid x_{<t}) = z_v - \underbrace{\log \sum_{v'} \exp(z_{v'})}_{=:\,\text{LSE}(z)}.
\end{equation}
Naively computing $\exp(z_v)$ for large logits overflows 32-bit floats; subtracting $\max_v z_v$ before exponentiating is numerically equivalent and keeps intermediate values bounded.

Second, softmax couples the probabilities of all tokens: increasing the logit of one token necessarily reduces the probability of every other token (the total must sum to one). This has a subtle consequence for interpretation: a probability of $0.9$ on token $A$ does not mean ``the model is 90\% confident in $A$ as a fact''---it means ``$A$ dominates all other tokens combined by a factor of roughly $9{:}1$ under this parameterization.'' The relative nature of softmax probabilities is the source of many calibration problems we will examine below.

\subsubsection{Worked Example}

Consider a toy vocabulary $V = \{\text{\emph{yes}}, \text{\emph{no}}, \text{\emph{maybe}}\}$ and a context for which the model produces logits $z = (2.0,\ 1.0,\ 0.1)$. The unnormalized exponentials are $(\exp(2),\ \exp(1),\ \exp(0.1)) \approx (7.389,\ 2.718,\ 1.105)$, summing to $\approx 11.212$. The resulting probabilities are
\[
q(\text{\emph{yes}}) \approx 0.659,\quad
q(\text{\emph{no}}) \approx 0.242,\quad
q(\text{\emph{maybe}}) \approx 0.099.
\]
If the ground-truth token is ``yes,'' the per-token cross-entropy contribution is $-\log 0.659 \approx 0.417$ nats. If we now shift every logit by $+1000$, the probabilities are numerically identical, but a naive implementation would overflow immediately. Subtracting the max ($+1002$ after shift) yields shifted logits $(-0,\ -1,\ -1.9)$, which is exactly $z - \max z$ in the original frame. Shift-invariance is therefore not a theoretical nicety; it is a daily engineering tool.

\subsubsection{The Decomposition of Cross-Entropy Into Per-Position Losses}

Training divides cleanly along positions. Because equation (\ref{eq:mle}) sums over positions $t$ as well as over sequences, the total loss on a batch decomposes exactly into a sum of per-position NLL contributions, each equivalent to a multinomial logistic regression problem over $V$. This decomposition is what makes teacher-forcing training efficient: a single forward pass computes the conditional distribution at every position in parallel, and backpropagation distributes the gradient across all of them simultaneously. It also means that a mis-tokenized example (one where a rare token has been split unusually) contributes anomalously large loss spikes to specific positions---a phenomenon we revisit in Week~3.

\subsection{Information-Theoretic Foundations}

To reason about training curves, perplexity benchmarks, and calibration, we need a modest amount of classical information theory. The canonical reference is \citet{cover2006elements}; the textbook treatment is \citet{bishop2006prml} or \citet{murphy2022pml}.

\subsubsection{Entropy}

Let $p$ be a distribution on a discrete set $\mathcal{X}$. Its \emph{entropy} in nats is
\begin{equation}
H(p) = -\sum_{x \in \mathcal{X}} p(x) \log p(x)
     = \E_{x \sim p}[-\log p(x)],
\end{equation}
and in bits the definition swaps $\log$ for $\log_2$. Entropy measures the average surprise under $p$: distributions concentrated on a single outcome have $H = 0$; the uniform distribution on $|V|$ outcomes has $H = \log |V|$ (the maximum). \citet{shannon1948} introduced this quantity as the expected number of bits required to encode a random draw from $p$ under the optimal code.

\subsubsection{Cross-Entropy and KL Divergence}

The cross-entropy between $p$ and $q$ is
\begin{equation}
H(p, q) = -\sum_{x} p(x) \log q(x) = \E_{x \sim p}[-\log q(x)],
\end{equation}
the expected number of bits required to encode samples from $p$ using a code optimized for $q$. The gap between $H(p, q)$ and $H(p)$ is the \emph{Kullback--Leibler divergence}:
\begin{equation}
\KL(p \parallel q) = H(p, q) - H(p) = \sum_x p(x) \log \frac{p(x)}{q(x)}.
\end{equation}
$\KL$ is non-negative, zero iff $p = q$, and \emph{asymmetric}---a crucial property that reappears in reverse-KL alignment objectives in Week~8.

The equation $H(p, q) = H(p) + \KL(p \parallel q)$ is the identity we used in equation (\ref{eq:ce-kl}). It makes transparent what cross-entropy training accomplishes: it minimizes the additional coding cost incurred by using $q$ in place of the true distribution $p$.

\subsubsection{Mutual Information}

For two random variables $X$ and $Y$ with joint $p(x, y)$ and marginals $p(x), p(y)$, the mutual information is
\begin{equation}
I(X; Y) = \KL\!\big(p(x, y) \parallel p(x)p(y)\big) = H(X) - H(X \mid Y).
\end{equation}
$I(X; Y)$ quantifies how much uncertainty about $X$ is resolved by observing $Y$. In LLM work, mutual information appears in several guises: in information-bottleneck arguments about representation learning, in the analysis of in-context learning (where the context tokens reduce uncertainty about the continuation), and in retrieval evaluation (where good retrieval increases $I(\text{answer}; \text{context})$).

\subsubsection{Perplexity}

Perplexity is the standard headline metric for language models. It is defined as the exponential of per-token cross-entropy:
\begin{equation}
\operatorname{PPL}(q_\theta; \mathcal{D})
  = \exp\!\left( \frac{1}{N_{\text{tok}}} \sum_{i,t} -\log q_\theta(x_t^{(i)} \mid x_{<t}^{(i)}) \right)
  = \exp\big( H(\hat p, q_\theta) \big).
\label{eq:perplexity}
\end{equation}
Perplexity has an appealing interpretation: it is the ``effective vocabulary size'' the model is uncertain over at each step. A perplexity of $1$ corresponds to a perfectly deterministic predictor; a perplexity of $|V|$ corresponds to the uniform distribution. Because $\exp$ is monotone, minimizing cross-entropy and minimizing perplexity are the same optimization problem---but perplexity is easier to compare across datasets and model sizes on a log scale.

Two caveats merit explicit mention. First, perplexity is \emph{tokenization-dependent}: a model with a smaller vocabulary mechanically achieves lower per-token perplexity because each token is less informative. Comparing perplexities across models trained with different tokenizers is therefore misleading, and we will see in Week~3 that this is a recurring source of confusion. Second, a low perplexity does not imply high task performance. A model can be excellent at modelling the statistics of Wikipedia prose while still being useless at answering a question. Week~9 is where we confront task-level evaluation directly.

\subsection{Uncertainty, Confidence, and Calibration}

\subsubsection{Three Quantities That Look the Same and Are Not}

A common source of error in LLM engineering is the collapse of three distinct quantities into a single intuitive notion of ``the model knows.'' They are:
\begin{itemize}
  \item \textbf{Probability}: the scalar $q_\theta(v \mid x_{<t})$ the model assigns to a specific token $v$ given its parameters and context. This is what the softmax computes.
  \item \textbf{Confidence}: a summary statistic of the full distribution $q_\theta(\cdot \mid x_{<t})$, typically operationalized as $1 - H(q_\theta) / \log |V|$ (normalized entropy), or as $\max_v q_\theta(v \mid x_{<t})$ (top-1 probability), or as $\log q_\theta(v_1) - \log q_\theta(v_2)$ (log-odds between top-2).
  \item \textbf{Correctness}: whether the sampled or argmax token matches an external ground truth. Correctness is defined relative to something outside the model---a labeler, a gold answer, a runtime assertion.
\end{itemize}
These three can, and routinely do, move independently. A model can assign $0.9$ probability to the correct token (high probability, high confidence, correct); assign $0.9$ to an incorrect token (high probability, high confidence, \emph{incorrect}); or spread probability evenly over five plausible tokens of which one is correct (low confidence, correct-on-average). The difference between the second and third cases is the entire subject of \emph{calibration}.

\subsubsection{Calibration and Reliability Diagrams}

A probabilistic classifier is said to be \emph{perfectly calibrated} if, among the set of predictions on which it outputs probability $p^\star$, the fraction of correct predictions equals $p^\star$. Formally, for any $p^\star \in [0, 1]$,
\begin{equation}
\Pr\!\big[\text{correct} \,\big|\, \hat q = p^\star \big] = p^\star.
\end{equation}
Perfect calibration is an \emph{extra} property beyond accuracy: a classifier can be accurate but uncalibrated (its confidences misstate the reliability) or calibrated but inaccurate (uniformly low confidence on a hard problem).

Reliability diagrams are the standard visual diagnostic. Bin the predicted probabilities into $K$ bins $B_1, \ldots, B_K$, and for each bin plot the mean predicted probability against the empirical accuracy. Perfect calibration corresponds to points on the diagonal; overconfidence lies below the diagonal (predicted $>$ actual), underconfidence above. Modern deep networks including LLMs are almost invariably \emph{overconfident} in the high-probability regime \citep{guo2017calibration}.

\subsubsection{Expected Calibration Error}

Reliability diagrams are qualitative; to quantify calibration we use the \emph{Expected Calibration Error} (ECE):
\begin{equation}
\operatorname{ECE} = \sum_{k=1}^{K} \frac{|B_k|}{N}\,\big| \operatorname{acc}(B_k) - \operatorname{conf}(B_k) \big|,
\end{equation}
where $\operatorname{acc}(B_k)$ is the empirical accuracy of predictions in bin $k$, $\operatorname{conf}(B_k)$ is the mean predicted probability in the bin, $|B_k|$ is the bin count, and $N = \sum_k |B_k|$ is the total number of predictions. ECE is a weighted average of absolute calibration gaps. A well-calibrated predictor has $\operatorname{ECE} \approx 0$; values above a few percent indicate a practically significant miscalibration.

ECE has known weaknesses---it is sensitive to binning, it ignores the sign of miscalibration, and it conflates different calibration failure modes. Alternatives include the Brier score,
\begin{equation}
\operatorname{Brier} = \frac{1}{N} \sum_{i=1}^{N} \sum_{v \in V} \big( q_i(v) - \mathbb{1}[y_i = v] \big)^2,
\end{equation}
and the marginal log-loss, which is simply per-instance NLL. Each of these is a \emph{proper scoring rule} in the sense defined below.

\subsubsection{Temperature Scaling}

Let $T > 0$ be a scalar. \emph{Temperature scaling} divides logits by $T$ before the softmax:
\begin{equation}
q_T(v \mid x_{<t}) = \frac{\exp(z_v / T)}{\sum_{v'} \exp(z_{v'} / T)}.
\end{equation}
Temperature $T < 1$ sharpens the distribution (more mass on the top-1); $T > 1$ flattens it; $T \to 0^+$ degenerates to a point mass on the argmax; $T \to \infty$ degenerates to uniform. Crucially, temperature preserves the \emph{ordering} of all tokens---it cannot turn a low-probability token into the most probable one.

Temperature scaling is the simplest and most common post-hoc calibration method. A single $T$ is fit on held-out data to minimize NLL; because $T$ is scalar, it does not change argmax predictions, and it has been shown empirically to recover most of the calibration deficit in image classifiers and language models \citep{guo2017calibration}. Temperature is also the primary decoding-time knob in inference stacks: the same mathematical operation that fixes calibration can be used to trade off diversity for determinism in generation. Temperature does not introduce new knowledge; it redistributes existing mass.

\subsubsection{Proper Scoring Rules}

A scoring rule $S(q, y)$ assigns a score to a predicted distribution $q$ in light of an observed outcome $y$. A scoring rule is \emph{proper} if reporting the true distribution is, in expectation, optimal:
\begin{equation}
\E_{y \sim p}[S(p, y)] \le \E_{y \sim p}[S(q, y)] \quad \text{for all } q.
\end{equation}
(The inequality is with respect to losses; reverse it if $S$ is a gain.) Proper scoring rules incentivize honest uncertainty reporting; improper ones reward overconfident predictions. Log-loss ($S(q, y) = -\log q(y)$) and the Brier score are the two most common proper scoring rules in classification. The log-loss we minimize during training is therefore not merely convenient; it is provably proper. The canonical survey is \citet{gneiting2007strictly}.

This is the theoretical backstop for cross-entropy training: we are not just minimizing some arbitrary convex surrogate; we are minimizing a proper scoring rule that is optimal \emph{in expectation} when $q = p$. Equivalently, the uniquely optimal strategy under log-loss is to report the true conditional distribution. The gap between that optimum and what a trained model actually does tells us how far we are from the optimum---and, informally, where there is room to grow.

\subsection{Sampling and Decoding as Probabilistic Objects}

Once the model has computed $q_\theta(\cdot \mid x_{<t})$, we must choose a token. This step is frequently treated as incidental, but it materially shapes system behavior.

\begin{itemize}
  \item \textbf{Greedy decoding} selects $\arg\max_v q_\theta(v \mid x_{<t})$. It is deterministic but pathological in long sequences, as it cannot recover from early mistakes and tends to produce repetitive, degenerate text \citep{holtzman2020curious}.
  \item \textbf{Stochastic sampling} draws $x_t \sim q_\theta(\cdot \mid x_{<t})$. This yields diverse outputs but can produce low-probability (and low-quality) tokens in the distribution's long tail.
  \item \textbf{Top-$k$ sampling} restricts sampling to the $k$ highest-probability tokens, renormalizes, and samples. It cuts the tail but ignores the shape of the head.
  \item \textbf{Top-$p$ (nucleus) sampling} restricts to the smallest set whose cumulative mass exceeds a threshold $p$, and samples from that set. It adapts to the shape of the distribution: sharp distributions yield small nuclei, flat distributions yield large ones.
  \item \textbf{Temperature} modifies any of the above by pre-scaling the logits. $T \to 0^+$ recovers greedy; $T \to \infty$ flattens to uniform.
\end{itemize}

A useful mental model is that decoding strategies define a \emph{second distribution} over outputs, layered on top of the model's learned conditional. The question ``what does the model say?'' is therefore under-specified without a decoding strategy. Week~5 treats decoding more formally; for now, the point is that sampling is a first-class design choice with measurable effects on calibration, diversity, and downstream task performance.

\subsection{Common Probabilistic Failure Modes}

With the probabilistic framing in place, several characteristic LLM failure modes resolve into familiar statistical pathologies.

\paragraph{Confident fabrication (hallucination).}
The model assigns high probability to a token sequence that happens to be factually wrong. Under the probabilistic lens this is unremarkable: the training objective (\ref{eq:mle}) rewards high probability on \emph{corpus-likely} sequences, and a fluent fabrication can be corpus-likely if the corpus contains many fluent fabrications in adjacent contexts. No component of cross-entropy training pushes the model toward truth; truth is not a variable in the optimization.

\paragraph{Overconfidence in rare events.}
Modern deep networks tend to concentrate mass aggressively. In the extreme, argmax probabilities drift toward 1 as models are scaled and trained longer---beyond what the empirical accuracy in that bin justifies. This is the phenomenon that reliability diagrams expose and temperature scaling partially corrects.

\paragraph{Token-boundary brittleness.}
Because cross-entropy is computed per token and tokenization is lossy (Week~3), semantically equivalent inputs can produce very different logit trajectories. The model may be confident on one phrasing of a prompt and unsure on a paraphrase. This is a calibration failure expressed at the level of inputs rather than outputs.

\paragraph{Exposure bias.}
During training the model always conditions on \emph{ground-truth} prefixes (teacher forcing). At inference it conditions on its own generations. Any small probability mass on wrong tokens at position $t$ compounds over positions $t+1, t+2, \ldots$, leading to distributional drift that the loss itself did not penalize. This is a structural consequence of the decomposition in equation (\ref{eq:mle}) and is one reason why long-form generation is harder than short-form.

\paragraph{Calibration drift after alignment.}
Post-training techniques (instruction tuning, RLHF; Week~8) reliably reshape \emph{which} tokens get probability mass without preserving \emph{how much} mass they get. Aligned models are often less calibrated than their base counterparts---the confidence signal that was useful in the base model becomes misleading after alignment. This is a critical consideration in production pipelines and is one of the motivations for retrieval-augmented and tool-using systems (Week~9).

\paragraph{A worked overconfidence vignette.}
Suppose a factual QA benchmark shows that an LLM's top-1 answer has $88\%$ average top-1 probability but only $71\%$ top-1 accuracy. The $17$-percentage-point gap is the calibration error. If a downstream system uses the top-1 probability as a confidence-for-abstention signal---abstaining whenever probability is below $0.7$---it will never abstain on the hard cases, because probability clusters above $0.7$ even when accuracy is poor. Fixing this is primarily a calibration task (temperature scaling on a held-out split), not a model-capacity task. A useful system-design principle follows: \emph{if a downstream decision depends on a threshold on model probability, validate that the probability is calibrated on the operating distribution before deployment.}

\subsection{Systems Engineering Implications}

From a systems engineering standpoint, an LLM should be treated as a stochastic subsystem with an opaque internal state and non-stationary performance. The following table summarizes the consequences of the statistical framing for each phase of a standard V-model:

\begin{center}
\begin{tabular}{|L{3.4cm}|L{8.0cm}|}
\hline
\textbf{Phase / Concern} & \textbf{Consequence of the Statistical Framing} \\
\hline
Requirements & Specifications must admit probabilistic outputs; ``the system shall return the correct answer'' is not enforceable against a probabilistic model. \\
\hline
Architecture & Downstream components must tolerate variance. Verification layers, retrieval, tool use, and human-in-the-loop review are architectural patterns motivated directly by the probabilistic substrate. \\
\hline
Verification \& Validation & Evaluation must report distributions, not point estimates. Calibration metrics (ECE, Brier) belong in V\&V alongside accuracy. \\
\hline
Operations & Distribution shift is inevitable: input distributions change over time. Continuous monitoring of loss, entropy, and acceptance rates is the operational analogue of calibration. \\
\hline
Incident Response & Failures are rarely deterministic bugs. Root-cause analysis must consider rare-event tail behavior and distributional assumptions. \\
\hline
\end{tabular}
\end{center}

These properties motivate the architectural patterns---verification layers, retrieval augmentation, structured output formats, and human-in-the-loop review---that Weeks 8 through 10 will develop in detail.

\subsection{Human Factors and Failure Modes}

A common early failure mode in LLM systems is \emph{automation bias}:
\begin{quote}
``The model sounds confident, therefore it must be correct.''
\end{quote}
This arises from a mismatch between human epistemic intuitions---where confident assertions come from knowledgeable interlocutors---and probabilistic text generation, where confidence is a statistical property of the next-token distribution. Humans infer reliability from tone; models produce tone as a by-product of fitting tone-heavy text. The fluency heuristic is a cousin: the more coherent the output reads, the more we trust it, independent of whether it is correct.

Effective first-line mitigations include structured outputs with explicit confidence fields, uncertainty visualization in user interfaces, and source attribution (retrieval citations, tool-call traces). We return to these in Week~9 in the context of system design and in Week~10 in the context of assurance cases.

\subsection{Bridges to Later Chapters}

The statistical machinery of this chapter is revisited, refined, or challenged in every subsequent week. Explicit bridges worth keeping in mind:
\begin{itemize}
  \item \textbf{Week 2 (Optimization)} studies how we actually find $\hat\theta$ in practice, and why the optimization dynamics of cross-entropy over deep networks defy the intuition of classical statistics.
  \item \textbf{Week 3 (Tokenization \& Embeddings)} makes precise what $x_t$ is as an object: a token rather than a word, and embedded in a vector space rather than a categorical index.
  \item \textbf{Week 4 (Transformers)} specifies how the conditional $q_\theta(\cdot \mid x_{<t})$ is computed from the prefix, filling in the black box that the derivations above took for granted.
  \item \textbf{Week 5 (Pretraining \& ICL)} revisits perplexity, decoding, and the emergence of in-context adaptation as a property of the pretraining objective.
  \item \textbf{Week 7 (Adaptation)} and \textbf{Week 8 (Alignment)} change $q_\theta$ after pretraining. Both typically \emph{reduce} calibration relative to the base model, which forces the discussion from this chapter to be revisited operationally.
  \item \textbf{Week 9 (RAG \& Tools)} substitutes $q_\theta(y \mid x)$ with $q_\theta(y \mid x, d)$ where $d$ is retrieved evidence, converting a pure language-modelling task into a conditional inference task with grounded context.
  \item \textbf{Week 10 (Governance \& Assurance)} treats probabilistic performance as an assurance-case claim and asks what evidence is needed to justify it.
\end{itemize}

\subsection{Key References}

\begin{itemize}
  \item \citet{bishop2006prml} --- \emph{Pattern Recognition and Machine Learning}; definitive treatment of probabilistic modelling and calibration.
  \item \citet{murphy2012mlpp} and \citet{murphy2022pml} --- \emph{Machine Learning: A Probabilistic Perspective} and the updated \emph{Probabilistic Machine Learning}; covers MLE, exponential families, and information theory.
  \item \citet{shannon1948} --- \emph{A Mathematical Theory of Communication}; the foundational paper introducing entropy, cross-entropy, and the source-coding theorems that underpin perplexity.
  \item \citet{cover2006elements} --- \emph{Elements of Information Theory}; the canonical reference for information-theoretic quantities.
  \item \citet{vapnik1998statistical} --- \emph{Statistical Learning Theory}; classical PAC framework for generalization, useful background for Week~2.
  \item \citet{guo2017calibration} --- empirical study of calibration in modern neural networks; introduces temperature scaling.
  \item \citet{gneiting2007strictly} --- survey of proper scoring rules in forecasting.
  \item \citet{holtzman2020curious} --- shows that maximum-likelihood decoding fails in open-ended generation; motivates the decoding toolkit.
\end{itemize}

\subsection{Recommended University Lectures}

\begin{itemize}
  \item Caltech CS156 --- \emph{Learning From Data} \citep{caltech_cs156}.
  \item MIT 6.036 --- \emph{Introduction to Machine Learning} \citep{mit_6036}.
  \item Stanford CS229 --- \emph{Machine Learning}, particularly the probability and optimization sections \citep{stanford_cs229}.
\end{itemize}

\subsection{Practitioner Lab}

\textbf{Objective}: implement multinomial logistic regression for a toy next-token task and observe probabilistic behavior first-hand.

\textbf{Tasks}:
\begin{enumerate}
  \item Train a simple next-token classifier on a short corpus.
  \item Inspect the output probability distributions on held-out inputs; plot entropies and top-$k$ gaps.
  \item Measure accuracy, entropy, and calibration error (ECE) on a held-out split.
  \item Fit a scalar temperature on a validation split; re-measure ECE; verify that accuracy is unchanged.
  \item Sample generations at $T \in \{0.2, 0.7, 1.0, 1.4\}$ and qualitatively compare.
  \item Construct a case where top-1 probability exceeds $0.9$ but the top-1 prediction is wrong; discuss what a downstream system should do in response.
\end{enumerate}

\textbf{Deliverable}: a Jupyter notebook with code and plots, plus a one-page reflection titled ``Where my probabilistic intuition failed.''

See \texttt{labs/week01/week01\_lab\_probabilistic\_foundations.ipynb}.

\subsection{Week 1 Takeaway}

\begin{quote}
An LLM is not a knowledge base or a reasoner; it is a probabilistic sequence model trained by minimizing cross-entropy against a corpus distribution. Cross-entropy is a proper scoring rule, which makes it a principled training target---but it measures distributional fidelity, not truth. Everything downstream, from hallucination to overconfidence to the surprising power of prompting, is a consequence of reading the model's outputs as probabilities and not as assertions.
\end{quote}

All subsequent components---Transformers, scaling, adaptation, alignment, retrieval, tools, agents, and governance---build on this foundation.
