\section{Week 9: LLM Systems — Retrieval, Tool Use, and Evaluation Harnesses}

\subsection{Learning Objectives}

By the end of this week, you should be able to:
\begin{itemize}
  \item Reason about LLMs as components in larger systems
  \item Explain Retrieval-Augmented Generation (RAG)
  \item Understand tool use as structured I/O, not “reasoning”
  \item Design evaluation harnesses for LLM systems
  \item Identify system-level failure modes and silent errors
\end{itemize}

\subsection{LLM Components Covered}

\textbf{System architecture}
\begin{itemize}
  \item Retrieval pipelines
  \item Prompt construction
  \item Tool interfaces
\end{itemize}

\textbf{Evaluation}
\begin{itemize}
  \item Task-specific test sets
  \item Regression testing
  \item Human-in-the-loop evaluation
\end{itemize}

\subsection{Conceptual Core: LLMs Are Not Standalone Systems}

Modern LLM deployments rarely rely on the model alone.
Instead, they combine:
\begin{itemize}
  \item pretrained models,
  \item external knowledge sources,
  \item structured tools,
  \item application logic.
\end{itemize}

The system boundary matters more than the model internals.

\subsection{Retrieval-Augmented Generation (RAG)}

RAG augments generation by conditioning on retrieved documents:

\[
P(y \mid x) \rightarrow P(y \mid x, d_1, \dots, d_k)
\]

where $d_i$ are retrieved from an external corpus.

Benefits:
\begin{itemize}
  \item reduced hallucination,
  \item up-to-date knowledge,
  \item traceable sources.
\end{itemize}

Risks:
\begin{itemize}
  \item retrieval errors,
  \item prompt injection,
  \item false sense of grounding.
\end{itemize}

\subsection{Tool Use as Interface Design}

Tool use is often framed as “reasoning,” but operationally it is:
\begin{itemize}
  \item structured input/output,
  \item constrained action spaces,
  \item orchestration by external code.
\end{itemize}

The LLM proposes actions; the system executes them.

\subsection{Prompt Construction as a System Artifact}

Prompts encode:
\begin{itemize}
  \item task specification,
  \item retrieved context,
  \item safety constraints.
\end{itemize}

They must be:
\begin{itemize}
  \item versioned,
  \item tested,
  \item audited.
\end{itemize}

Prompt changes are system changes.

\subsection{Evaluation Harnesses}

Evaluating LLM systems requires:
\begin{itemize}
  \item representative test cases,
  \item clear success criteria,
  \item automated regression detection.
\end{itemize}

Naive metrics (e.g.\ BLEU) are often misleading.

\subsection{Systems Engineering Implications}

\textbf{Requirements}
\begin{itemize}
  \item Correctness under distribution shift
  \item Graceful failure modes
\end{itemize}

\textbf{Architecture}
\begin{itemize}
  \item Modular components
  \item Clear interfaces and contracts
\end{itemize}

\textbf{Verification \& Validation}
\begin{itemize}
  \item End-to-end testing
  \item Red-teaming of retrieval and tools
\end{itemize}

\textbf{Operations}
\begin{itemize}
  \item Logging and traceability
  \item Continuous evaluation
\end{itemize}

\subsection{Human Factors and Failure Modes}

Common risks include:
\begin{itemize}
  \item automation bias,
  \item overtrust in retrieved content,
  \item opaque failure chains.
\end{itemize}

Users often attribute errors to the model rather than the system.

\subsection{Key References}

\begin{itemize}
  \item \citet{lewis2020retrieval} — RAG
  \item \citet{yao2023react} — tool use and reasoning-action loops
\end{itemize}

\subsection{Recommended University Lectures}

\begin{itemize}
  \item Stanford CS324 — LLM systems and evaluation  
  \url{https://stanford-cs324.github.io/winter2022/}

  \item Berkeley AI Systems lectures  
  \url{https://bair.berkeley.edu}
\end{itemize}

\subsection{Practitioner Lab}

See \texttt{labs/week09/week09\_lab\_llm\_systems\_rag\_eval.ipynb}.

\subsection{Week 9 Takeaway}

\begin{quote}
Most LLM failures are system failures, not model failures.
\end{quote}


