\chapter{Chapter 5 Refresher: The Transformer --- Attention as the Core Computational Primitive --- Attention, Shapes, and Complexity}

\scenelabel
Attention replaces recurrence's sequential bottleneck with explicit pairwise comparisons implemented as batched matrix multiply---linear algebra becomes architecture instead of notation garnish. But attention alone is position-blind: the formula $\mathrm{softmax}(QK^\top/\sqrt{d_k})V$ treats all positions identically unless position is injected explicitly. This chapter covers both the attention mechanism and the positional encoding family that completes it.

\paragraph{Step A --- bilinear compatibility.}
\textbf{Problem.} Position \(t\) must aggregate information from other positions weighted by relevance.
\textbf{Insight.} Dot products score compatibility between transformed representations (queries vs keys).
\textbf{Lineage.} Bahdanau-style attention (2015); scaled dot-product formulation crystallised in Vaswani et al.\ (2017).
\textbf{Mental model.} Think ``matching embeddings,'' not mysticism---everything reduces to tensor contractions.

\paragraph{Step B --- softmax as differentiable selector.}
\textbf{Problem.} Hard argmax blocks gradients; we need smooth convex combinations.
\textbf{Insight.} Softmax (Chapter~1 tools) turns logits into mixing weights---Preface Step~10 intuition specialised row-wise.
\textbf{Lineage.} Statistical mechanics exponential tilts $\to$ neural attention gates.
\textbf{Mental model.} Temperature inherited from decoding chapters scales sharpness identically in spirit.

\paragraph{Step C --- masks encode causal laws.}
\textbf{Problem.} Autoregressive generation forbids peeking rightward.
\textbf{Insight.} Additive \(-\infty\) constraints zero softmax mass---enforce equality constraints algebraically before nonlinear squash.
\textbf{Lineage.} Decoder masking folklore codified in Transformer stacks.
\textbf{Mental model.} Same trick handles padding tokens---different semantics, identical mathematics.

\paragraph{Step D --- quadratic surface area.}
\textbf{Problem.} Pairwise attention allocates \(T\times T\) logits per layer.
\textbf{Insight.} Long contexts explode memory/compute---motivates FlashAttention-style engineering outside this refresher's scope but inside systems reasoning.
\textbf{Lineage.} Dao et al.\ IO-aware kernels (2022 era).
\textbf{Mental model.} Preface Step~6 variance arguments resurface when chunking randomises attention pathways.

\paragraph{Step E --- injecting position.}
\textbf{Problem.} Attention is permutation-equivariant; without positional information ``the cat sat on the mat'' and ``the mat sat on the cat'' produce identical representations.
\textbf{Insight.} Add a position-dependent signal to each token representation before or inside attention. Three families appear in the main text: absolute sinusoidal (Vaswani), relative bias (ALiBi), and rotation-based (RoPE).
\textbf{Lineage.} Vaswani (2017) sinusoids $\to$ Shaw et al.\ relative encodings (2018) $\to$ Su et al.\ RoPE (2021) $\to$ Press et al.\ ALiBi (2022).
\textbf{Mental model.} RoPE rotates each query and key by an angle proportional to its position; relative position appears as the \emph{difference} of angles, which survives the dot product.

If shapes align, attention is honest matrix cookbookery---when shapes drift, bugs hide silently.

\section*{Notation Introduced in This Chapter}
\begin{itemize}[leftmargin=1.5em]
  \item $T$: sequence length.
  \item $d_{\text{model}}$: model hidden width.
  \item $H$: number of attention heads.
  \item $d_k$: per-head key/query dimension.
  \item $Q,K,V$: query/key/value matrices.
  \item $M$: additive mask matrix applied before softmax.
  \item $\mathbf{R}_\theta$: $2\times 2$ rotation matrix by angle $\theta$.
  \item $m, n$: token position indices; $\theta_i$: per-dimension base frequency in RoPE.
\end{itemize}

\section*{What You Need To Recall Precisely}
\begin{itemize}[leftmargin=1.5em]
  \item \textbf{Scaled attention:}
  \[
    \mathrm{Attn}(Q,K,V)=\mathrm{softmax}\!\left(\frac{QK^\top+M}{\sqrt{d_k}}\right)V.
  \]
  Read row-wise: row $t$ of $QK^\top$ lists dot-product similarities between position-$t$ query vectors and every key vector; softmax turns those logits into nonnegative weights summing to $1$; multiplying those weights into $V$ mixes value vectors---attention outputs are convex combinations of values attended-to.

  \textbf{Why divide by $\sqrt{d_k}$?}
  If components of queries and keys are $\mathcal{O}(1)$ with roughly independent variation across $d_k$ dimensions, dot-product magnitudes grow \(\Theta(\sqrt{d_k})\) in expectation (length scales like Euclidean norm). Without scaling, softmax logits become sharply peaked as width grows, driving gradients toward sparse argmax-like behaviour and harming trainability. Rescaling keeps typical logits \(\mathcal{O}(1)\) across widths.

  \item \textbf{Mask semantics:} causal mask forbids looking right; padding mask blocks non-content locations.
  \item \textbf{Head decomposition:} independent projections create multiple interaction subspaces.
  \item \textbf{Residual and normalization:} residual streams pass representations forward as \(x\leftarrow x+\mathrm{Block}(x)\), letting blocks learn corrections around identity rather than full maps from scratch. Normalisation (LayerNorm/RMSNorm inside Transformer blocks) recentres/rescales activations per token or position so dot products do not drift orders of magnitude mid-training---pairs with residual paths to control gradient scale through depth.
  \item \textbf{Complexity source:} score matrix is $T\times T$, so interaction cost grows with sequence length.

  \item \textbf{Positional encodings --- three families.}

  \emph{Sinusoidal (absolute).} Add a fixed vector $\mathbf{p}_t \in \mathbb{R}^{d_{\text{model}}}$ to each token embedding before the first layer:
  \[
    p_{t,2i}   = \sin\!\left(\frac{t}{10000^{2i/d_{\text{model}}}}\right),\quad
    p_{t,2i+1} = \cos\!\left(\frac{t}{10000^{2i/d_{\text{model}}}}\right).
  \]
  Different dimensions oscillate at different frequencies, giving each position a unique fingerprint. The key property: $\mathbf{p}_t \cdot \mathbf{p}_s$ depends only on the offset $t - s$, so relative distance is encoded implicitly.

  \emph{ALiBi (relative bias).} Instead of modifying embeddings, subtract a head-specific linear bias from attention logits before softmax:
  \[
    \mathrm{score}_{t,s} = \frac{\mathbf{q}_t \cdot \mathbf{k}_s}{\sqrt{d_k}} - m \cdot |t - s|,
  \]
  where $m > 0$ is a per-head slope. Distant tokens receive larger penalties; the model learns to prefer local context without any positional vector in the embedding.

  \emph{RoPE (rotary position embedding).} Encode position by \emph{rotating} query and key vectors before the dot product. For a pair of scalar dimensions $(x_{2i}, x_{2i+1})$ at position $t$, apply a 2D rotation by angle $t\,\theta_i$:
  \[
    \mathbf{R}_{t\theta_i}
    = \begin{pmatrix} \cos t\theta_i & -\sin t\theta_i \\
                      \sin t\theta_i &  \cos t\theta_i \end{pmatrix},
    \qquad \theta_i = 10000^{-2i/d_k}.
  \]
  The critical property follows from the rotation group: for any two positions $m$ and $n$,
  \[
    (\mathbf{R}_{m\theta_i}\,\mathbf{x})^\top
    (\mathbf{R}_{n\theta_i}\,\mathbf{y})
    = \mathbf{x}^\top \mathbf{R}_{(n-m)\theta_i}\,\mathbf{y}.
  \]
  The dot product of a rotated query at position $m$ with a rotated key at position $n$ depends only on the \emph{relative} offset $n - m$, not on the absolute positions. This is the algebraic reason RoPE gives length-generalisation properties that absolute encodings lack.

  \textbf{Why this matters for Chapter~5.} The reader derives all three families. RoPE is the default in Llama, Mistral, and most open-weight models released since 2023; understanding the rotation-matrix identity is the prerequisite for following that derivation.
\end{itemize}

\section*{How These Results Build on Each Other}
\begin{enumerate}[leftmargin=1.5em]
  \item Project input vectors into $Q,K,V$.
  \item Form pairwise compatibility scores via $QK^\top$.
  \item Apply structural constraints with mask $M$.
  \item Normalize row-wise with softmax.
  \item Aggregate values and return contextualized representations.
\end{enumerate}

\section*{Reminder Glossary (Term by Term)}
\begin{itemize}[leftmargin=1.5em]
  \item \textbf{Query:} representation of what current position is asking for.
  \item \textbf{Key:} representation of what each position offers.
  \item \textbf{Value:} content payload mixed by attention weights.
  \item \textbf{Mask:} hard structural restriction on allowable dependencies.
  \item \textbf{Head:} one parallel attention subspace.
  \item \textbf{KV cache:} stored keys/values for efficient autoregressive decoding.
  \item \textbf{Rotation matrix $\mathbf{R}_\theta$:} a $2\times2$ matrix that rotates a 2D vector by angle $\theta$; its transpose is its inverse ($\mathbf{R}_\theta^\top = \mathbf{R}_{-\theta}$), so $\mathbf{R}_\theta^\top \mathbf{R}_\phi = \mathbf{R}_{\phi-\theta}$.
  \item \textbf{RoPE:} rotary position embedding; encodes absolute position as a rotation so that dot products yield relative-position information.
  \item \textbf{ALiBi:} attention with linear biases; penalises long-range attention logits with a linear distance term, no positional vectors needed.
\end{itemize}

\section*{Worked Mini Example (Shape Sanity)}
Let $T=4$, $d_{\text{model}}=8$, heads $H=2$, so $d_k=4$.
\[
Q,K,V \in \mathbb{R}^{4\times 4}\ \text{per head},\quad QK^\top \in \mathbb{R}^{4\times 4}.
\]
After softmax over rows, multiply by $V$ to return $\mathbb{R}^{4\times 4}$ per head, concatenate to $\mathbb{R}^{4\times 8}$.

\section*{Worked Example (Reasoning Chain with Causal Mask)}
For token position $t=3$ in a length-4 causal sequence, allowed keys are positions $\{1,2,3\}$ only.  
In row 3 of $QK^\top$, mask sets column 4 score to $-\infty$ before softmax.  
Result: attention weight at $(3,4)$ is exactly 0.  
Interpretation: causality is enforced at score level, not corrected later at output level.

\section*{Worked Mini Example (RoPE relative-position identity)}
Let $d_k = 2$ (one dimension pair), positions $m = 1$, $n = 3$, angle $\theta = \pi/6$.
\[
  \mathbf{R}_{m\theta} = \mathbf{R}_{\pi/6} =
  \begin{pmatrix} \cos(\pi/6) & -\sin(\pi/6) \\ \sin(\pi/6) & \cos(\pi/6) \end{pmatrix}
  = \begin{pmatrix} \sqrt{3}/2 & -1/2 \\ 1/2 & \sqrt{3}/2 \end{pmatrix}.
\]
\[
  \mathbf{R}_{n\theta} = \mathbf{R}_{\pi/2} =
  \begin{pmatrix} 0 & -1 \\ 1 & 0 \end{pmatrix}.
\]
For arbitrary $\mathbf{x}, \mathbf{y} \in \mathbb{R}^2$:
\[
  (\mathbf{R}_{\pi/6}\,\mathbf{x})^\top (\mathbf{R}_{\pi/2}\,\mathbf{y})
  = \mathbf{x}^\top \mathbf{R}_{\pi/6}^\top \mathbf{R}_{\pi/2}\,\mathbf{y}
  = \mathbf{x}^\top \mathbf{R}_{(\pi/2 - \pi/6)}\,\mathbf{y}
  = \mathbf{x}^\top \mathbf{R}_{\pi/3}\,\mathbf{y}.
\]
The result depends only on the offset $n - m = 2$ (encoded via angle $\pi/3 = 2 \times \pi/6$), not on the absolute positions 1 and 3. This is the key property in full $d_k$-dimensional RoPE: decompose into $d_k/2$ independent 2D rotation pairs, apply the same argument in each pair.

\section*{Intuition Checkpoints}
\begin{itemize}[leftmargin=1.5em]
  \item ``Masking'' is a constraint on attention scores, not a post-hoc output edit.
  \item Quadratic attention cost refers to score interactions, not all block components equally.
  \item KV cache changes decoding-time behavior, not pretraining-time complexity class.
  \item RoPE encodes absolute position but the dot product reads only relative position --- the rotation group guarantees this algebraically; it is not an approximation.
  \item ALiBi requires no positional vectors and generalises well to lengths longer than seen in training; sinusoidal PE does not generalise as cleanly beyond the training context window.
\end{itemize}

\section*{Further Refresh}
\begin{itemize}[leftmargin=1.5em]
  \item Shape and vector geometry reminders: see Chapter 4 refresher.
  \item Softmax stability reminders: see Chapter 1 refresher.
  \item \textbf{Vaswani et al. (2017)} --- architectural equations and design rationale.
  \item \textbf{Annotated Transformer} --- equation-by-equation implementation mapping.
\end{itemize}
