\chapter{Chapter 3 Refresher: Neural Networks}
\label{chap:neural-networks}
\label{sec:math-refresher-neural-networks}

\scenelabel
Chapter~\ref{chap:neural-networks} names the map \(f_\theta\) that converts contexts (embedded as vectors) into logits consumed by the softmax head from Chapter~\ref{chap:stat-foundations}. Probability Preface Steps~2--3 built distributions over \emph{symbols}; here we build differentiable functions producing \emph{parameters} of those distributions.

\paragraph{Step A --- Affine versus nonlinear composition.}
\textbf{Problem.} Stack linear maps naïvely and nothing new appears (matrix multiplication absorbs depth).
\textbf{Insight.} Insert nonlinearities so each layer genuinely bends space.
\textbf{Lineage.} McCulloch--Pitts abstract neurons; Rosenblatt perceptron (1958); Minsky--Papert XOR diagnosis (1969).
\textbf{Mental model.} Ask ``would removing \(\varphi\) reduce this entire stack to one matrix?'' If yes, redesign.

\paragraph{Step B --- Depth becomes tractable.}
\textbf{Problem.} Naïve depth amplifies or shrinks signals exponentially.
\textbf{Insight.} Backprop applies Preface Step~8's log-product intuition in reverse: derivatives multiply along layers too.
\textbf{Lineage.} Werbos thesis (1974); Rumelhart--Hinton--Williams \emph{Nature} (1986).
\textbf{Mental model.} Forward multiply matrices; backward multiply transposes---same \(\mathcal{O}(P)\) order as one forward pass.

\paragraph{Step C --- Initialise like a physicist.}
\textbf{Problem.} Random weights inject variance into activations; depth compounds variance sums (Preface Step~6).
\textbf{Insight.} Match \(\mathrm{Var}\) layer-to-layer so neither explosions nor vanishing dominate early training.
\textbf{Lineage.} Glorot \& Bengio (2010); He et al.\ (2015).
\textbf{Mental model.} Treat fan-in as degrees of freedom contributing independent squared terms.

\paragraph{Step D --- Normalise and short-cut.}
\textbf{Problem.} Optimisers still distort scale mid-training; identity maps are easiest to learn.
\textbf{Insight.} Normalisation recentres activations; residuals learn perturbations around identity (He et al., 2016; Ba et al., LayerNorm).
\textbf{Mental model.} Depth searches neighbourhoods of stable highways rather than rebuilding maps from scratch each layer.

This refresher installs calculus-chain-rule literacy distinct from the probability chain rule (Preface Step~3): same English phrase, different theorem---always label which you mean.

\section*{Notation Introduced in This Chapter}
\begin{itemize}[leftmargin=1.5em]
  \item \(x,h,z\): vectors (column vectors by default); \(h^{(\ell)}\) activations at layer \(\ell\).
  \item \(W^{(\ell)}\): weight matrix for layer \(\ell\); \(b^{(\ell)}\): bias vector.
  \item \(\varphi\): generic scalar activation applied \emph{elementwise}; \(\varphi'\) its derivative.
  \item \(\sigma(\cdot)\): \emph{only} the logistic sigmoid \(\sigma(z)=1/(1+e^{-z})\) (overload avoided elsewhere in the reader).
  \item \(\odot\): Hadamard (elementwise) product of two vectors of the same shape.
  \item \(f \circ g\): composition, \((f \circ g)(x)=f(g(x))\).
  \item \(J\) or \(\partial u/\partial v\): Jacobian matrix of partial derivatives (shape dictated by numerator layout below).
  \item \(\mathcal{L}\): scalar loss; \(\theta\): flattened parameters (as in Chapter~\ref{chap:optimization} refresher).
\end{itemize}

\section*{What You Need To Recall Precisely}
\begin{itemize}[leftmargin=1.5em]
  \item \textbf{Affine layer.}
  \[
    z = W h_{\mathrm{in}} + b,\qquad W\in\mathbb{R}^{d_{\mathrm{out}}\times d_{\mathrm{in}}},\;
    h_{\mathrm{in}}\in\mathbb{R}^{d_{\mathrm{in}}},\; b,z\in\mathbb{R}^{d_{\mathrm{out}}}.
  \]
  Without an elementwise nonlinearity, any composition of affine maps is still a single affine map; \emph{depth buys nothing}.
  Concrete sketch with two layers: $z_2 = W_2(W_1 x + b_1)+b_2 = (W_2 W_1)x + (W_2 b_1 + b_2)$, still affine in $x$. Adding \(\varphi\) between layers breaks closure because \(\varphi(W_1 x+b_1)\) is generally non-affine in \(x\).

  \item \textbf{MLP skeleton.}
  With pre-activations \(z^{(\ell)}\), post-activations \(h^{(\ell)}=\varphi(z^{(\ell)})\) for hidden layers, and \(h^{(0)}=x\),
  \[
    z^{(\ell)} = W^{(\ell)} h^{(\ell-1)} + b^{(\ell)}.
  \]
  The language-modelling head typically applies softmax to logits \(z^{(L)}\) (Chapter~\ref{chap:stat-foundations} refresher).

  \item \textbf{Two ``chain rules'' (disambiguation).}
  \begin{itemize}[leftmargin=1.5em]
    \item \emph{Probability chain rule} (Chapter~\ref{chap:stat-foundations}): factorises a joint as a product of conditionals, e.g.\ \(p(x_{1:T})=\prod_t p(x_t\mid x_{<t})\).
    \item \emph{Calculus chain rule}: propagates derivatives through composed differentiable maps. Same phrase, different theorem.
  \end{itemize}

  \item \textbf{Scalar calculus chain rule.}
  If \(y=g(u)\) and \(u=h(x)\) are differentiable scalars, \(\displaystyle\frac{dy}{dx}=\frac{dy}{du}\frac{du}{dx}\). For nested \(y=f^{(L)}\!\circ\cdots\circ f^{(1)}(x)\), multiply local derivatives along the graph from output backward.

  \item \textbf{Jacobian view for vectors (reading backprop).}
  Warmup scalar-on-vector case: if \(z\in\mathbb{R}^m\) feeds scalar \(\mathcal{L}(z)\) and \(z=z(\theta)\), chain rule gives \(\partial \mathcal{L}/\partial \theta_j = \sum_i (\partial \mathcal{L}/\partial z_i)(\partial z_i/\partial \theta_j)\)---a dot product of the upstream sensitivity \(\nabla_z \mathcal{L}\) with each column of \(\partial z/\partial \theta\). Packing columns yields matrix-vector form below.

  Take scalar \(\mathcal{L}(z)\) with \(z\in\mathbb{R}^m\) depending on \(\theta\in\mathbb{R}^r\). With \emph{column} gradients \(g_z=\nabla_z \mathcal{L}\in\mathbb{R}^m\) and \(g_\theta=\nabla_\theta\mathcal{L}\in\mathbb{R}^r\),
  \[
    g_\theta = \Bigl(\frac{\partial z}{\partial \theta}\Bigr)^{\!\top} g_z,
  \]
  where \(\partial z/\partial \theta\) has shape \(m\times r\) (row \(i\): \(\partial z_i/\partial \theta\)). Each layer implements one structured Jacobian--vector product; backprop is the reverse topological walk applying those products without forming full Jacobians explicitly.

  \item \textbf{Elementwise \(\varphi\) and Hadamard.}
  If \(h=\varphi(z)\) elementwise and upstream column gradient is \(g_h=\nabla_h \mathcal{L}\), then
  \[
    g_z = g_h \odot \varphi'(z),
  \]
  with \(\varphi'(z)\) applied elementwise (convention at nondifferentiable points, e.g.\ ReLU at \(0\), fixed for implementations).

  \item \textbf{Affine backward pattern (conceptual).}
  For \(z = W h + b\), sensitivities satisfy \(g_h = W^{\top} g_z\) and \(g_W = g_z\, h^{\top}\) (up to batch averaging conventions). You do not need to memorise full layouts if you remember: \emph{multiply by \(W\) forward, by \(W^{\top}\) backward on activations}.

  \item \textbf{Softmax plus cross-entropy gradient.}
  Let logits \(z\in\mathbb{R}^{|V|}\), probabilities \(p=\mathrm{softmax}(z)\), and one-hot target \(e_k\) for class/token \(k\). With \(\mathcal{L}_{\mathrm{CE}}=-\log p_k\),
  \[
    \nabla_z \mathcal{L}_{\mathrm{CE}} = p - e_k.
  \]
  Sketch: write \(\log p_k = z_k - \log\sum_j e^{z_j}\). Differentiate: \(\partial (\log\sum_j e^{z_j})/\partial z_i = p_i\) because softmax derivatives telescope inside the log-sum-exp. Subtracting the indicator \(e_k\) leaves residual \(p_i\) on every coordinate except the target coordinate where it becomes \(p_k-1\). Packaging coordinates yields \(p-e_k\). This identity is the bridge between Chapter~\ref{chap:stat-foundations}'s probabilistic head and Chapter~\ref{chap:neural-networks}'s layered map producing logits---gradients vanish exactly when \(p\) places mass \(1\) on the observed token.

  \item \textbf{Hyperplane and XOR.}
  A linear classifier uses \(\mathrm{sign}(w^{\top}x+b)\); the decision boundary is \(\{x:w^{\top}x+b=0\}\), an affine hyperplane. XOR labels on \(\{(0,0),(0,1),(1,0),(1,1)\}\) are not linearly separable in \(\mathbb{R}^2\): no single line separates the two label classes. A hidden layer folds space so a second boundary can finish the job.

  \item \textbf{Variance sketch for initialisation (one layer).}
  If \(z=\sum_{j=1}^{n} w_j x_j\) with \(w_j\) independent, zero mean, common variance \(\mathrm{Var}(w)\), and fixed \(x\),
  \[
    \mathrm{Var}(z)=\mathrm{Var}(w)\sum_{j=1}^{n} x_j^2.
  \]
  Feeding forward through many layers without controlling \(\mathrm{Var}(z)\) leads to exponential growth or shrinkage of activation scale (exploding / vanishing signals). Xavier/He scaling chooses \(\mathrm{Var}(w)\) proportional to \(1/n_{\mathrm{in}}\) or \(2/n_{\mathrm{in}}\) depending on \(\varphi\) so typical squared norms stay stable layer-to-layer---the reader chapter makes this operational.

  \item \textbf{Universal approximation (one sentence).}
  Wide shallow nets can approximate continuous functions on compact sets under mild conditions on \(\varphi\); this is about \emph{existence of representation}, not whether SGD finds it quickly or with reasonable width---trainability is separate.
\end{itemize}

\section*{How These Results Build on Each Other}
\begin{enumerate}[leftmargin=1.5em]
  \item Chapter~\ref{chap:stat-foundations} fixes the probabilistic head (softmax, CE); gradients at logits land on \(p-e_k\).
  \item Chapter~\ref{chap:optimization} consumes \(\nabla_\theta \mathcal{L}\) as an abstract vector; Chapter~\ref{chap:neural-networks} explains how software obtains it through layered Jacobians.
  \item Affine maps propagate signals forward (\(W,b\)) and sensitivities backward (\(W^{\top}\)).
  \item Elementwise \(\varphi\) applies nonlinearity forward and gates gradients backward via \(\odot \varphi'(z)\).
  \item Composing many layers iterates these two patterns (calculus chain rule), yielding backpropagation when implemented reversibly.
  \item Initialisation and normalisation keep forward magnitudes and backward factors in a trainable range; residuals add an identity path so optimisation can make small corrections around a stable baseline.
\end{enumerate}

\section*{Reminder Glossary (Term by Term)}
\begin{itemize}[leftmargin=1.5em]
  \item \textbf{Pre-activation \(z\):} input to \(\varphi\) before nonlinearity.
  \item \textbf{Post-activation \(h\):} output \(\varphi(z)\).
  \item \textbf{Hadamard product \(\odot\):} multiply matching coordinates of two vectors.
  \item \textbf{Vanishing gradients:} backward factors shrink exponentially with depth so early layers barely move.
  \item \textbf{Exploding gradients:} backward factors grow until norms overflow numerically.
  \item \textbf{Dead ReLU:} unit stuck with \(z\le 0\) forever receives zero \(\varphi'(z)\) and stops learning.
  \item \textbf{BatchNorm / LayerNorm / RMSNorm:} re-centre/rescale activations (reader compares where statistics are taken); purpose includes stabilising scale through depth.
  \item \textbf{Residual / skip connection:} add input of a block to its output so layers learn deltas around identity.
\end{itemize}

\section*{Worked Example (XOR Geometry + One Composition Chain)}
\textbf{Part A (geometry).} XOR assigns label $+1$ to $(0,1)$ and $(1,0)$ and label $-1$ to $(0,0)$ and $(1,1)$. Suppose a linear scoring rule $\mathrm{sign}(w_1 x_1 + w_2 x_2 + b)$ separates the classes with nonzero margin along each labelled point (ideal separator---relax later). Evaluate linear score $s(x)=w_1 x_1 + w_2 x_2 + b$ on the four corners. XOR symmetry forces contradictory inequalities:
\begin{itemize}[leftmargin=1.5em]
  \item Separating $(0,1)$ from $(0,0)$ implies $s(0,1)>s(0,0)$ hence $w_2>0$.
  \item Separating $(1,0)$ from $(0,0)$ implies $w_1>0$.
  \item Separating $(1,1)$ from $(1,0)$ implies $s(1,1)<s(1,0)$ hence $w_2<0$.
\end{itemize}
The second and third requirements contradict each other. No affine hyperplane in $\mathbb{R}^2$ induces opposite signs on exactly those XOR-labelled subsets without contradiction---so linear threshold units alone fail.

\textbf{Part B (scalar depth intuition).} Let \(f(x)=\varphi(w_2\,\varphi(w_1 x)+b_1)+b_2\) with scalar \(x\) and smooth \(\varphi\). Then
\[
\frac{df}{dx}= \varphi'(u_2)\, w_2\, \varphi'(u_1)\, w_1,\quad u_1=w_1 x+b_1,\ u_2=w_2 \varphi(u_1)+b_2.
\]
Each layer contributes a factor \(\varphi'(u)\, w\); depth multiplies many such factors---either shrink toward \(0\) or grow past \(1\) unless \(\varphi'\), weights, and stabilisers are chosen carefully. This scalar picture is the skeleton of vanishing/exploding gradients in vector nets.

\textbf{Practical reading implication:} when Chapter~\ref{chap:neural-networks} of the reader walks backpropagation, read each layer as supplying one Jacobian--vector product like the factors above; when it discusses Xavier/He, read \(\mathrm{Var}(z)\) reasoning like Part B scaled to fan-in \(n\).

\section*{Intuition Checkpoints}
\begin{itemize}[leftmargin=1.5em]
  \item Say ``probability chain rule'' or ``calculus chain rule'' explicitly when both appear in one sitting.
  \item Depth without elementwise \(\varphi\) is equivalent to one affine map.
  \item \(\nabla_\theta \mathcal{L}\) from Chapter~\ref{chap:optimization} is the same object software computes by reverse-mode differentiation through the graph defined here.
  \item \(\odot\) with \(\varphi'(z)\) is where ReLU sparsity (zeros) creates dead-unit risk.
  \item Representation theorems do not guarantee cheap optimisation; judge pipelines by trainability evidence, not approximation slogans alone.
\end{itemize}

\section*{Pointers}
\begin{itemize}[leftmargin=1.5em]
  \item Probability, softmax, cross-entropy, logits: see Chapter 1 refresher.
  \item Gradients as vectors, SGD/Adam, clipping, scaling budgets: see Chapter 2 refresher.
  \item Embedding geometry and dot products reused inside attention: see Chapter 4 refresher.
\end{itemize}

\section*{References and What To Use Them For}
\begin{itemize}[leftmargin=1.5em]
  \item \textbf{Goodfellow--Bengio--Courville, \emph{Deep Learning}} --- backpropagation as computational graphs, activations, initialisation practice.
  \item \textbf{Strang, \emph{Linear Algebra}} --- matrix multiplication, transpose, norms.
  \item \textbf{Cristianini \& Shawe-Taylor (or similar ML math notes)} --- Jacobian and gradient layout conventions if confusion persists.
  \item \textbf{Rumelhart--Hinton--Williams (1986)} --- historical chain-rule-through-networks narrative.
  \item \textbf{Glorot \& Bengio (2010); He et al.\ (2015)} --- variance-preserving initialisation motivations (original papers optional; reader Chapter~\ref{chap:neural-networks} operationalises scaling with \(\varphi\)).
\end{itemize}
