\chapter{Chapter 10 Refresher: LLM Systems --- Retrieval, Tool Use, and Evaluation Harnesses --- Retrieval Metrics and Evaluation Arithmetic}

\scenelabel
Retrieval-augmented answering concatenates two stochastic modules: recall relevant evidence (set-valued event), then condition generation on that evidence (Preface Step~2 again). One scalar cannot isolate which module failed.

\paragraph{Step A --- Set-valued relevance.}
\textbf{Problem.} Truth may admit multiple supporting documents.
\textbf{Insight.} Precision/recall generalise Preface Step~1 counting logic to partial overlaps between retrieved and gold sets.
\textbf{Lineage.} Classical IR (Manning et al.); neural hybrids layer embedding geometry from Chapter~4.

\paragraph{Step B --- Conditional generation quality.}
\textbf{Problem.} Good retriever + weak synthesiser still ships nonsense.
\textbf{Insight.} Evaluate \(\mathcal{L}\) or task metrics \emph{given} oracle contexts vs live contexts---difference isolates grounding gaps.

\paragraph{Step C --- Faithfulness as constrained decoding.}
\textbf{Problem.} Fluent generators invent citations.
\textbf{Insight.} Faithfulness metrics approximate hard logical entailment with softer overlap checks---engineering surrogate for impossible full semantics.

\paragraph{Step D --- Sparse retrieval scoring and hybrid fusion.}
\textbf{Problem.} Dense retrievers miss exact-match keywords; sparse retrievers miss paraphrases. Neither alone suffices for private or technical corpora.
\textbf{Insight.} BM25 is a principled lexical scorer with IDF weighting, TF saturation, and length normalisation. Reciprocal Rank Fusion (RRF) combines BM25 and dense rankings without needing calibrated scores---just ordinal ranks.
\textbf{Lineage.} BM25 (Robertson \& Walker, 1994 BM25 refinements); RRF (Cormack et al.\ 2009).
\textbf{Mental model.} RRF's $k=60$ offset prevents a rank-1 result from dominating; both retrievers contribute robustly even when their score scales are incomparable.

Decomposition honors Preface Step~5 linearity instinct: diagnose components before averaging them blindly.

\section*{Notation Introduced in This Chapter}
\begin{itemize}[leftmargin=1.5em]
  \item $k$: retrieval cutoff (also used as the RRF offset constant, conventionally 60; context distinguishes the two).
  \item $R_k$: top-$k$ retrieved set.
  \item $G$: gold relevant set.
  \item Precision@$k$, Recall@$k$: standard ranking metrics.
  \item $q$: query embedding/vector; also used as query term string.
  \item $d$/$c$: document/chunk embedding/vector.
  \item $t$: a query term.
  \item $\mathrm{TF}(t,c)$: term frequency of $t$ in chunk $c$.
  \item $\mathrm{IDF}(t)$: inverse document frequency of $t$.
  \item $N$: corpus size (number of documents).
  \item $\mathrm{df}(t)$: number of documents containing term $t$.
  \item $|c|$: length of chunk $c$ in tokens; $\mathrm{avgdl}$: corpus average chunk length.
  \item $k_1, b$: BM25 saturation and length-normalisation parameters (typical: $k_1\in[1.2,2.0]$, $b=0.75$).
  \item $r_{\mathrm{bm25}}(c,q)$, $r_{\mathrm{dense}}(c,q)$: rank of chunk $c$ in BM25 and dense retrievals for query $q$.
\end{itemize}

\section*{What You Need To Recall Precisely}
\begin{itemize}[leftmargin=1.5em]
  \item \textbf{Ranking metrics.}
  Let \(R_k\) be top-$k$ retrieved IDs and \(G\) gold relevant IDs with $|G|$ possibly larger than $k$.
  \[
    \mathrm{Prec@}k=\frac{|R_k \cap G|}{k},\qquad
    \mathrm{Rec@}k=\frac{|R_k \cap G|}{|G|}.
  \]
  Precision stresses cleanliness of short rankings; recall stresses coverage of all positives---trade-offs arise whenever $|G|$ is large relative to $k$. Mean reciprocal rank (MRR) averages \(1/\text{rank of first relevant}\) across queries---rewarding systems that surface \emph{one} correct doc quickly rather than maximising exhaustive recall.
  \item \textbf{BM25 scoring formula.}
  For a query $q$ and chunk $c$, BM25 sums over query terms $t \in q$:
  \[
    \mathrm{BM25}(c,q) = \sum_{t\in q}
      \mathrm{IDF}(t)\cdot
      \frac{\mathrm{TF}(t,c)\cdot(k_1+1)}
           {\mathrm{TF}(t,c) + k_1\!\left(1-b+b\cdot\dfrac{|c|}{\mathrm{avgdl}}\right)}.
  \]
  \emph{IDF term.} Measures how rare term $t$ is across the corpus:
  \[
    \mathrm{IDF}(t) = \log\!\left(\frac{N - \mathrm{df}(t) + 0.5}{\mathrm{df}(t) + 0.5}+1\right).
  \]
  Rare terms get high IDF; stop-words appearing in almost every document get IDF near zero.

  \emph{TF saturation.} Raw term frequency is not linear: a document with 10 occurrences is not 10$\times$ more relevant than one with 1. The numerator $\mathrm{TF}(t,c)\cdot(k_1+1)$ divided by $\mathrm{TF}(t,c)+k_1$ is a saturating function that asymptotes at $(k_1+1)$ as $\mathrm{TF}\to\infty$; $k_1$ controls how fast saturation sets in.

  \emph{Length normalisation.} Long documents contain more terms by chance. The $(1-b+b\cdot|c|/\mathrm{avgdl})$ denominator term penalises long documents proportionally, with $b=0$ disabling normalisation entirely and $b=1$ fully correcting for length.

  \emph{Why it matters.} BM25 is the standard baseline for lexical retrieval in every RAG pipeline. Private corpora with specialised terminology (medical, legal, engineering) benefit from BM25's IDF weighting, which down-weights corpus-wide common terms and up-weights rare domain terms without any learned representation.

  \item \textbf{Reciprocal Rank Fusion (RRF).}
  Given a BM25 ranked list and a dense-embedding ranked list, the hybrid score for chunk $c$ against query $q$ is:
  \[
    \mathrm{RRF}(c,q) = \frac{1}{k + r_{\mathrm{bm25}}(c,q)} + \frac{1}{k + r_{\mathrm{dense}}(c,q)},
    \qquad k = 60.
  \]
  Only the ordinal ranks matter---no score calibration required. The offset $k=60$ prevents rank-1 from contributing a disproportionately large $1/1$ term; it acts as a floor that smooths the contribution of highly-ranked items.

  \emph{Why $k=60$?} Cormack et al.\ (2009) found empirically that $k=60$ is robust across many collection types; larger $k$ gives equal weight to all results, smaller $k$ amplifies top-ranked items. 60 is a practical default, not a derived constant.

  \emph{When to use RRF.} Whenever BM25 and dense retrieval are both available (the common hybrid RAG architecture). RRF is parameter-free beyond $k$, requiring no score normalisation---scores from BM25 (typically 0--30) and dense retrievers (cosine similarity, 0--1) are on incompatible scales but ranks are not.

  \item \textbf{Sparse vs dense retrieval:} lexical overlap and embedding similarity capture different relevance modes; hybrid retrieval via RRF combines both without score calibration.
  \item \textbf{Fusion methods:} rank fusion stabilizes retrieval when signals disagree.
  \item \textbf{Evaluation decomposition:} retrieval-stage errors and generation-stage errors require separate diagnostics.
  \item \textbf{Grounding metrics:} answer support must be checked against retrieved evidence, not linguistic plausibility.
\end{itemize}

\section*{How These Results Build on Each Other}
\begin{enumerate}[leftmargin=1.5em]
  \item Define relevance and evidence assumptions.
  \item BM25 scores each chunk by IDF-weighted, length-normalised term frequency.
  \item Dense retrieval scores each chunk by embedding cosine/dot product.
  \item RRF fuses both ranked lists via reciprocal rank sum, no calibration required.
  \item Measure retrieval coverage/precision (Precision@k, Recall@k, MRR).
  \item Measure generation quality conditioned on retrieved context.
  \item Measure grounding/faithfulness of final claims.
  \item Localize failure stage before proposing fix.
\end{enumerate}

\section*{Reminder Glossary (Term by Term)}
\begin{itemize}[leftmargin=1.5em]
  \item \textbf{Gold relevance set:} trusted set of truly relevant items.
  \item \textbf{Precision@k:} fraction of top-$k$ retrieved items that are relevant.
  \item \textbf{Recall@k:} fraction of relevant items captured in top-$k$.
  \item \textbf{MRR (Mean Reciprocal Rank):} average of $1/\text{rank of first relevant doc}$ across queries; rewards surfacing one correct result quickly.
  \item \textbf{BM25:} Best Match 25; probabilistic term-weighting retrieval model with IDF, TF saturation, and length normalisation. The default sparse baseline.
  \item \textbf{IDF (Inverse Document Frequency):} log-scaled measure of term rarity; rare terms receive higher weight.
  \item \textbf{TF saturation:} BM25's sub-linear treatment of term frequency; a term repeated 10 times scores less than 10$\times$ a single occurrence.
  \item \textbf{Length normalisation ($b$):} BM25 parameter controlling penalty for longer documents; $b=0.75$ is the standard default.
  \item \textbf{Dense retrieval:} embedding-based retrieval scoring chunks by dot product or cosine similarity to a query vector.
  \item \textbf{RRF (Reciprocal Rank Fusion):} rank-combination formula $1/(k+r_1) + 1/(k+r_2)$ that merges two ranked lists without score calibration; $k=60$ default.
  \item \textbf{Hybrid retrieval:} architecture running both BM25 and dense retrieval in parallel, fusing results (typically via RRF).
  \item \textbf{Reranker:} second-stage scoring model for refined ranking.
  \item \textbf{Faithfulness:} proportion of answer claims supported by context.
\end{itemize}

\section*{Worked Mini Example (BM25 Single Term)}
Corpus of $N=1000$ documents; term ``\texttt{fiduciary}'' appears in $\mathrm{df}=5$ of them. Query $q=\{\texttt{fiduciary}\}$, chunk $c$ has $\mathrm{TF}(t,c)=2$, $|c|=80$ tokens, $\mathrm{avgdl}=100$. Parameters: $k_1=1.5$, $b=0.75$.

\[
  \mathrm{IDF} = \log\!\left(\frac{1000 - 5 + 0.5}{5 + 0.5}+1\right)
              = \log(182.4) \approx 5.21.
\]
\[
  \text{length-norm denom} = 1 - 0.75 + 0.75\cdot\tfrac{80}{100} = 0.25 + 0.60 = 0.85.
\]
\[
  \text{TF factor} = \frac{2 \cdot (1.5+1)}{2 + 1.5\cdot 0.85} = \frac{5.0}{3.275} \approx 1.53.
\]
\[
  \mathrm{BM25}(c,q) = 5.21 \times 1.53 \approx 7.97.
\]
Key observations: the rare term (\texttt{fiduciary} in only 5/1000 docs) earns high IDF; the TF factor is saturated at 1.53 rather than the raw 2; the shorter-than-average chunk ($|c|/\mathrm{avgdl}=0.8$) receives a slight length bonus.

\section*{Worked Mini Example (RRF Fusion)}
Three chunks retrieved by BM25 and a dense encoder. Ranks (1-indexed):

\begin{center}
\begin{tabular}{lcc}
Chunk & $r_{\mathrm{bm25}}$ & $r_{\mathrm{dense}}$ \\ \hline
A & 1 & 3 \\
B & 2 & 1 \\
C & 3 & 2 \\
\end{tabular}
\end{center}

RRF scores with $k=60$:
\[
  \mathrm{RRF}(A) = \tfrac{1}{61} + \tfrac{1}{63} \approx 0.0164 + 0.0159 = 0.0323,
\]
\[
  \mathrm{RRF}(B) = \tfrac{1}{62} + \tfrac{1}{61} \approx 0.0161 + 0.0164 = 0.0325,
\]
\[
  \mathrm{RRF}(C) = \tfrac{1}{63} + \tfrac{1}{62} \approx 0.0159 + 0.0161 = 0.0320.
\]
Final ranking: $B > A > C$. Chunk $B$ wins by being rank-1 in dense retrieval---even though it was rank-2 in BM25---because the fusion treats both signals symmetrically. The margin is small: rank differences of 1--2 near the top of the list yield very similar RRF scores due to the $k=60$ offset.

\section*{Worked Mini Example (Precision and Recall@k)}
Suppose the gold relevant set has 4 items. Top-5 retrieval returns 3 relevant items.
\[
\mathrm{Precision@5}=\frac{3}{5}=0.6,\qquad
\mathrm{Recall@5}=\frac{3}{4}=0.75.
\]
Interpretation: ranking quality and evidence coverage are both incomplete, but in different ways.

\section*{Worked Example (Failure Localization)}
Pipeline A improves final answer score from 68 to 72.  
But decomposition reveals:
\begin{itemize}[leftmargin=1.5em]
  \item retrieval recall@5 dropped from 0.81 to 0.70,
  \item generation conditioned on gold context improved substantially.
\end{itemize}
Interpretation: generator improved, retriever regressed.  
Action: fix retrieval stage rather than over-tuning generation prompts.

\section*{Intuition Checkpoints}
\begin{itemize}[leftmargin=1.5em]
  \item Better answer fluency can hide retrieval failures.
  \item High retrieval recall can coexist with poor final answers if synthesis is weak.
  \item Metric definitions require explicit edge-case conventions.
  \item BM25 is not a neural model --- it has no learned parameters, runs in microseconds, and degrades gracefully on unseen terminology (the term just gets zero TF). That makes it robust for cold-start private corpora before any fine-tuning is possible.
  \item The TF saturation in BM25 is a deliberate design choice grounded in probability theory (Robertson--Sp\"{a}rck Jones weighting). It prevents one heavily repeated term from swamping the entire score.
  \item RRF's $k=60$ offset means that rank 1 contributes $1/61$ rather than $1/1$: a roughly 60$\times$ smaller maximum score. This prevents any single system from dominating when results disagreeing strongly near the top.
  \item Hybrid retrieval (BM25 + dense + RRF) is the practical default for RAG systems because neither modality alone is sufficient: BM25 misses paraphrases; dense retrieval misses exact-match keywords and OOV terms.
\end{itemize}

\section*{Further Refresh}
\begin{itemize}[leftmargin=1.5em]
  \item Embedding geometry: see Chapter 4 refresher.
  \item Calibration/uncertainty interpretation: see Chapter 1 refresher.
  \item Systems throughput constraints: see Chapter 7 refresher.
  \item \textbf{Manning et al., \emph{IIR}} --- BM25, ranking metrics, IR evaluation.
  \item \textbf{FAISS/HNSW papers} --- ANN retrieval behavior and tradeoffs.
\end{itemize}
