\section{Week 7 Refresher: Low-Rank Adaptation and Quantization}

\subsection*{Setting the Scene}
Week 7 studies approximation as a design tool. The evolution from full fine-tuning to parameter-efficient adaptation and quantized deployment was driven by memory/cost constraints. The key insight is to treat each approximation as a hypothesis with measurable error, not a free performance gain.

\subsection*{Notation Introduced in This Chapter}
\begin{itemize}[leftmargin=1.5em]
  \item $W$: base weight matrix, $\Delta W$: adaptation update.
  \item $A,B$: low-rank factors in LoRA update.
  \item $d,k$: layer input/output dimensions.
  \item $r$: LoRA rank.
  \item $\alpha$: LoRA scaling constant.
  \item $\rho$: trainable parameter fraction versus full fine-tuning.
\end{itemize}

\subsection*{What You Need To Recall Precisely}
\begin{itemize}[leftmargin=1.5em]
  \item \textbf{LoRA update form:}
  \[
    \Delta W = \frac{\alpha}{r}BA,\quad B\in\mathbb{R}^{d\times r},\ A\in\mathbb{R}^{r\times k}.
  \]
  \item \textbf{Trainable parameter fraction:}
  \[
    \rho=\frac{r(d+k)}{dk}.
  \]
  \item \textbf{Quantization model:} real values mapped to finite codes plus scale metadata.
  \item \textbf{Error tradeoff:} memory/bandwidth gains versus approximation and calibration drift risk.
  \item \textbf{Budget accounting:} adaptation and base-model memory are separate components.
\end{itemize}

\subsection*{How These Results Build on Each Other}
\begin{enumerate}[leftmargin=1.5em]
  \item Choose adaptation parameterization (full vs low-rank).
  \item Compute trainable parameter and memory deltas.
  \item Apply numeric approximation (quantization).
  \item Measure quality drift and failure concentration.
  \item Decide deployment gate from explicit thresholds.
\end{enumerate}

\subsection*{Reminder Glossary (Term by Term)}
\begin{itemize}[leftmargin=1.5em]
  \item \textbf{Rank:} dimensionality of adaptation subspace.
  \item \textbf{LoRA scaling factor:} controls update magnitude relative to rank.
  \item \textbf{Quantizer:} mapping from continuous values to discrete levels.
  \item \textbf{Calibration set:} data used to estimate quantization scales.
  \item \textbf{Drift threshold:} maximum acceptable quality loss for deployment.
\end{itemize}

\subsection*{Worked Mini Example (Parameter Fraction)}
Let $d=k=4096$ and $r=16$:
\[
\rho=\frac{16(4096+4096)}{4096\cdot4096}
=\frac{131072}{16777216}\approx 0.0078.
\]
This is about $0.78\%$ of full fine-tuning parameters for that layer.

\subsection*{Worked Example (Quality Gate Under Approximation)}
Suppose full fine-tuned baseline scores 78.4 on target benchmark.  
LoRA+quantized candidate scores 77.1 overall, but subgroup analysis shows:
\begin{itemize}[leftmargin=1.5em]
  \item common cases: -0.6
  \item rare long-context cases: -3.8
\end{itemize}
Reasoning: average drop may appear acceptable, but concentrated degradation violates deployment intent for long-context tasks.  
Decision: reject current quantization profile or introduce subgroup-aware gate.

\subsection*{Intuition Checkpoints}
\begin{itemize}[leftmargin=1.5em]
  \item A low-rank update is a hypothesis about task-adaptation dimensionality, not a theorem.
  \item Quantization quality is distribution-dependent; outlier channels can dominate failure.
  \item ``Near parity'' must be defined metric-by-metric, not as a vague label.
\end{itemize}

\subsection*{Further Refresh}
\begin{itemize}[leftmargin=1.5em]
  \item Rank and linear-algebra background: see Week 3 refresher.
  \item Optimizer/memory arithmetic: see Week 2 refresher.
  \item Systems bottlenecks and throughput: see Week 6 refresher.
  \item \textbf{LoRA (Hu et al.)} --- low-rank adaptation mechanics.
  \item \textbf{QLoRA (Dettmers et al.)} --- quantization plus adapter training.
\end{itemize}
