\section{Week 11 Refresher: Latency, Queues, and Reliability Budgets}

\subsection*{Setting the Scene}
Week 11 brings probability back into systems operations. Queueing theory, retry arithmetic, and budget constraints determine whether an otherwise capable model can be deployed reliably. The key insight is compounding: small per-request probabilities become large system effects at scale.

\subsection*{Notation Introduced in This Chapter}
\begin{itemize}[leftmargin=1.5em]
  \item $\lambda$: request arrival rate.
  \item $\mu$: service rate.
  \item $\rho=\lambda/\mu$: utilization.
  \item $\mathbb{E}[W]$: expected waiting time.
  \item $p_f$: per-attempt failure probability.
  \item $\mathbb{E}[K]$: expected number of attempts.
\end{itemize}

\subsection*{What You Need To Recall Precisely}
\begin{itemize}[leftmargin=1.5em]
  \item \textbf{Latency decomposition:} prefill latency plus per-output-token decode latency.
  \item \textbf{Queueing baseline:} in M/M/1,
  \[
    \mathbb{E}[W]=\frac{1}{\mu-\lambda},
  \]
  so wait time grows sharply near saturation.
  \item \textbf{Retry expectation:} repeated failures produce nonlinear expected attempts.
  \item \textbf{Token cost model:} base cost is linear in tokens, expected cost depends on retries and aborts.
  \item \textbf{Termination guards:} bounded steps/tokens/cost enforce finite agent behavior.
\end{itemize}

\subsection*{How These Results Build on Each Other}
\begin{enumerate}[leftmargin=1.5em]
  \item Estimate traffic intensity $(\lambda/\mu)$.
  \item Predict queue delay and saturation risk.
  \item Incorporate retries into effective load and expected cost.
  \item Enforce runtime guards for bounded execution.
\end{enumerate}

\subsection*{Reminder Glossary (Term by Term)}
\begin{itemize}[leftmargin=1.5em]
  \item \textbf{Arrival rate $\lambda$:} requests per unit time.
  \item \textbf{Service rate $\mu$:} processing capacity per unit time.
  \item \textbf{Utilization $\rho$:} $\lambda/\mu$.
  \item \textbf{Retry multiplier:} expected attempts per successful request.
  \item \textbf{Budget guard:} hard bound on cost/tokens/steps.
\end{itemize}

\subsection*{Worked Mini Example (Retries)}
Let per-attempt failure probability be $p_f=0.2$ and assume independent retries with no cap.
\[
\mathbb{E}[K]=\frac{1}{1-p_f}=\frac{1}{0.8}=1.25.
\]
If average successful request cost is \$0.04, expected cost with retries is roughly
\[
0.04\times 1.25 = \$0.05.
\]

\subsection*{Worked Example (Queue and Retry Coupling)}
If base arrival rate is 80 req/s and retries add 25\% expected attempts, effective arrival becomes:
\[
\lambda_{\text{eff}} = 80 \times 1.25 = 100\ \text{req/s}.
\]
If service rate is $\mu=110$ req/s, utilization rises from $0.73$ to $0.91$.  
Interpretation: retry policy can push system into tail-latency regime even when baseline looked safe.

\subsection*{Intuition Checkpoints}
\begin{itemize}[leftmargin=1.5em]
  \item Systems can meet average SLOs while failing badly in tails.
  \item Queueing approximations are useful planning tools, not exact traffic simulators.
  \item Reliability controls (backoff, circuit breaking) alter both latency and cost.
\end{itemize}

\subsection*{Further Refresh}
\begin{itemize}[leftmargin=1.5em]
  \item Systems scaling and bottlenecks: see Week 6 refresher.
  \item Evaluation and trust boundaries: see Week 9 refresher.
  \item Risk thresholding and controls: see Week 10 refresher.
  \item \textbf{Harchol-Balter} --- queueing and latency scaling.
  \item \textbf{SRE workbooks} --- retries/backoff/overload control design.
\end{itemize}
