\chapter{Chapter 11 Refresher: Governance, Assurance, and the Future of AI Systems --- Risk Arithmetic and Assurance Logic}
\label{chap:governance}

\scenelabel
Risk arithmetic translates qualitative hazards into inequalities you can monitor---probability returns as operational thresholds on empirical frequencies (Preface Steps~1--4 recycled for compliance narratives).

\paragraph{Step A --- Hazards as events.}
\textbf{Problem.} Stakeholders disagree verbally unless \(\Omega\) carved crisply.
\textbf{Insight.} Define hazardous outcomes as measurable events; attach severities/likelihoods only after measurability clears.
\textbf{Lineage.} Leveson STAMP lineage; ISO/IEC flavour standards.

\paragraph{Step B --- Controls multiply imperfectly.}
\textbf{Problem.} Layered mitigations stack but rarely independent.
\textbf{Insight.} Product-form residual models mirror naive independence yet expose correlation caveats---same scepticism as Preface Step~6 covariance warnings.

\paragraph{Step C --- Evidence graphs.}
\textbf{Problem.} Auditors demand provenance.
\textbf{Insight.} Assurance cases are Bayesian-flavoured narratives without always updating posteriors---still DAGs connecting claims to sensors.

Governance makes Preface probability language accountable to humans outside ML conferences.

\section*{Notation Introduced in This Chapter}
\begin{itemize}[leftmargin=1.5em]
  \item $R$: baseline risk score.
  \item $R_{\text{res}}$: residual risk after controls.
  \item $\eta_i$: efficacy of control $i$.
  \item $m_t$: monitored metric at time $t$.
  \item $m_0$: baseline metric value.
  \item $\epsilon$: escalation threshold.
\end{itemize}

\section*{What You Need To Recall Precisely}
\begin{itemize}[leftmargin=1.5em]
  \item \textbf{Risk decomposition:} severity and likelihood are different axes with different uncertainty.
  \item \textbf{Layered controls.}
  With baseline scalar risk score \(R\) and independent-ish controls each attenuating fraction \(\eta_i\in[0,1]\),
  \[
    R_{\mathrm{res}} = R \prod_{i}(1-\eta_i).
  \]
  Interpret \(\eta_i\) as fractional reductions attributable to distinct safeguards under optimistic independence assumptions; shared vulnerabilities violate independence and make multiplicative models overly cheerful---inflate residuals mentally whenever controls depend on the same latent failures.
  \item \textbf{Trigger logic:} indicator functions implement threshold decisions.
  \item \textbf{Assurance structure:} claim $\to$ evidence $\to$ control action $\to$ artifact.
  \item \textbf{Correlation caveat:} multiplicative risk-reduction models are optimistic under shared-cause failures.
\end{itemize}

\section*{How These Results Build on Each Other}
\begin{enumerate}[leftmargin=1.5em]
  \item Define risk factors and their measurement status.
  \item Model baseline and post-control residual risk.
  \item Apply threshold policy to operational metrics.
  \item Attach each decision to auditable evidence.
\end{enumerate}

\section*{Reminder Glossary (Term by Term)}
\begin{itemize}[leftmargin=1.5em]
  \item \textbf{Hazard:} undesirable outcome class to prevent/mitigate.
  \item \textbf{Control efficacy:} estimated fractional risk reduction from a control.
  \item \textbf{Residual risk:} remaining risk after controls.
  \item \textbf{Threshold policy:} decision rule on monitored metrics.
  \item \textbf{Assurance artifact:} evidence item supporting a claim.
\end{itemize}

\section*{Worked Mini Example (Single vs Two Controls)}
If baseline risk score is $R=1.0$, and control efficacies are $\eta_1=0.4$, $\eta_2=0.3$:
\[
R_1=R(1-\eta_1)=0.6,\qquad
R_{12}=R(1-\eta_1)(1-\eta_2)=0.42.
\]
Interpretation: layered controls reduce residual risk multiplicatively under the model assumptions.

\section*{Worked Example (Threshold and Escalation)}
Suppose drift metric baseline is $m_0=0.12$ and policy threshold is $\epsilon=0.05$.  
Current observed $m_t=0.19$ gives deviation $|m_t-m_0|=0.07>\epsilon$.  
Escalation rule triggers high-review mode.  
Reasoning chain: measured metric -> threshold test -> policy action -> logged evidence.

\section*{Intuition Checkpoints}
\begin{itemize}[leftmargin=1.5em]
  \item Multiplicative control models can be optimistic if failures are correlated.
  \item Governance thresholds are policy design decisions, not immutable laws.
  \item Audit readiness is about traceable evidence chains, not only numeric scores.
\end{itemize}

\section*{Further Refresh}
\begin{itemize}[leftmargin=1.5em]
  \item Metric reliability and uncertainty: see Chapter 1 refresher.
  \item Evaluation decomposition and harness logic: see Chapter 10 refresher.
  \item \textbf{Leveson, \emph{Engineering a Safer World}} --- hazard/control argument structure.
  \item \textbf{NIST AI RMF} --- operational governance controls and documentation patterns.
\end{itemize}
