\chapter{Chapter 1 Refresher: Statistical Learning Foundations for Large Language Models --- Probability, Likelihood, and Calibration}

\scenelabel
Chapter~1 of the reader applies the probability spine summarised in the \emph{Preface: Probability Building Blocks}: Steps~2--4 turn sequential text into products of conditionals; Steps~8--10 justify cross-entropy training as likelihood plus coding intuition; Step~7 motivates interpreting finite corpora as imperfect proxies for expectations.

\paragraph{Reasoning cadence (Strang-style).}
Each numbered derivation below answers an explicit question that bothered practitioners historically:
\begin{enumerate}[leftmargin=1.5em]
  \item \textbf{What object is ``predict next token''?} Conditional probability (restrict and renormalise to context).
  \item \textbf{How do those pieces assemble into a story probability?} Iterated conditioning---Markov's linguistic intuition modernised.
  \item \textbf{How do we fit parameters?} Fisher-style maximum likelihood $\Rightarrow$ sums of log probabilities.
  \item \textbf{How do we interpret the scalar loss?} Shannon/Kullback--Leibler information decomposition separates irreducible noise from model mismatch.
  \item \textbf{Why softmax logits?} Analytic convenience plus numerical stabilisation via shift invariance (engineering cousin of log-sum-exp proofs).
  \item \textbf{What does calibration add that loss omits?} Reliability of stated probabilities versus empirical hit rates (psychometrics meets scoring rules).
\end{enumerate}

The critical insight for Chapter~1 is that language modeling does not invent probability theory; it \emph{sequences} classical moves under computational constraints. Logarithms convert products into sums because curvature and gradient variance behave better that way---not because theoreticians prefer prettier notation.

This refresher chapter specialises the preface steps into LM notation; reread the preface whenever conditioning identities feel unmotivated.

\section*{Notation Introduced in This Chapter}
\begin{itemize}[leftmargin=1.5em]
  \item $x_t$: token at position $t$ in a sequence; $x_{<t}$ denotes all preceding tokens (the context).
  \item $V$: vocabulary; $|V|$ is vocabulary size.
  \item $p_\theta(\cdot \mid x_{<t})$: model's conditional distribution over next tokens given context.
  \item $z_t\in\mathbb{R}^{|V|}$: vector of logits before softmax at position $t$.
\end{itemize}

\section*{What You Need To Recall Precisely}
\begin{itemize}[leftmargin=1.5em]
  \item \textbf{Conditional probability as prediction primitive.} If $A$ and $B$ are events with $P(B)>0$, then
  \[
    P(A\mid B)=\frac{P(A\cap B)}{P(B)}.
  \]
  For language modeling, $A$ is ``next token equals $x_t$'' and $B$ is ``context is $x_{<t}$''.
  \textit{Insight.} Conditioning formalises ``given everything written so far.''
  \textit{Lineage.} Kolmogorov-era formalism; psychological predecessors debated inverse probability for centuries.
  \textit{Mental model.} Always ask what slice of $\Omega$ you conditioned on---ambiguous context strings yield ambiguous measures.

  \item \textbf{Chain rule as sequence constructor.}
  Start from the definition of conditional probability in joint form,
  \[
    P(A,B)=P(A\mid B)\,P(B).
  \]
  Identify $(A,B)$ with ``token $x_T$ takes its realised value \emph{and} prefix $x_{<T}$ takes its realised values''. Then $P(x_{1:T})=P(x_T\mid x_{<T})\,P(x_{<T})$. Peel the same identity off $P(x_{<T})$ to introduce $P(x_{T-1}\mid x_{<T-1})$, and continue until only $P(x_1)$ remains. Induction yields
  \[
    p(x_{1:T}) = \prod_{t=1}^{T} p(x_t \mid x_{<t}).
  \]
  Reading direction: first factor is ``probability of $x_1$ before anything is observed'' (often simplified under a start token convention); each later factor is ``next-token prediction given everything to the left''. This is \emph{not} a modeling trick---it is bookkeeping forced by the definition of conditional probability once you adopt an ordering on positions.
  \textit{Insight.} Long-range dependence enters only through what you put inside each conditional---the factorisation itself assumes nothing about Markov structure.
  \textit{Lineage.} Used implicitly by Markov (1913); standard autoregressive NLP since Bengio et al.\ (2003) and contemporaries.
  \textit{Mental model.} If a factor equals zero, later narrative probability collapses---numerical underflow in logs mirrors this.

  \item \textbf{Likelihood to optimization objective.} Maximum likelihood chooses parameters that maximize probability of observed data. Because products are awkward numerically, we apply log:
  \[
    \arg\max_\theta \prod_i a_i(\theta)=\arg\max_\theta \sum_i \log a_i(\theta).
  \]
  Then we minimize negative average log-likelihood (NLL).
  \textit{Insight.} Likelihood rewards parameters that assign mass to what happened; it is silent about what \emph{could} happen unless regularised.
  \textit{Lineage.} Fisher's likelihood programme; modern empirical risk minimisation inherits the template.
  \textit{Mental model.} $\log$ turns ``multiply tiny probabilities'' into ``sum manageable negatives''---mirror of Preface Step~8.

  \item \textbf{Expectation as average under a distribution.} If $X$ is discrete:
  \[
    \mathbb{E}[f(X)] = \sum_x p(x)f(x).
  \]
  This lets us interpret empirical losses as estimates of population quantities.
  \textit{Insight.} Expectation linearises aggregation---bias/variance decompositions for estimators start here.
  \textit{Lineage.} Integral formulation ascends to Lebesgue; discrete view suffices for finite vocabularies.
  \textit{Mental model.} Preface Step~5 bridges to finite-batch stochastic optimisation noise.

  \item \textbf{Entropy, cross-entropy, and KL in one chain.}
  For distributions $p$ (reference/true) and $q$ (model):
  \[
    H(p) = -\sum_x p(x)\log p(x),\quad
    H(p,q) = -\sum_x p(x)\log q(x),
  \]
  \[
    D_{\mathrm{KL}}(p\|q)=\sum_x p(x)\log\frac{p(x)}{q(x)}.
  \]
  Then
  \[
    H(p,q) = H(p) + D_{\mathrm{KL}}(p\|q).
  \]
  Interpretation in coding language: $H(p,q)$ is how many nats you pay on average if you compress draws from $p$ using an optimal code designed for $q$. Split that bill into two parts: $H(p)$ pays for irreducible uncertainty in $p$ itself; $D_{\mathrm{KL}}(p\|q)$ pays the \emph{extra} bits from model mismatch. Because $H(p)$ does not depend on $q$, minimising $H(p,q)$ over model parameters is the same optimisation problem as minimising $D_{\mathrm{KL}}(p\|q)$; cross-entropy loss is simply the sample estimate of $H(p,q)$.
  \textit{Insight.} Optimising CE pushes $q$ toward empirical frequencies while respecting parameterisation constraints.
  \textit{Lineage.} Shannon (entropy/coding); Kullback--Leibler (1951); revived obsessively in deep learning loss engineering.
  \textit{Mental model.} KL is asymmetric---treat direction carefully when KL penalties anchor policies or priors.

  \item \textbf{Softmax as probability map from logits:}
  \[
    \mathrm{softmax}(z)_i = \frac{e^{z_i}}{\sum_j e^{z_j}}.
  \]
  \textbf{Shift invariance (numerical stability):}
  \[
    \mathrm{softmax}(z) = \mathrm{softmax}(z - c\mathbf{1}) \quad \text{for any scalar } c.
  \]
\nobreak
  Proof sketch: multiplying every numerator and the normalising denominator by $e^{-c}$ leaves all ratios unchanged. Choosing $c=\max_i z_i$ forces every exponent to be $\le 0$, so each $e^{z_i-c}\in(0,1]$ and the sum cannot explode upward before you take logs downstream.
  \textit{Insight.} Softmax is the exponential-family ``linear predictor $\to$ simplex'' map used in multinomial logistic regression for decades.
  \textit{Lineage.} Statistical mechanics analogies (Boltzmann/Gibbs); modern LM heads inherit the template unchanged.
  \textit{Mental model.} Large logits amplify sensitivity---temperature scaling literally rescales that curvature.

  \item \textbf{Perplexity as exponentiated cross-entropy.}
  \[
    \mathrm{PPL}=\exp(H)
  \]
  when $H$ is in nats. Perplexity is interpretable only with fixed tokenization and log base.
  \textit{Insight.} Perplexity converts cross-entropy into ``effective branching factor'' language Shannon favoured when estimating English entropy.
  \textit{Lineage.} Shannon's guessing-game estimates; modern LM tooling reproduces the metaphor for engineers who dislike logs.
  \textit{Mental model.} Changing tokenizer granularity rescales typical perplexities---never compare runs blindly.

  \item \textbf{Calibration distinction (fit vs honesty of confidence).}
  Training loss averages $-\log q_\theta(\text{true token})$---it measures whether probability mass sits on what happened in data. \emph{Calibration} asks a different question: when the model assigns predicted probability $s$ to its top choice (or to events it phrases as ``confident''), does that choice hit roughly $s$ fraction of the time on held-out slices? A model can achieve moderately good cross-entropy yet assign $0.9$ to answers that are only correct $60\%$ of the time (overconfidence), or spread mass flatly and look calibrated while guessing poorly. Always separate ``did we assign probability to what occurred?'' from ``do stated probabilities match empirical frequencies?''.
  \textit{Insight.} Proper scoring rules optimise sharpness \emph{and} honesty under correctly specified beliefs; finite-parameter nets rarely satisfy both simultaneously without explicit calibration machinery.
  \textit{Lineage.} Brier (1950); modern neural calibration literature (Guo et al., 2017, among others).
  \textit{Mental model.} Deploy metrics conditional on confidence bins---aggregate accuracy hides structured lies about uncertainty.
\end{itemize}

\section*{How These Results Build on Each Other}
\begin{enumerate}[leftmargin=1.5em]
  \item Conditional probability defines next-token prediction.
  \item Chain rule extends next-token predictions to full-sequence probabilities.
  \item Likelihood turns full-sequence probabilities into a training objective.
  \item Log transform converts objective from product form to additive form.
  \item Empirical averaging converts sample objective to dataset objective.
  \item Cross-entropy/KL decomposition gives interpretation of what minimizing the objective does.
  \item Softmax provides a valid distribution over vocabulary for each context.
  \item Calibration analysis audits whether confidence values are trustworthy after fitting.
\end{enumerate}

\section*{Reminder Glossary (Term by Term)}
\begin{itemize}[leftmargin=1.5em]
  \item \textbf{Random variable:} a map from outcomes to numerical values.
  \item \textbf{Conditional probability:} probability under restricted information.
  \item \textbf{Likelihood:} probability of observed data viewed as function of parameters.
  \item \textbf{Log-likelihood:} logarithm of likelihood; additive over samples/tokens.
  \item \textbf{NLL:} negative log-likelihood; minimized in training.
  \item \textbf{Entropy $H(p)$:} uncertainty of true distribution.
  \item \textbf{Cross-entropy $H(p,q)$:} expected code length when coding $p$ with model $q$.
  \item \textbf{KL divergence $D_{\mathrm{KL}}(p\|q)$:} mismatch penalty from using $q$ when truth is $p$.
  \item \textbf{Logits:} pre-softmax scores in $\mathbb{R}^{|V|}$.
  \item \textbf{Softmax:} normalization from logits to a probability simplex.
  \item \textbf{Perplexity:} exponentiated average NLL/cross-entropy (with fixed log base).
  \item \textbf{Calibration:} correspondence between predicted confidence and observed accuracy.
  \item \textbf{ECE:} binned estimate of calibration gap.
\end{itemize}

\section*{Worked Example (Non-Toy, End-to-End Reasoning)}
Assume a validation slice with $N=1000$ predicted tokens. We observe:
\begin{itemize}[leftmargin=1.5em]
  \item Mean NLL $=1.10$ nats.
  \item Mean confidence of top-1 predictions $=0.78$.
  \item Actual top-1 accuracy $=0.64$.
\end{itemize}
Now reason in the chain used in Chapter~1:

\textbf{Step 1: Loss to perplexity.}
\[
\mathrm{PPL}=\exp(1.10)\approx 3.00.
\]
Interpretation: average uncertainty corresponds to about three equiprobable choices per token on this slice.

\textbf{Step 2: Loss does not equal calibration.}  
Despite moderate loss, confidence-accuracy gap is
\[
\Delta_{\mathrm{cal}} = 0.78 - 0.64 = 0.14.
\]
The model is overconfident by 14 percentage points on average.

\textbf{Step 3: Why this matters for downstream chapters.}  
Chapter 9 (alignment) can reduce harmful outputs while still leaving calibration issues.  
Chapter 10 (RAG/evaluation) can improve factual grounding while confidence remains miscalibrated.  
Therefore the distinction between fit quality and confidence quality is operational, not philosophical.

\textbf{Step 4: Practical intervention path.}
\begin{enumerate}[leftmargin=1.5em]
  \item Keep objective fit diagnostics (NLL/perplexity).
  \item Add calibration diagnostics (reliability bins, ECE).
  \item Apply temperature scaling on held-out data.
  \item Re-check both loss and calibration after scaling.
\end{enumerate}

\section*{Intuition Checkpoints}
\begin{itemize}[leftmargin=1.5em]
  \item A lower NLL does \emph{not} guarantee calibrated confidence.
  \item KL is a divergence, not a metric; direction matters.
  \item Perplexity is only comparable under matched tokenization and log base.
  \item ECE is a binned approximation and depends on binning and sample size.
\end{itemize}

\section*{If You Get Stuck}
\begin{itemize}[leftmargin=1.5em]
  \item Rewrite every probability expression in words before manipulating symbols.
  \item Track whether each quantity is theoretical, empirical estimate, or model output.
  \item Check if the reasoning step is algebraic (identity) or statistical (estimation).
\end{itemize}

\section*{References and What To Use Them For}
\begin{itemize}[leftmargin=1.5em]
  \item \textbf{Murphy, \emph{Probabilistic Machine Learning}} --- conditional probability, expectation, estimation viewpoint.
  \item \textbf{Bishop, \emph{Pattern Recognition and Machine Learning}} --- likelihood-based learning, cross-entropy objectives.
  \item \textbf{Cover--Thomas, \emph{Elements of Information Theory}} --- entropy, cross-entropy, KL interpretation.
  \item \textbf{Guo et al. (2017)} --- practical calibration diagnostics and temperature scaling.
  \item \textbf{Wasserman, \emph{All of Statistics}} --- fast refresh on estimation, uncertainty, and empirical risk language.
\end{itemize}
